---
title: "Completely randomized designs and one-way inference"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Completely randomized designs and one-way inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

The one-way completely randomized design is the simplest setting, but it still illustrates a central principle: a rank test answers a hypothesis about distributions or relative effects, not automatically a hypothesis about arithmetic means.



## State of the art

Kruskal-Wallis remains a useful omnibus reference for independent groups. Modern rank-based analysis adds interpretable relative effects, studentized permutation procedures, and simultaneous contrasts. For unbalanced designs, pseudo-ranks are preferable when the target is an unweighted relative treatment effect because ordinary ranks can implicitly change the estimand with sample allocation.

## `np_crd()`: three examples

```r
# 1. Four cultivars
x1 <- simulate_agri("crd", seed = 21)
np_crd(yield ~ treatment, x1)

# 2. Ordinal severity response
x2 <- x1
x2$severity <- ordered(cut(x2$yield, quantile(x2$yield, c(0, .25, .5, .75, 1)),
                           include.lowest = TRUE, labels = 1:4))
np_crd(as.numeric(severity) ~ treatment, x2, method = "kruskal")

# 3. Unequal sample sizes
x3 <- x1[-c(1, 2, 8), ]
np_crd(yield ~ treatment, x3, method = "kruskal")
```

## `agri_rank()`: three examples

```r
d <- agri_design(yield ~ treatment, x1, design = "crd")
agri_rank(d, method = "auto")
agri_rank(d, method = "kruskal", estimand = "distribution")
if (requireNamespace("permuco", quietly = TRUE)) agri_rank(d, method = "permuco")
```

## `agri_effects()`: three examples

```r
fit <- agri_rank(d)
agri_effects(fit)
agri_effects(fit, ci = TRUE, B = 199, seed = 22)
if (requireNamespace("rankFD", quietly = TRUE)) {
  fd <- agri_design(yield ~ treatment, x3, design = "factorial")
  agri_effects(agri_rank(fd, method = "rankFD"))
}
```

## Reporting

Do not translate a significant Kruskal-Wallis result into “the means differ.” Describe the response scale and the estimand actually represented by the method. Pairwise comparisons should follow the scientific comparison structure, with multiplicity control where appropriate.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.
