---
title: "Hierarchical Plot Designs, Trends, ANCOVA, and Power"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchical Plot Designs, Trends, ANCOVA, and Power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.8,
  fig.height = 4.4,
  dpi = 130
)
```

**Hierarchical designs vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** designs with more than one randomization level, ordered-dose trend
tests, rank-based covariance adjustment, and simulation-based power.

---

# 1. Why this vignette exists

A field experiment often cannot randomize everything at the same scale.
Irrigation is applied to large areas because the equipment demands it. Tillage
is applied to whole strips because a tractor cannot turn inside a plot. Sowing
date is applied to a whole section because the operation happens on one day.

The consequence is that the experiment has **more than one experimental unit**,
and therefore more than one error stratum. Each effect must be tested against
the variation appropriate to the level at which it was randomized.

Getting this wrong is not a matter of efficiency. Testing a whole-plot factor
against the subplot residual makes it far too easy to detect, and the error is
invisible in the output: the analysis runs, produces a small p-value, and is
wrong in a predictable direction.

> **Every effect is tested against the variation of the unit to which it was
> randomized. A design with three randomization levels has three error strata,
> and no analysis that collapses them is admissible.**

## 1.1 What else is here

Three further topics share this vignette because they all concern getting the
**structure** right before the inference:

| Topic | The structural question it answers |
|---|---|
| trend tests | is the ordered dose response monotone, and is that the right question |
| ANCOVA | can a measured covariate remove variation the blocking could not |
| power | is the design capable of detecting what matters |

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. identify how many experimental units a field layout created, and name each;
2. declare a split-plot, split-split-plot and strip-plot correctly;
3. explain which effect is tested against which stratum, in each of the three;
4. recognise the specific failure that follows from collapsing strata;
5. distinguish a strip-plot from a split-plot, and say why the interaction
   differs between them;
6. choose an ordered-dose trend test rather than an unordered comparison, and
   say what it gains;
7. apply a rank-based covariance adjustment and state the assumption it makes;
8. distinguish a legitimate covariate from one that is affected by treatment;
9. run a simulation-based power study whose data-generating assumption is
   explicit;
10. read a power table as a design decision rather than as a post-hoc excuse;
11. recognise the documented calibration limitation of the permutation backend
    in two of these designs, and route around it.

---

# 3. The hierarchical module in one map

| Function | Design | Randomization levels |
|---|---|---|
| `np_splitplot()` | split-plot | 2 |
| `np_splitsplit()` | split-split-plot | 3 |
| `np_stripplot()` | strip-plot | 2, crossed |
| `agri_trend()` | any, with an ordered quantitative treatment | as declared |
| `agri_ancova()` | any, with a measured covariate | as declared |
| `agri_power()` | any | as generated |

## 3.1 The one table that prevents most errors

```{r strata-map}
data.frame(
  design = c("split-plot", "split-plot", "split-plot",
             "strip-plot", "strip-plot", "strip-plot"),
  effect = c("whole plot", "subplot", "interaction",
             "strip A", "strip B", "interaction"),
  tested_against = c("whole-plot error", "subplot error", "subplot error",
                     "strip-A error", "strip-B error", "residual error")
)
```

Note the difference in the last rows. In a split-plot the interaction is tested
against the subplot error. In a strip-plot it has an error of its own. That is
the whole distinction between the two designs, and it is why they are not
interchangeable names for the same layout.

---

# Part I. Split-plot designs

# 4. What makes a design a split-plot

## 4.1 The physical situation

One factor is applied to large areas because it must be. A second factor is then
randomized **within** those areas. The large area is the whole plot; the
subdivision is the subplot.

| Example | Whole plot | Subplot |
|---|---|---|
| irrigation regime, then cultivar | irrigation | cultivar |
| tillage system, then nitrogen rate | tillage | nitrogen |
| sowing date, then variety | sowing date | variety |
| fumigation, then rootstock | fumigation | rootstock |

## 4.2 The statistical consequence

There are two randomizations, so there are two experimental units and two error
strata.

The whole-plot factor has as many independent units as there are whole plots,
which is typically far fewer than the number of subplots. Its comparison is
therefore **less precise**, and the design knows this even when the analyst does
not.

## 4.3 Fitting one

```{r splitplot-fit}
sp <- simulate_agri("split_plot", seed = 501, n = 5)
str(sp)

fit_sp <- np_splitplot(
  yield ~ irrigation * cultivar,
  data       = sp,
  block      = block,
  whole_plot = irrigation,
  subplot    = cultivar,
  method     = "auto"
)

fit_sp
```

```{r splitplot-anova}
anova(fit_sp)
```

## 4.4 Reading the table

| Effect | Randomized to | Tested against | Typically |
|---|---|---|---|
| `irrigation` | the whole plot | whole-plot error | fewer units, less precise |
| `cultivar` | the subplot | subplot error | more units, more precise |
| `irrigation:cultivar` | the subplot | subplot error | as precise as the subplot factor |

A frequent surprise for readers new to split-plots: the **interaction is tested
more precisely than the whole-plot main effect**. That is not an anomaly. The
interaction is a comparison among subplots, and subplots are numerous.

## 4.5 The figure

Three figures answer three different questions, and it is worth being explicit
about which is which. The first shows the observations as laid out, before any
model. The second asks whether the subplot factor behaves the same way under each
whole-plot treatment. The third summarises the fitted effects.

```{r splitplot-plot-data, fig.cap = "The observations themselves, grouped by the declared strata. Look here first, before any test."}
agri_plot(fit_sp$design, type = "data")
```

```{r splitplot-plot, fig.cap = "Interaction figure for the split-plot. Non-parallel lines suggest that the subplot factor behaves differently under each whole-plot treatment."}
agri_plot(fit_sp, type = "interaction")
```

```{r splitplot-plot-effects, fig.cap = "Fitted effects. Note that the whole-plot factor carries the wider uncertainty, because it is tested against the coarser stratum."}
agri_plot(fit_sp, type = "effects")
```

## 4.6 The error that this design invites

```{r splitplot-wrong}
# Analysed as an ordinary factorial in blocks: the hierarchy is discarded.
des_wrong <- agri_design(yield ~ irrigation * cultivar, sp,
                         design = "rcbd", block = block)
anova(agri_rank(des_wrong, method = "ART"))
```

Compare that table against the split-plot output of section 4.3. The subplot
factor and the interaction are largely unaffected, because they were already
being tested against the subplot error. The **whole-plot factor** is where the
damage occurs: analysed as a factorial, it is tested against a residual that is
far too small, and it becomes far too easy to declare significant.

This is among the most common analysis errors in agronomy, and it is almost
always invisible in the output.

## 4.7 Interpretation

Report the design as a split-plot in the methods, name the whole plot and the
subplot, and give the degrees of freedom of both strata. A reader can then check
that the whole-plot factor was tested against the right denominator.

---

# 5. Split-split-plot designs

## 5.1 Three randomization levels

```{r splitsplit-fit}
ssp <- simulate_agri("split_split", seed = 502, n = 4)

fit_ssp <- np_splitsplit(
  yield ~ irrigation * cultivar * timing,
  data       = ssp,
  block      = block,
  whole_plot = irrigation,
  subplot    = cultivar,
  subsubplot = timing,
  method     = "auto"
)

fit_ssp
```

```{r splitsplit-anova}
anova(fit_ssp)
```

## 5.2 The strata

Written as a nesting, with block $R$, whole-plot factor $A$, subplot factor $B$
and sub-subplot factor $C$, the randomization hierarchy is

$$R \;\rightarrow\; R{:}A \;\rightarrow\; R{:}A{:}B \;\rightarrow\; R{:}A{:}B{:}C.$$

The first three terms are the block, whole-plot and subplot strata. The last
level holds the individual sub-subplot observations. Each arrow marks one act of
randomization, and an effect belongs to the stratum created by the act that
assigned it.

| Effect | Stratum |
|---|---|
| whole plot | whole-plot error, $R{:}A$ |
| subplot, and its interaction with the whole plot | subplot error, $R{:}A{:}B$ |
| sub-subplot, and every interaction involving it | sub-subplot error, the remainder |

Precision increases as you descend. The three-way interaction is the most
precisely tested term in the entire experiment, and the whole-plot main effect
the least.

## 5.3 Why the declaration cannot be partial

```{r splitsplit-refused, error = TRUE}
np_splitsplit(yield ~ irrigation * cultivar * timing, ssp,
              block = block, whole_plot = irrigation, subplot = cultivar)
```

Omitting a level would leave the analysis unable to identify which stratum an
effect belongs to. The package refuses rather than assuming.

## 5.4 A practical warning about size

```{r splitsplit-units}
data.frame(
  level = c("whole plot", "subplot", "sub-subplot"),
  independent_units = c(
    length(unique(interaction(ssp$block, ssp$irrigation, drop = TRUE))),
    length(unique(interaction(ssp$block, ssp$irrigation, ssp$cultivar,
                              drop = TRUE))),
    nrow(ssp))
)
```

The whole-plot factor is being compared across a handful of units, whatever the
total plot count suggests. Read that table before designing the experiment, not
after analysing it. If the factor of primary interest is the whole-plot factor, a
split-split-plot is usually the wrong design for it.

---

# 6. Strip-plot designs

## 6.1 A different physical situation

Both factors are applied in strips across the whole block, one set of strips
running in each direction. The plots are the intersections.

```{r strip-fit}
strip <- simulate_agri("strip_plot", seed = 503, n = 5)

fit_strip <- np_stripplot(
  yield ~ irrigation * nitrogen,
  data    = strip,
  block   = block,
  strip_a = irrigation,
  strip_b = nitrogen,
  method  = "auto"
)

fit_strip
```

```{r strip-anova}
anova(fit_strip)
```

## 6.2 Why the interaction has its own stratum

In a split-plot, the subplot factor and the interaction are both comparisons
among subplots inside a whole plot, so they share an error term.

In a strip-plot, neither factor is nested inside the other. Each main effect is a
comparison among strips in its own direction, and the interaction is a comparison
among **intersections**, which is a third kind of unit. With block $R$ and the two
strip factors $A$ and $B$, the primary strata are

$$A: R{:}A, \qquad B: R{:}B, \qquad A \times B: \text{the intersection remainder}.$$

Note that these are **crossed**, not nested: $R{:}A$ and $R{:}B$ sit side by side
rather than one inside the other. That is the structural difference from the
split-plot, and it is the reason the interaction cannot borrow either main
effect's denominator.

| Design | Strata | Interaction tested against |
|---|---|---|
| split-plot | 2 | the subplot error, shared with the subplot factor |
| strip-plot | 3 | its own residual, shared with nothing |

## 6.3 The consequence of confusing them

A strip-plot analysed as a split-plot tests the interaction against the wrong
denominator. Depending on which stratum is larger, the interaction becomes either
too easy or too hard to detect, and the direction of the error cannot be
predicted from the output.

## 6.4 Recognising a strip-plot in the field plan

Ask two questions:

1. Was factor A applied in strips running one way across the block?
2. Was factor B applied in strips running the **other** way?

If both answers are yes, it is a strip-plot, whatever the analysis in the
previous paper on the same trial did.

## 6.5 Interpretation

Report all three strata and their degrees of freedom. A strip-plot reported with
two strata has been analysed as something else.

## 6.6 Where the package stops

A note on what these three functions will not do for you. The package does not
silently coerce an incomplete or structurally ambiguous hierarchical experiment
into a simpler independent design. If plots are missing, or if the declaration
does not identify which stratum an effect belongs to, the analysis is refused
rather than approximated. Missingness and imbalance have to be judged against the
**actual randomization strata**, not against the data frame, and no function here
can make that judgement on your behalf.

This is a deliberate boundary. A package that quietly falls back to a factorial
when the split-plot declaration is incomplete would produce a table that looks
right and is not.

---

# 7. An engine the package refuses, and why

`permuco` is **not admissible** for any design with nested field strata. The
package refuses it for split-plot, split-split-plot and strip-plot, and the
automatic mode selects ART instead.

The cause is not a programming error in `permuco`. It is a mismatch of scope.
`permuco::aovperm` implements the repeated-measures `Error(subject/within)`
form, in which each subject contributes one observation per within-cell. A field
split-plot carries several sub-units inside every whole plot, so without one
observation per cell the sub-plot stratum is never built, and the terms that
belong to it are tested against the whole-plot mean square.

Under a true null, across 300 simulated experiments, the sub-plot and
sub-sub-plot factors were rejected **not once** at either the 5% or the 10%
level, and the smallest p-value observed for the sub-sub-plot factor was 0.18.
The test was not conservative, it was switched off.

```{r perm-warning}
data.frame(
  design = c("factorial", "rcbd", "split-plot", "split-split-plot", "strip-plot"),
  permuco_admissible = c("yes", "yes", "NO", "NO", "NO"),
  use_instead = c("-", "-", "ART", "ART", "ART")
)
```

Two words in that table were chosen deliberately. **Admissible**, not
*calibrated*, because this is a decision the package enforces rather than an
empirical result that more simulation might revise. And **ART** alone, not
"rankFD or ART", because `rankFD` does not represent nested field strata either.
Recommending it here would repeat the very error this section documents.

The direction of the failure matters for anyone revisiting an old analysis. The
inflated denominator produces **false negatives**, not false positives. If you
ran one of these three designs through `permuco` in an earlier version, the
results to re-examine are the ones that came out non-significant.

---

# Part II. Ordered quantitative treatments

# 8. When the levels have an order, use it

## 8.1 The situation

A trial applies nitrogen at 0, 40, 80, 120 and 160 kg/ha. Those are not five
unrelated categories. They lie on a line, and the scientific question is almost
never "do the five differ" but "does yield rise with rate, and how".

Treating an ordered treatment as unordered throws away that information, and the
cost is power. An unordered test spends four degrees of freedom looking for any
pattern; a trend test spends one looking for the pattern you expect.

```{r trend-data}
set.seed(601)
trend_dat <- expand.grid(
  block = factor(1:6),
  dose  = c(0, 40, 80, 120, 160)
)

trend_dat$yield <-
  4.5 +
  0.045 * trend_dat$dose -
  0.00012 * trend_dat$dose^2 +
  0.15 * as.numeric(trend_dat$block) +
  rt(nrow(trend_dat), df = 5)

des_trend <- agri_design(
  yield ~ dose,
  trend_dat,
  design       = "rcbd",
  block        = block,
  quantitative = dose
)
des_trend
```

Note `quantitative = dose`. That argument is the declaration that the levels are
ordered and numeric, and it is what makes the trend test available.

## 8.2 The trend test

```{r trend-fit}
tr <- agri_trend(
  des_trend,
  treatment = dose,
  B         = 1999,
  seed      = 601
)
tr
```

## 8.3 What the test does, and what it does not

The procedure assigns **scores** to the ordered levels and tests whether the
within-block ranks move systematically with those scores. By default the scores
are the dose values themselves, which targets a monotone trend.

| It answers | It does not answer |
|---|---|
| does the response move systematically with dose | what shape the response has |
| in which direction | where the maximum is |
| with what strength of evidence | what to recommend |

That last column is the important one. A trend test is a **screening** device for
an ordered factor. It is not a description of the response, and it cannot support
a recommendation.

## 8.4 The specific trap: a hump-shaped response

The data above were generated with a quadratic term, so yield rises and then
flattens. A linear trend test can miss such a response entirely when the rise and
the fall roughly cancel.

```{r trend-hump}
set.seed(605)
hump <- expand.grid(block = factor(1:6), dose = c(0, 40, 80, 120, 160))
# A strong hump with its peak in the middle of the range.
hump$yield <- 5 + 0.09 * hump$dose - 0.00056 * hump$dose^2 +
  0.15 * as.numeric(hump$block) + rnorm(nrow(hump), 0, 0.5)

des_hump <- agri_design(yield ~ dose, hump, design = "rcbd",
                        block = block, quantitative = dose)
agri_trend(des_hump, treatment = dose, B = 999, seed = 605)
```

```{r trend-hump-means}
round(tapply(hump$yield, hump$dose, mean), 2)
```

Read the two together. The means show an unmistakable hump. Whether the linear
trend test detects anything depends on how symmetric that hump is about the
centre of the dose range, and a perfectly symmetric one would give a trend
statistic near zero.

## 8.5 The correct escalation

| Question | Tool | Vignette |
|---|---|---|
| does the response move with dose at all | `agri_trend()` | this one |
| what shape does it have | `agri_np_regression()` | *Nonparametric Regression* |
| where does it stop rising | `agri_np_sizer()` | *Distribution-Free Uncertainty* |
| what rate to recommend, with an interval | `agri_np_optimum_test()` | *Optima, Quantiles, and Block Structure* |

A trend test is the first rung of that ladder, not a substitute for it.

## 8.6 Choosing scores

The `scores` argument controls what "ordered" means.

| Scores | Targets | Use when |
|---|---|---|
| the dose values | a linear response in dose | doses are on a meaningful scale |
| ranks of the doses | any monotone response | the spacing is arbitrary |
| log doses | a linear response in log dose | the response is proportional |

Choosing the scores after seeing which gives the smaller p-value is the same
error as choosing an engine that way. Decide from the biology.

## 8.7 Interpretation

Report the scores used and the number of permutations. A trend statistic without
its scores is not reproducible, because a different scoring can reverse the
conclusion.

---

# Part III. Covariance adjustment

# 9. When a measured covariate can help

## 9.1 The situation

Blocking removes variation the experimenter could **anticipate** and lay out.
A covariate removes variation that could only be **measured**: initial plant
size, soil organic matter, stand count at emergence, pre-treatment disease
level.

```{r ancova-data}
set.seed(602)
anc <- expand.grid(
  block     = factor(1:6),
  treatment = factor(c("control", "bio", "chemical")),
  rep       = 1:2
)

anc$initial_biomass <- runif(nrow(anc), 8, 15)
anc$final_biomass <-
  10 +
  1.6 * anc$initial_biomass +
  c(control = 0, bio = 2, chemical = 4)[anc$treatment] +
  0.3 * as.numeric(anc$block) +
  rt(nrow(anc), 5)
str(anc)
```

## 9.2 The adjustment

```{r ancova-fit}
if (requireNamespace("permuco", quietly = TRUE)) {
  anc_fit <- agri_ancova(
    final_biomass ~ treatment,
    data           = anc,
    covariates     = initial_biomass,
    block          = block,
    np             = 1999,
    seed           = 602,
    rank_response  = TRUE
  )
  print(anc_fit$omnibus)
}
```

## 9.3 What it buys

```{r ancova-without}
des_noanc <- agri_design(final_biomass ~ treatment, anc, design = "rcbd",
                         block = block)
anova(agri_rank(des_noanc, method = "ART"))
```

The covariate absorbs the initial-size variation, which would otherwise sit in
the residual. Where the covariate is strongly related to the response, as here,
the gain in precision can be large.

## 9.4 The condition that makes it legitimate

This is the part most often violated, and it is not statistical but temporal.

> **The covariate must be unaffected by the treatment.**

The safe rule: a covariate measured **before** the treatment was applied is
legitimate. One measured afterwards requires an argument that the treatment could
not have influenced it, and that argument is usually not available.

| Covariate | Measured | Legitimate |
|---|---|---|
| plant height at transplanting | before treatment | yes |
| soil organic matter at sowing | before treatment | yes |
| stand count at emergence | before treatment, usually | yes, if treatment began later |
| leaf area at harvest | after treatment | no |
| disease severity at flowering | after treatment | no |

## 9.5 Why adjusting for an affected covariate is worse than not adjusting

If the treatment changed the covariate, adjusting for it removes part of the
treatment effect itself. The adjusted comparison then answers a question nobody
asked: what the treatments would have done had they not changed the covariate,
which is a counterfactual the experiment cannot support.

The result is usually an attenuated and biased treatment effect, in a direction
that cannot be signed without knowing the causal structure.

## 9.6 A diagnostic worth running

```{r ancova-check}
# Does the covariate differ across treatments? If it does, and it was measured
# after treatment began, the adjustment is not admissible.
round(tapply(anc$initial_biomass, anc$treatment, mean), 3)
kruskal.test(initial_biomass ~ treatment, data = anc)$p.value
```

Here the covariate does not differ across treatments, which is what randomization
should produce for a pre-treatment measurement. A strong difference in a
pre-treatment covariate is worth investigating: it can indicate that the
randomization was not executed as planned.

## 9.7 `rank_response` and what it means

Setting `rank_response = TRUE` ranks the response before the adjustment, which
keeps the analysis in the rank-based family and protects against a skewed
response or an extreme plot.

The trade-off is that the adjustment is then linear in the covariate **on the
rank scale**, which is a different assumption from linearity on the original
scale. Where the relationship is strongly non-linear, a nonparametric regression
on the covariate is a better route than an ANCOVA of either kind.

## 9.8 Interpretation

Report the covariate, when it was measured, the evidence that it was unaffected
by treatment, and whether the response was ranked. Those four facts are what make
the adjustment checkable.

---

# Part IV. Power

# 10. The question, put correctly

"How many replicates do I need" has no answer. "How many replicates do I need to
detect a difference of agronomic size D with probability P, given material of
variability S" has one.

The four quantities are inseparable. Fixing any three determines the fourth.

## 10.1 Why this package makes you write the assumption down

```{r power-fit}
generator <- function(i) {
  simulate_agri("rcbd", seed = 8000 + i, n = 6)
}

analyzer <- function(dat) {
  np_rcbd(yield ~ treatment, dat, block = block, method = "friedman")
}

pwr_demo <- agri_power(
  generator = generator,
  analyzer  = analyzer,
  nsim      = 100,
  alpha     = 0.05,
  seed      = 603
)

pwr_demo
```

`agri_power()` takes two **functions** rather than a formula and a few numbers.
That is deliberate.

A closed-form power calculation hides its data-generating assumption inside the
formula. Normal errors, equal variances, no ties, no missing plots. Field
material frequently satisfies none of those, and the resulting power estimate is
optimistic in a way the analyst never sees.

Writing the generator as code forces the assumption into the open, where a
co-author or a reviewer can disagree with it.

## 10.2 Varying the design

```{r power-vary}
make_gen <- function(nblocks) function(i)
  simulate_agri("rcbd", seed = 9000 + i, n = nblocks)

pw <- do.call(rbind, lapply(c(3, 5, 8), function(nb) {
  r <- agri_power(make_gen(nb), analyzer, nsim = 60, alpha = 0.05, seed = 1)
  data.frame(blocks = nb, power = round(r$power, 3))
}))
pw
```

## 10.3 Reading a power table

Power rises with replication, steeply at first and then slowly. The practical
reading is usually the reverse of the intended one: not "how much power do I get
from eight blocks", but **"below how many blocks is this experiment not worth
running"**.

## 10.4 The most common design failure

Four blocks, three treatments, and a difference of agronomic interest that is
small relative to plot-to-plot variation. The experiment will almost certainly
return a non-significant result, and that result distinguishes nothing: it is
compatible with no effect and with an effect of practical importance.

The remedy is a design decision taken before the field is laid out. No method in
this package repairs it afterwards.

## 10.5 Post-hoc power is not a remedy

Computing power from the observed effect size, after a non-significant result, is
circular: it is a monotone function of the p-value and adds no information.

What is legitimate, and useful, is to report the effect size the design **could**
have detected. That is computed from the design and the assumed variability, not
from the observed result, and it turns "we found nothing" into "we could not have
found anything smaller than D".

## 10.6 Interpretation

Report the generator's assumptions, the number of simulations, the seed, and the
detectable difference. A power statement without its data-generating assumption
is not checkable.

---

# Part V. Common mistakes, and the function that prevents each

# 11. Testing a whole-plot factor against the subplot error

**The mistake.** Analysing a split-plot as a factorial in blocks.

**Why it is wrong.** The whole-plot factor is tested against a residual belonging
to a different randomization level, which is much smaller. The effect becomes far
too easy to declare significant. See section 4.6.

**What prevents it.** `np_splitplot()` with `whole_plot=` and `subplot=` records
both units and uses the right stratum for each effect.

---

# 12. Calling a strip-plot a split-plot

**The mistake.** Using `np_splitplot()` when both factors were applied in
crossing strips.

**Why it is wrong.** The interaction has its own error stratum in a strip-plot
and does not in a split-plot. The interaction is tested against the wrong
denominator, in a direction that cannot be predicted. See section 6.

**What prevents it.** `np_stripplot()` with `strip_a=` and `strip_b=`.

---

# 13. Declaring a hierarchical design partially

**The mistake.** Omitting `subsubplot=` from a split-split-plot.

**Why it is wrong.** The analysis cannot identify which stratum an effect belongs
to.

**What prevents it.** The declaration is refused by name. See section 5.3.

---

# 14. Using the permutation backend in split-plot, split-split or strip-plot designs

**The mistake.** Choosing `method = "permuco"` for those three designs because
permutation tests are assumption-light.

**Why it is wrong.** The backend does not build the sub-plot stratum at all, so
the terms belonging to it are tested against the whole-plot mean square. Across
300 null replicates it produced zero rejections at 5% and at 10%. The test is
not merely conservative, it never rejects. The failure is a false negative, so
`rankFD` is no substitute either, since it does not represent nested strata.
Use `ART`.

**What prevents it.** The package refuses the call. `method = "permuco"` on any
of the three designs raises an error naming the violated assumption, and the
automatic mode selects `ART`. In earlier versions nothing prevented it, and
`permuco` was in fact the first automatic choice, which is why section 7 asks
you to revisit analyses run before this release.

---

# 15. Treating an ordered dose as unordered

**The mistake.** Comparing five nitrogen rates as five unrelated categories.

**Why it is wrong.** It spends four degrees of freedom looking for any pattern
when one would do, and reports letters instead of a shape.

**What prevents it.** `quantitative = dose` in the declaration makes
`agri_trend()` available, and the regression vignettes take it further. See
section 8.

---

# 16. Reading a trend test as a description of the response

**The mistake.** "A significant linear trend was found, so yield increases with
nitrogen up to 160 kg/ha."

**Why it is wrong.** A trend test says the response moves systematically. It says
nothing about the shape, and it can be significant for a response that rises then
falls. See section 8.4.

**What prevents it.** The escalation table of section 8.5.

---

# 17. Adjusting for a covariate the treatment affected

**The mistake.** Using leaf area at harvest as a covariate for yield.

**Why it is wrong.** The adjustment removes part of the treatment effect, and the
adjusted comparison answers a counterfactual the experiment cannot support. See
section 9.5.

**What prevents it.** Only the analyst's judgement, informed by section 9.4. The
package cannot know when the covariate was measured.

---

# 18. Reporting post-hoc power

**The mistake.** Computing power from the observed effect after a
non-significant result.

**Why it is wrong.** It is a monotone function of the p-value and adds no
information. See section 10.5.

**What prevents it.** `agri_power()` requires a generator, which forces the
effect size to be stated independently of the data.

---

# Part VI. Compact selection guide

# 19. Choose the design function by the randomization

| The field plan says | Use | Declare |
|---|---|---|
| one factor applied to large areas, a second inside them | `np_splitplot()` | `block`, `whole_plot`, `subplot` |
| three nested levels of application | `np_splitsplit()` | plus `subsubplot` |
| both factors in crossing strips | `np_stripplot()` | `block`, `strip_a`, `strip_b` |
| everything randomized at one scale | `np_factorial()` or `np_rcbd()` | `block` if present |

# 20. Choose the extra analysis by the question

| The question | Use |
|---|---|
| does the response move with an ordered dose | `agri_trend()` |
| can a pre-treatment measurement sharpen the comparison | `agri_ancova()` |
| is this design big enough | `agri_power()`, before the field is laid out |
| what shape does the response have | `agri_np_regression()`, in the regression vignettes |

---

# Part VII. Minimum reporting checklist

# 21. What the methods section must contain

1. the design, named, with every randomization level identified;
2. the experimental unit for each factor;
3. the degrees of freedom of every error stratum;
4. the engine, and for split-plot, split-split or strip-plot designs, why `permuco` was not
   used;
5. for a trend test: the scores and the number of permutations;
6. for an ANCOVA: the covariate, when it was measured, and evidence that it was
   unaffected by treatment;
7. for a power statement: the generator's assumptions, the number of
   simulations, and the seed;
8. the package version.

# 22. A worked methods paragraph

> The experiment was a split-plot in five randomized complete blocks, with
> irrigation randomized to whole plots and cultivar to subplots within them. The
> whole-plot experimental unit was the irrigation strip and the subplot unit the
> individual plot. Analysis used design-aware rank-based inference
> (`np_splitplot()` from agriRank 0.14.0), which tests each effect against the
> error stratum of the unit to which it was randomized: irrigation against the
> whole-plot error, cultivar and the interaction against the subplot error.
> Initial plant biomass, measured before treatments were applied, was used as a
> covariate (`agri_ancova()`, 1999 permutations, seed 602); it did not differ
> across treatments (p = 0.43), consistent with correct randomization.

---

# 23. Where to go next

| If you now want | Read |
|---|---|
| the design declaration in detail | *Design Foundations, CRD, and RCBD* |
| comparisons and compact letters | *Effects, Conover, Contrasts, and Factorial Inference* |
| the shape of the dose response | *Nonparametric and Shape-Aware Regression* |
| a rate to recommend, with an interval | *Optima, Quantiles, and How the Block Enters the Model* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# Part VIII. Glossary

# 24. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **whole plot** | the larger unit, to which the first factor was randomized |
| **subplot** | the smaller unit inside a whole plot |
| **sub-subplot** | the smallest unit, in a three-level design |
| **strip** | a band across the whole block, to which a factor was applied |
| **error stratum** | the residual variation appropriate to one randomization level |
| **denominator** | the error stratum against which an effect is tested |
| **ordered treatment** | levels lying on a meaningful scale, not merely labelled |
| **scores** | the numeric values assigned to ordered levels in a trend test |
| **covariate** | a measured variable used to remove variation, not a treatment |
| **pre-treatment covariate** | one measured before treatments were applied |
| **power** | the probability of detecting a stated effect, given the design |
| **detectable difference** | the smallest effect the design could find with stated probability |
| **calibration** | whether a test rejects at its nominal rate under the null |

---

# Selected methodological references

On the designs themselves:

- Federer, W. T., and King, F. (2007). *Variations on Split Plot and Split Block Experiment Designs*. Wiley.
- Gomez, K. A., and Gomez, A. A. (1984). *Statistical Procedures for Agricultural Research*, 2nd edition. Wiley.

On the rank-based and permutation machinery used here:

- Brunner, E., Konietschke, F., Bathke, A. C., and Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349-366. https://doi.org/10.1111/insr.12418
- Konietschke, F., and Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142-158. https://doi.org/10.32614/RJ-2023-029
- Pauly, M., Brunner, E., and Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461-473. https://doi.org/10.1111/rssb.12073
- Wobbrock, J. O., Findlater, L., Gergle, D., and Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *Proceedings of CHI 2011*, 143-146. https://doi.org/10.1145/1978942.1978963
- Elkin, L. A., Kay, M., Higgins, J. J., and Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *Proceedings of UIST 2021*, 754-768. https://doi.org/10.1145/3472749.3474784

On the engine discussed in section 7, cited so that the reader can check the
scope of the method for themselves:

- Frossard, J., and Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). https://doi.org/10.18637/jss.v099.i15

The package also ships a verified RIS library under `inst/references/`.
