---
title: "Repeated Measures and Missing Longitudinal Data"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repeated Measures and Missing Longitudinal Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.8,
  fig.height = 4.4,
  dpi = 130
)
```

**Repeated measures vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** measurements taken more than once on the same unit, and what to do
when some of them are absent.

---

# 1. Why this vignette exists

Measuring the same plant four times does not give four independent
observations. It gives one plant, observed four times, and the difference
between those two statements is the whole content of this vignette.

The dependence is not a nuisance to be removed. It is information: a plant that
starts tall tends to stay tall, and a trajectory is a richer datum than an
endpoint. But it must be represented, and an analysis that treats the four
measurements as four replicates inflates the apparent replication fourfold, with
every interval correspondingly too narrow.

> **Declare the subject. Analyse trajectories, not endpoints. When values are
> missing, describe the pattern before modelling it, and report the assumption
> as part of the methods.**

## 1.1 The second half of the vignette

Longitudinal agronomic data are almost never complete. A plant dies, a plot is
flooded, a measurement day is rained out, a leaf is lost to a herbivore.

Missingness is not a technical inconvenience to be imputed away. **Why** a value
is absent determines which analyses remain valid, and that question is
scientific, not computational.

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. distinguish a repeated-measures design from a design with more replicates;
2. declare `subject` and `within` correctly, and explain what each asserts;
3. explain why an omnibus repeated-measures test answers a question about
   trajectories rather than endpoints;
4. choose among the available backends on structural grounds;
5. read a missing-data report and say what it does and does not establish;
6. distinguish MCAR, MAR and MNAR, and name a plausible agronomic mechanism for
   each;
7. explain why the choice among them cannot be made from the data alone;
8. run a missing-data sensitivity analysis and interpret its output;
9. recognise the experimental label on the native wild-bootstrap engine and say
   what it means for a manuscript;
10. recognise the deliberate refusal of an additional block stratum in the
    repeated-measures workflow, and why it is preferable to a silent
    approximation.

---

# 3. The repeated-measures module in one map

| Function | What it does | Requires |
|---|---|---|
| `agri_design(design = "repeated")` | declare subject and within factors | `subject`, `within` |
| `np_repeated()` | the one-step wrapper | the same |
| `agri_repeated()` | fit with an explicit backend | a declared design |
| `agri_missing_report()` | describe the missingness pattern | a declared design |
| `agri_missing_sensitivity()` | does the conclusion survive the assumption | a declared design |
| `incomplete_wild_rank_test()` | the native experimental engine | a declared design |
| `agri_plot(type = "missing")` | see where the gaps are | a fitted object |

## 3.1 Which backend, and when

```{r backend-map}
data.frame(
  backend = c("nparLD", "MANOVA.RM", "native_wild"),
  handles_missing = c("no", "no", "yes"),
  status = c("established", "established", "EXPERIMENTAL"),
  use_when = c("complete data, classical rank ATS",
               "complete data, resampling WTS and MATS",
               "incomplete data, and only after reading section 12")
)
```

---

# Part I. Complete repeated measures

# 4. Declaring the design

```{r rm-design}
rm <- simulate_agri(
  "repeated",
  seed = 701,
  n    = 10
)
str(rm)

des_rm <- agri_design(
  height ~ treatment * time,
  rm,
  design  = "repeated",
  subject = subject,
  within  = time
)
des_rm
```

## 4.1 What `subject` asserts

`subject` names the unit that was measured repeatedly. Declaring it asserts:

1. that measurements sharing a subject are **not** independent;
2. that the subject, not the measurement, is the replicate;
3. and therefore that the treatment comparison has as many independent units as
   there are subjects, not as there are rows.

## 4.2 What `within` asserts

`within` names the factor that varies **inside** a subject. Time is the usual
one, but a within factor can also be a leaf position, a soil depth, or a
sampling date.

The distinction matters because a between-subject factor and a within-subject
factor are tested with different precision, for the same reason a whole-plot and
a subplot factor are in a split-plot. A within-subject comparison is made inside
the subject and is therefore free of between-subject variation.

| Factor | Varies | Compared | Typically |
|---|---|---|---|
| `treatment` | between subjects | across subjects | less precise |
| `time` | within subjects | inside each subject | more precise |
| `treatment:time` | within subjects | inside each subject | more precise |

## 4.3 The declaration cannot be inferred

```{r rm-refused, error = TRUE}
np_repeated(height ~ treatment * time, rm)
```

Nothing in the data file distinguishes forty independent plants from ten plants
measured four times. The subject column exists in both cases; only the
experimenter knows which it is.

## 4.4 Interpretation

Write down how many independent subjects the experiment has, before analysing
anything. If that number is small, the trajectory analysis will be imprecise
whatever the total row count suggests.

---

# 5. Fitting a complete repeated-measures analysis

```{r rm-fit}
fit_rm <- agri_repeated(
  des_rm,
  backend = "auto",
  B       = 999,
  iter    = 999,
  seed    = 701
)

fit_rm
```

## 5.1 The three effects, and what each asks

| Effect | Question |
|---|---|
| `treatment` | averaged over time, do the treatments differ |
| `time` | averaged over treatments, does the response change through the cycle |
| `treatment:time` | do the treatments follow different trajectories |

The interaction is read first, as always. If present, the main effects are
averages over conditions in which the behaviour differs.

## 5.2 A repeated-measures analysis is about trajectories

This is worth stating plainly, because it is the commonest misuse.

A repeated-measures analysis answers questions about the **shape of the response
over time**. If the scientific question concerns the final value, the final value
should be analysed as a single response in its own design, which is simpler and
usually more powerful.

| The question is | Analyse |
|---|---|
| does the treatment change final yield | the final measurement, as an RCBD |
| does the treatment change how fast plants grow | the trajectory, as repeated measures |
| does the treatment delay senescence | the trajectory |
| does the treatment change the maximum reached | the maximum per subject, as an RCBD |

Running a repeated-measures analysis to answer an endpoint question spends
degrees of freedom on time effects nobody asked about.

## 5.3 The figure

```{r rm-plot, fig.cap = "Trajectories by treatment. Each line is a treatment mean over subjects; the question is whether the lines are parallel."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(
    ggplot2::ggplot(rm, ggplot2::aes(x = time, y = height,
                                     colour = treatment, group = treatment)) +
      ggplot2::stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
      ggplot2::stat_summary(fun = mean, geom = "point", size = 2) +
      ggplot2::labs(x = "Occasion", y = "Height", colour = "Treatment") +
      agri_theme()
  )
}
```

---

# 6. Choosing a backend

## 6.1 nparLD

```{r nparld}
if (requireNamespace("nparLD", quietly = TRUE)) {
  fit_nparld <- agri_repeated(des_rm, backend = "nparLD", seed = 701)
  print(fit_nparld)
}
```

`nparLD` implements the classical rank-based analysis of longitudinal data. It
computes relative treatment effects and an ANOVA-type statistic, and it is the
established reference for complete designs of this shape.

## 6.2 MANOVA.RM

```{r manovarm}
if (requireNamespace("MANOVA.RM", quietly = TRUE)) {
  fit_mrm <- agri_repeated(des_rm, backend = "MANOVA.RM", iter = 999,
                           seed = 701)
  print(fit_mrm)
}
```

`MANOVA.RM` provides a Wald-type statistic with a resampling reference
distribution, and a modified ANOVA-type statistic. The resampling matters: the
Wald-type statistic is known to be liberal in small samples, and the resampled
version corrects it.

## 6.3 Which to use

| Situation | Prefer |
|---|---|
| complete data, one within factor | `nparLD` |
| complete data, small samples | `MANOVA.RM` with resampling |
| complete data, several responses | `agri_multivariate()`, see the multivariate vignette |
| incomplete data | `native_wild`, having read section 12 |

## 6.4 Comparing them is legitimate; choosing on p is not

Running two backends and reporting the agreement is a sensitivity analysis.
Running two and reporting the one that reached significance is a selection
procedure whose properties are unknown. The rule is the same throughout this
package.

---

# Part II. Missing longitudinal data

# 7. Describe before modelling

```{r missing-data}
rm_miss <- simulate_agri(
  "repeated_missing",
  seed         = 801,
  n            = 12,
  missing_rate = 0.15
)

des_miss <- agri_design(
  height ~ treatment * time,
  rm_miss,
  design  = "repeated",
  subject = subject,
  within  = time
)

miss_report <- agri_missing_report(des_miss)
miss_report
```

## 7.1 Three separate questions

A single overall percentage answers none of these well:

| Question | Why it matters |
|---|---|
| how much is missing | determines how much power remains |
| where is it concentrated | one treatment losing plants is a finding, not a nuisance |
| does the pattern relate to the design | determines which assumptions are tenable |

## 7.2 The figure

```{r missing-plot, fig.cap = "Where the gaps are. Concentration in one treatment or at one occasion is the pattern to look for."}
agri_plot(
  agri_rank(
    des_miss,
    method             = "incomplete_wild",
    B                  = 299,
    seed               = 801,
    missing_assumption = "MCAR"
  ),
  type = "missing"
)
```

## 7.3 What the report establishes, and what it cannot

The report describes the **observed** pattern. It cannot establish why values are
absent, because the reason involves the values that were not recorded.

That limitation is fundamental rather than technical, and it is the subject of
the next section.

---

# 8. MCAR, MAR, MNAR

## 8.1 The three mechanisms

| Assumption | Means | An agronomic mechanism |
|---|---|---|
| **MCAR** | missingness is unrelated to anything | a technician dropped a tray; a measurement day was rained out for the whole trial |
| **MAR** | missingness depends on **observed** values | plants below a height threshold at week 2 were not measured at week 4, and week 2 is recorded |
| **MNAR** | missingness depends on the **unobserved** value itself | plants died because they were about to be small, and the size they would have reached is unrecorded |

## 8.2 Why the choice cannot be made from the data

MAR and MNAR are **not distinguishable** from the observed data. Both produce the
same observed pattern; they differ in what happened to the values you do not
have.

This is not a limitation of any particular method. No amount of computation
recovers information that was never recorded, and any procedure claiming to
choose among them from the data alone is smuggling in an assumption.

## 8.3 What follows

The assumption is a **scientific** statement about the mechanism, made from
knowledge of the experiment, and it belongs in the methods section beside the
design.

| If the mechanism was | State |
|---|---|
| an equipment failure or a rained-out day | MCAR, and say why |
| a protocol that skipped plants based on a recorded earlier value | MAR, and name the value |
| plant death related to the treatment | MNAR, and say the analysis is a sensitivity analysis rather than an estimate |

## 8.4 The most consequential case in agronomy

Plant death is frequently **MNAR**, and it is frequently treated as MCAR.

If a treatment kills the weakest plants, the survivors in that treatment are a
selected subset, and any analysis of survivors overstates the treatment's
performance. The missingness is then not a nuisance at all: it is one of the main
results, and it should be analysed as a survival outcome in its own right.

## 8.5 Interpretation

Report the assumption, the reason for it, and a sensitivity analysis under an
alternative. An analysis that does not state its missingness assumption has made
one anyway.

---

# 9. The native incomplete engine

```{r wild-fit}
fit_miss <- incomplete_wild_rank_test(
  des_miss,
  B                  = 1999,
  seed               = 801,
  statistic          = "ATS",
  missing_assumption = "MCAR"
)

fit_miss
```

## 9.1 What it does

The engine implements rank-based quadratic-form inference for incompletely
observed repeated measurements, with a wild-bootstrap reference distribution. It
uses the observed cells directly rather than imputing the absent ones, which is
what makes it applicable when a subject has some but not all occasions.

Note that `missing_assumption` is a **required** argument with no silent default
behaviour. The engine will not proceed without an explicit statement of what is
being assumed, which is the point.

## 9.2 The statistic

| `statistic` | Is | Behaves |
|---|---|---|
| `"ATS"` | ANOVA-type statistic | conservative, robust in small samples |
| `"WTS"` | Wald-type statistic | liberal in small samples unless resampled |

For agronomic sample sizes, `"ATS"` with resampling is the safer default.

---

# 10. Sensitivity to the assumption

```{r missing-sens}
sens_miss <- agri_missing_sensitivity(
  des_miss,
  B         = 999,
  seed      = 802,
  statistic = "ATS"
)

sens_miss
```

## 10.1 This is not imputation

An imputation procedure fills the gaps and then analyses the completed data as
if it were observed. A sensitivity analysis does something different and more
honest: it asks whether the conclusion **changes** under different assumptions
about why values are absent.

| | Imputation | Sensitivity |
|---|---|---|
| Produces | one completed dataset | several analyses |
| Claims | the filled values are plausible | nothing about the filled values |
| Answers | what is the estimate | is the estimate robust to the assumption |
| Fails when | the mechanism is MNAR | never, but it may report instability |

## 10.2 How to read the output

| Outcome | Reading |
|---|---|
| conclusions agree across assumptions | the result does not rest on the missingness assumption; say so in one sentence |
| conclusions differ | **that is the finding**; report all of them and state which was pre-specified |

The second case is uncomfortable and it is where most of the value lies. A
conclusion that survives only under MCAR, in an experiment where plants died, is
not a conclusion.

## 10.3 Interpretation

Report the primary assumption, the sensitivity range, and the mechanism you
believe operated. Those three together are what a reviewer needs.

---

# Part III. Two deliberate refusals

# 11. An additional block stratum is refused

```{r block-refused, error = TRUE}
blocked_rm <- rm_miss
blocked_rm$block <- factor(rep(1:4, length.out = nrow(blocked_rm)))

blocked_des <- agri_design(
  height ~ treatment * time,
  blocked_rm,
  design  = "repeated",
  subject = subject,
  within  = time,
  block   = block
)

agri_rank(blocked_des, method = "auto")
```

## 11.1 Why the refusal is preferable

A repeated-measures experiment laid out in blocks has **three** levels of
structure: the block, the subject inside the block, and the occasion inside the
subject. The rank-based engines available here represent two of them.

Two responses are possible. Accept the block and treat it approximately, or
refuse and say so. The package refuses, for a specific reason: an approximation
that is not validated would produce a p-value with unknown calibration, and the
analyst would have no way to detect it.

## 11.2 What to do instead

| Option | When it is defensible |
|---|---|
| drop the block from the declaration | when between-block variation is demonstrably small; report the decision |
| analyse each block separately and combine | when blocks are few and interpretable |
| analyse a derived endpoint in the full design | when the scientific question is about an endpoint anyway |
| use a mixed model outside this package | when the subject-and-block structure is the scientific object |

The last option is not a defeat. This package is deliberately narrow, and a
three-level longitudinal structure is a legitimate place to leave it.

---

# 12. The experimental label

The native incomplete repeated-measures engine is labelled **experimental** in
the package documentation, and it keeps that label until the validation plan in
`VALIDATION_PLAN.md` is complete.

## 12.1 What remains to be verified

```{r validation-items}
data.frame(
  item = c("intermediate rank quantities", "relative effects",
           "covariance matrices", "WTS/ATS/MATS",
           "frozen wild-bootstrap sequences", "Type-I error",
           "power", "missingness rates", "ordinal data", "ties",
           "heteroscedasticity", "row and label invariance",
           "comparison with reference code"),
  status = "pending"
)
```

## 12.2 What the label means in practice

Running without error is **not** evidence of correct Type-I error. An engine can
produce plausible p-values that are systematically too small or too large, and
nothing in the output reveals it.

For a manuscript, the defensible positions are:

1. use an established backend on complete data, and report the missingness
   separately;
2. or use the native engine and describe it as experimental, with the
   sensitivity analysis of section 10 as the primary evidence.

What is not defensible is using it without the label.

---

# Part IV. Common mistakes, and the function that prevents each

# 13. Counting measurements as replicates

**The mistake.** Ten plants measured four times, analysed as forty observations.

**Why it is wrong.** The experiment has ten independent units. Every interval is
too narrow by roughly a factor of two.

**What prevents it.** `agri_design(design = "repeated", subject =)` records the
unit, and the analysis uses it. See section 4.

---

# 14. Running a repeated-measures analysis for an endpoint question

**The mistake.** Analysing four occasions when the question concerns final
yield.

**Why it is wrong.** It spends degrees of freedom on time effects nobody asked
about, and the answer to the endpoint question is buried in an interaction.

**What prevents it.** Nothing automatic. See the decision table in section 5.2.

---

# 15. Analysing only the survivors

**The mistake.** Dropping subjects with any missing occasion, and analysing the
complete cases.

**Why it is wrong.** If death is related to treatment, the survivors are a
selected subset, and the comparison is biased in favour of the treatment that
killed more. See section 8.4.

**What prevents it.** `agri_missing_report()` shows where the losses are
concentrated, and the incomplete engine uses the observed cells rather than
discarding subjects.

---

# 16. Not stating the missingness assumption

**The mistake.** Reporting an analysis of incomplete data without saying what was
assumed.

**Why it is wrong.** An assumption was made regardless. Leaving it unstated moves
it from the methods section, where a reader can disagree with it, into the code,
where they cannot.

**What prevents it.** `incomplete_wild_rank_test()` requires
`missing_assumption` explicitly. See section 9.1.

---

# 17. Treating a sensitivity analysis as a robustness ritual

**The mistake.** Running `agri_missing_sensitivity()`, observing that it ran, and
reporting the primary analysis unchanged.

**Why it is wrong.** The output is the point. If the conclusions differ across
assumptions, that difference is a result.

**What prevents it.** Reading section 10.2.

---

# 18. Using the experimental engine without its label

**The mistake.** Reporting the native wild-bootstrap engine as an established
method.

**Why it is wrong.** Its calibration study is incomplete. See section 12.

**What prevents it.** The label is in the function documentation, in
`VALIDATION_PLAN.md`, and here.

---

# Part V. Compact selection guide

# 19. Choose by the completeness and the question

| Situation | Use |
|---|---|
| complete data, trajectory question | `agri_repeated(backend = "nparLD")` |
| complete data, small samples | `agri_repeated(backend = "MANOVA.RM")` |
| complete data, endpoint question | analyse the endpoint with `np_rcbd()` |
| incomplete data | `agri_missing_report()`, then `incomplete_wild_rank_test()` |
| incomplete data, always | `agri_missing_sensitivity()` as well |
| repeated measures inside blocks | see section 11.2 |
| death related to treatment | analyse survival too, see *Time-to-Event and Ranking Data* |

---

# Part VI. Minimum reporting checklist

# 20. What the methods section must contain

1. the number of subjects, and the number of occasions per subject;
2. the subject and the within factor, named;
3. that the subject is the experimental unit for the between-subject factor;
4. the backend, and why it suits the completeness of the data;
5. the number of resampling iterations and the seed;
6. the amount of missingness and where it is concentrated;
7. the missingness assumption, and the mechanism believed to have operated;
8. the sensitivity analysis and its outcome;
9. if the native engine was used, its experimental status;
10. the package version.

# 21. A worked methods paragraph

> Plant height was measured on 12 plants per treatment at four occasions.
> The subject was the individual plant and the within-subject factor was
> occasion; the design was declared and validated with `agri_design()` from
> agriRank 0.14.0. Of the 144 scheduled measurements, 15% were absent,
> distributed across treatments without evident concentration
> (`agri_missing_report()`). Because the losses followed a rained-out
> measurement day affecting the whole trial, missingness was taken as completely
> at random. Inference used the native rank-based wild-bootstrap engine
> (`incomplete_wild_rank_test()`, ANOVA-type statistic, 1999 resamples, seed
> 801), which is labelled experimental in the package documentation. A
> sensitivity analysis across missingness assumptions
> (`agri_missing_sensitivity()`, 999 resamples) did not alter the conclusions and
> is reported in Table S3.

---

# 22. Where to go next

| If you now want | Read |
|---|---|
| the design declaration in detail | *Design Foundations, CRD, and RCBD* |
| comparisons after the omnibus test | *Effects, Conover, Contrasts, and Factorial Inference* |
| several responses on the same subjects | *Multivariate, Multi-Environment, Batch, and Sensitivity Workflows* |
| plant death as an outcome in its own right | *Time-to-Event and Ranking Data* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# Part VII. Glossary

# 23. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **subject** | the unit measured repeatedly; the replicate for a between-subject factor |
| **occasion** | one measurement time on a subject |
| **within factor** | a factor varying inside a subject |
| **between factor** | a factor varying across subjects |
| **trajectory** | the sequence of measurements on one subject |
| **MCAR** | missing completely at random, unrelated to anything |
| **MAR** | missing at random, related only to observed values |
| **MNAR** | missing not at random, related to the unobserved value itself |
| **complete case** | a subject with no missing occasion |
| **ATS** | ANOVA-type statistic, conservative in small samples |
| **WTS** | Wald-type statistic, liberal in small samples unless resampled |
| **wild bootstrap** | a resampling scheme that multiplies residuals by random signs |
| **sensitivity analysis** | repeating an analysis under alternative assumptions |

---

# Selected methodological references

- Amro, L., Konietschke, F., and Pauly, M. (2024). Multiplication-combination tests for incomplete paired data. *Statistics in Medicine*.
- Brunner, E., Bathke, A. C., and Konietschke, F. (2018). *Rank and Pseudo-Rank Procedures for Independent Observations in Factorial Designs*. Springer.
- Friedrich, S., Konietschke, F., and Pauly, M. (2019). Resampling-based analysis of multivariate data and repeated measures designs with the R package MANOVA.RM. *The R Journal*, 11(2), 380-400.
- Little, R. J. A., and Rubin, D. B. (2019). *Statistical Analysis with Missing Data*, 3rd edition. Wiley.
- Noguchi, K., Gel, Y. R., Brunner, E., and Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1-23. https://doi.org/10.18637/jss.v050.i12

The package also ships a verified RIS library under `inst/references/`.
