---
title: "Nonparametric and Shape-Aware Regression for Agronomic Gradients"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonparametric and Shape-Aware Regression for Agronomic Gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.8,
  fig.height = 4.4,
  dpi = 130
)
```

**Regression vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** fitting a curve to a quantitative agronomic gradient without assuming
its functional form.

---

# 1. Why this vignette exists

A fertilizer rate, a salinity level, an irrigation depth and a sowing density
are not categories. They lie on a scale, and the scientific content of the
experiment is the **shape** of the response along that scale.

The habitual analysis, comparing four or five rates as unrelated categories,
answers a question nobody asked and reports letters where a curve belongs. The
habitual alternative, fitting a quadratic because it is available, imposes a
shape the biology did not promise: a quadratic is symmetric about its peak and
falls away as fast as it rose, which almost no crop response does.

This vignette fits curves without imposing a functional form, and where a shape
**is** known from the biology, it imposes that shape rather than a convenient
polynomial.

> **Let the data choose the shape, unless the biology already fixed it. Never
> let a p-value choose the engine.**

## 1.1 Where this sits

| Stage | Vignette |
|---|---|
| declaring the design the curve is fitted inside | *Design Foundations* |
| the factorial alternative to a curve | *Effects, Conover, Contrasts, and Factorial Inference* |
| whole-number treatments | *Integer-Support Nonparametric Regression* |
| where the response stops changing, and intervals for a plot | *Distribution-Free Uncertainty and Model Checking* |
| a rate to recommend, with an interval on its location | *Optima, Quantiles, and How the Block Enters the Model* |

This vignette produces the curve. Those two turn it into a recommendation.

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. decide whether a quantitative treatment should be analysed as a factor, as a
   gradient, or as both;
2. name the fifteen engines available and say which family each belongs to;
3. choose an engine from the structure of the data and the biology, not from
   the fit;
4. impose a shape constraint when the biology fixes the shape, and explain what
   the constraint buys;
5. retain a declared block in a regression, and recognise the engines that
   cannot;
6. read the three explained-variation indices and say what each one answers;
7. compare engines predictively without letting the comparison feed back into
   inference;
8. produce a resampling band for an engine with no analytic interval;
9. fit and read a response surface over two gradients;
10. fit one curve per level of a qualitative factor;
11. test whether a prespecified parametric equation is too restrictive, without
    adopting it;
12. handle missing values explicitly rather than by silent deletion.

---

# 3. The regression module in one map

## 3.1 The engines, by family

```{r engines}
data.frame(
  engine = c("loess", "smoothing_spline", "kernel", "isotonic", "cobs",
             "discrete_kernel", "unimodal_isotonic",
             "umbrella", "integer_grid",
             "theil_sen", "siegel", "quantile", "smooth_quantile",
             "gam", "scam"),
  family = c(rep("strictly nonparametric", 7),
             rep("constrained / projected", 2),
             rep("rank-robust or semiparametric", 6))[1:15],
  shape_imposed = c("none", "none", "none", "monotone", "as declared",
                    "none", "one peak", "one peak", "none",
                    "none", "none", "none", "none", "none", "as declared")
)
```

## 3.2 The functions

| Function | Answers |
|---|---|
| `agri_np_regression()` | fit a curve |
| `agri_np_predict()` | predict, with an analytic interval where one is defined |
| `agri_np_plot()` | the standard figures |
| `agri_np_derivative()` | the slope of the fitted curve |
| `agri_np_optimum()` | the fitted extreme, as a bare point |
| `agri_np_bootstrap()` | a resampling band for the curve |
| `agri_np_compare()` | predictive comparison of engines |
| `agri_np_diagnostics()` | explained variation and predictive error |
| `agri_np_curves()` | several engines overlaid on the data |
| `agri_np_significance()` | bootstrap tests for kernel fits |
| `agri_np_specification()` | is a prespecified equation too restrictive |

---

# Part I. Factor or gradient?

# 4. The decision is about the treatment, not the fit

Nitrogen applied at 0, 40, 80, 120 and 160 kg/ha can be analysed two ways.

| As a factor | As a gradient |
|---|---|
| assumes nothing about the shape between levels | assumes the response varies smoothly between them |
| answers: do these five levels differ | answers: what shape, and where is it changing |
| cannot interpolate | can interpolate, inside the tested range |
| spends 4 degrees of freedom | spends fewer, and gains precision if the shape is real |
| output: letters | output: a curve |

Decide on scientific grounds. If the levels were chosen as points on a continuum
and interpolation between them is meaningful, the gradient analysis answers the
real question. If they are distinct managements that happen to carry numeric
labels, they are a factor.

## 4.1 Reporting both is not indecision

They answer different questions. A manuscript that reports only the factorial
analysis of a quantitative treatment has left the shape of the response
undescribed, which is usually the part a reader wants.

---

# Part II. The strictly nonparametric engines

# 5. A first fit

```{r ss-fit}
set.seed(2001)
dose_data <- data.frame(dose = seq(0, 220, length.out = 55))
dose_data$yield <- 5.2 + 0.075 * dose_data$dose -
  0.00022 * dose_data$dose^2 + rnorm(nrow(dose_data), 0, 0.65)

fit_ss <- agri_np_regression(
  yield ~ dose,
  data   = dose_data,
  method = "smoothing_spline"
)
fit_ss
```

## 5.1 The standard interrogation

```{r ss-interrogate}
agri_np_predict(fit_ss, data.frame(dose = c(50, 100, 150, 200)))
agri_np_diagnostics(fit_ss)
```

```{r ss-derivative}
head(agri_np_derivative(fit_ss))
agri_np_optimum(fit_ss)
```

## 5.2 The figures

```{r ss-plot, fig.cap = "Fitted curve over the observed points. The observed data are shown because a curve without them cannot be judged."}
agri_np_plot(fit_ss, type = "fit")
```

```{r ss-deriv-plot, fig.cap = "The slope of the fitted curve. Where it crosses zero is where the fitted response turns over."}
agri_np_plot(fit_ss, type = "derivative")
```

## 5.3 What a smoothing spline assumes

It assumes the response is **smooth**, and it chooses how smooth by
cross-validation. It does not assume a functional form, monotonicity, or a single
peak.

That freedom has a cost: with few distinct predictor values, or with noise
comparable to the signal, the fitted curve can follow features that are not
there. Section 9 gives the diagnostic that reveals it.

## 5.4 Interpretation

Report the fitted curve **with the observed points**, always. A smooth line
without data is not evidence; it is a drawing.

---

# 6. Shape constraints, when the biology fixes the shape

## 6.1 The situation

Some agronomic responses have a known direction. Biomass does not rise with
salinity. Germination does not fall with temperature over the sub-optimal range.
Where that knowledge exists, imposing it is not an assumption added to the data:
it is information the data did not have to spend degrees of freedom rediscovering.

```{r iso-data}
set.seed(2002)
sal <- data.frame(ec = seq(0.5, 6.0, length.out = 60))
sal$biomass <- 31 * exp(-0.12 * sal$ec^1.45) + rnorm(nrow(sal), 0, 1.1)

fit_iso <- agri_np_regression(
  biomass ~ ec,
  data   = sal,
  method = "isotonic",
  shape  = "decreasing"
)
fit_iso
```

```{r iso-plot, fig.cap = "Isotonic regression with a declared decreasing constraint. The fit is a step function, which is what the method estimates."}
agri_np_plot(fit_iso)
```

## 6.2 Isotonic regression fits steps, and that is honest

The fitted object is a step function, not a smooth curve. That is not a
limitation to be hidden: isotonic regression estimates the best **monotone**
function, and the best monotone fit to finite noisy data is a step function.
Drawing a smooth line through it would suggest a precision the estimate does not
have.

## 6.3 A smooth constrained alternative

```{r scam-fit}
if (requireNamespace("scam", quietly = TRUE)) {
  fit_scam <- agri_np_regression(
    biomass ~ ec,
    data   = sal,
    method = "scam",
    shape  = "decreasing"
  )
  print(agri_np_plot(fit_scam, interval = TRUE))
}
```

SCAM imposes the same monotone constraint but through a penalized spline basis,
so the result is smooth and carries an analytic interval. Where a smooth
constrained curve is wanted for presentation, this is the route.

## 6.4 The available constraints

```{r shapes}
data.frame(
  shape = c("increasing", "decreasing", "convex", "concave",
            "increasing_convex", "increasing_concave",
            "decreasing_convex", "decreasing_concave"),
  agronomic_example = c(
    "germination over sub-optimal temperature",
    "biomass over salinity",
    "cost over input rate",
    "yield over a nutrient below the optimum",
    "accelerating growth phase",
    "diminishing returns to fertilizer",
    "decelerating decline",
    "accelerating decline under stress")
)
```

## 6.5 The constraint must come from biology, not from the plot

```{r shape-refused, error = TRUE}
# Isotonic regression requires an explicit direction. It will not infer one.
agri_np_regression(biomass ~ ec, sal, method = "isotonic")
```

Choosing the constraint after looking at the scatter is fitting the data twice:
once to pick the shape and once to estimate it. The stated uncertainty then does
not cover the first step.

## 6.6 Interpretation

Report the constraint, and the biological reason for it, in the methods. A
constrained fit whose constraint is not justified is a stronger claim than an
unconstrained one, not a weaker one.

---

# Part III. Keeping the design

# 7. A block in a regression

```{r gam-block}
set.seed(2003)
rcbd <- expand.grid(
  block = factor(1:5),
  N     = seq(0, 200, length.out = 12)
)
rcbd$yield <- 6 + 0.075 * rcbd$N - 0.00021 * rcbd$N^2 +
  as.numeric(rcbd$block) * 0.35 + rnorm(nrow(rcbd), 0, 0.55)

if (requireNamespace("mgcv", quietly = TRUE)) {
  fit_gam <- agri_np_regression(
    yield ~ N,
    data   = rcbd,
    method = "gam",
    block  = block
  )
  print(summary(fit_gam))
}
```

```{r gam-plot, fig.cap = "Block-adjusted fitted response with its analytic interval. The band covers the curve, not a plot."}
if (exists("fit_gam")) print(agri_np_plot(fit_gam, predictor = "N", interval = TRUE))
```

## 7.1 Engines that cannot carry a block are refused

```{r block-refused, error = TRUE}
agri_np_regression(yield ~ N, rcbd, method = "loess", block = block)
```

The refusal is the same principle as in the design vignette. LOESS fits a curve
in one predictor and has no mechanism for an adjustment factor. Applying it to
blocked data would leave the between-block variation in the residual, which is
what blocking removed.

## 7.2 Which engines keep a block

```{r block-capable}
data.frame(
  keeps_block = c("gam", "scam", "kernel", "quantile", "smooth_quantile",
                  "umbrella", "integer_grid"),
  refuses_block = c("loess", "smoothing_spline", "theil_sen", "siegel",
                    "cobs", "isotonic", "unimodal_isotonic")
)
```

## 7.3 Interpretation

If the experiment was blocked, the regression must keep the block. If the
preferred engine cannot, the choice is between a different engine and a
justified decision to ignore the blocking, and the second requires an argument
in the methods.

---

# 8. Quantile regression, and the other engines

## 8.1 Conditional quantiles

```{r quantile-fit}
if (requireNamespace("quantreg", quietly = TRUE)) {
  fit_q50 <- agri_np_regression(yield ~ N, rcbd, method = "quantile",
                                tau = 0.50, block = block)
  fit_q10 <- agri_np_regression(yield ~ N, rcbd, method = "quantile",
                                tau = 0.10, block = block)
  nd <- rcbd[1:5, ]
  print(data.frame(N = nd$N,
                   q50 = round(as.numeric(agri_np_predict(fit_q50, nd)), 2),
                   q10 = round(as.numeric(agri_np_predict(fit_q10, nd)), 2)))
}
```

The median describes the typical plot; the tenth percentile describes the plot a
grower meets in a bad year. A treatment that raises one without the other changes
the risk, not only the level, and the two are different recommendations. The
smooth version and the full fan of quantiles belong to *Optima, Quantiles, and
How the Block Enters the Model*.

## 8.2 Kernel regression with mixed predictor types

```{r kernel-fit}
if (requireNamespace("np", quietly = TRUE)) {
  set.seed(2004)
  kd <- expand.grid(
    cultivar = factor(c("A", "B")),
    dose     = seq(0, 180, length.out = 20)
  )
  kd$yield <- 5 + 0.06 * kd$dose - 0.00018 * kd$dose^2 +
    ifelse(kd$cultivar == "B", 1.2, 0) + rnorm(nrow(kd), 0, .45)
  fit_kernel <- agri_np_regression(yield ~ dose + cultivar, kd,
                                   method = "kernel")
  print(agri_np_diagnostics(fit_kernel))
}
```

Kernel regression handles continuous and categorical predictors in one fit, with
bandwidths chosen by cross-validation. It is the most flexible engine here and
the most data-hungry.

## 8.3 Rank-robust slopes

Theil-Sen and Siegel regression estimate a slope as a median of pairwise slopes.
They are the natural choice when the relationship is close to linear and a few
extreme plots would dominate an ordinary fit. They also return interpretable
coefficients, which the smoothers deliberately do not.

```{r theilsen}
ts_fit <- agri_np_regression(yield ~ dose, dose_data, method = "theil_sen")
coef(ts_fit)
confint(ts_fit)
```

## 8.4 Why the smoothers refuse coefficients

```{r coef-refused, error = TRUE}
coef(fit_ss)
```

A smoothing spline has coefficients, but they are basis weights with no
agronomic meaning. Reporting one as "the effect of nitrogen" would invite a
reading the model does not support, so the package refuses by name.

---

# Part IV. Choosing an engine honestly

# 9. Explained variation, and what each index answers

```{r r2}
agri_np_diagnostics(fit_ss, cv = TRUE, seed = 1)$r2
```

| Index | Question | Fails to notice |
|---|---|---|
| `pseudo_r2` | how much variation does the fitted curve reproduce | overfitting: it always rises with flexibility |
| `cv_r2` | how much would it reproduce on a plot it has not seen | nothing, but it is noisy in small experiments |
| `spearman_r2` | how much of the **ordering** does it get right | the size of the departures |
| `effective_df` | how much flexibility was actually used | nothing; read it beside the other three |

## 9.1 The pattern that reveals overfitting

A flexible engine typically shows a **larger** `pseudo_r2`, a **larger**
`effective_df` and a **smaller** `cv_r2` than a rigid one on the same data.

```{r overfit-demo}
cmp <- do.call(rbind, lapply(c("smoothing_spline", "loess", "theil_sen"),
  function(m) {
    f <- agri_np_regression(yield ~ dose, dose_data, method = m)
    r <- agri_np_diagnostics(f, cv = TRUE, seed = 1)$r2
    data.frame(engine = m, pseudo_r2 = round(r$pseudo_r2, 3),
               cv_r2 = round(r$cv_r2, 3), edf = round(r$effective_df, 2))
  }))
cmp
```

Read the gap between `pseudo_r2` and `cv_r2`. A large gap is the signature of a
curve following noise, and it is visible only because the two are reported
together.

## 9.2 Why a single R-squared beside a flexible fit is uninformative

A sufficiently flexible smoother can reproduce any data set exactly, giving
`pseudo_r2 = 1` and no predictive ability at all. Reporting that number alone
tells the reader nothing about the response and everything about the flexibility
that was permitted.

## 9.3 Interpretation

Report `pseudo_r2` and `cv_r2` together, always, with `effective_df` beside
them.

---

# 10. Predictive comparison of engines

```{r compare}
agri_np_compare(
  yield ~ dose,
  dose_data,
  methods = c("smoothing_spline", "loess"),
  kfold   = 5,
  metric  = "RMSE"
)
```

```{r compare-block}
if (requireNamespace("mgcv", quietly = TRUE)) {
  print(agri_np_compare(yield ~ N, rcbd, methods = c("gam", "kernel"),
                        block = block, kfold = 4))
}
```

## 10.1 What this table is for, and what it is not

It reports **cross-validated predictive error**. It says which engine predicts a
held-out plot best.

It does **not** say which engine gives the correct p-value, and choosing the
engine that produced the most convenient p-value is the single most common way
to invalidate a nonparametric analysis. The package separates the two operations
deliberately: nothing in `agri_np_compare()` feeds back into inference.

## 10.2 The failures column

An engine that cannot carry the declared block is recorded as a failure rather
than dropped from the table, so the comparison cannot mislead by omission.

## 10.3 Several engines on one figure

```{r curves, fig.cap = "Three engines over the same data. Where they agree, the shape is not an artefact of the smoother."}
agri_np_curves(yield ~ dose, dose_data,
               methods = c("smoothing_spline", "loess", "theil_sen"))
```

Overlaying engines is a useful sanity check. A feature that appears under one
smoother and not the others is a property of that smoother.

## 10.4 Interpretation

Report the comparison table in supplementary material, and state explicitly that
the inferential engine was not selected from it.

---

# 11. Resampling bands

```{r bootstrap}
if (requireNamespace("mgcv", quietly = TRUE)) {
  bb <- agri_np_bootstrap(fit_gam, B = 99, n = 80, seed = 2005)
  print(head(as.data.frame(bb)))
}
```

## 11.1 When you need one

Several engines have no analytic interval: isotonic, LOESS, Theil-Sen and the
kernel fits among them. `agri_np_bootstrap()` supplies a resampling band for any
of them.

## 11.2 What the band covers

It covers the **fitted curve**, not a future plot. That distinction is developed
in *Distribution-Free Uncertainty and Model Checking*, and it matters: a plot
varies far more than the average response, typically by a factor of several.

## 11.3 Pointwise or simultaneous

```{r bootstrap-band}
if (exists("fit_gam")) {
  pw <- as.data.frame(agri_np_bootstrap(fit_gam, B = 99, n = 40, seed = 1,
                                        band = "pointwise"))
  st <- as.data.frame(agri_np_bootstrap(fit_gam, B = 99, n = 40, seed = 1,
                                        band = "simultaneous"))
  print(data.frame(band = c("pointwise", "simultaneous"),
                   mean_width = round(c(mean(pw$upper - pw$lower),
                                        mean(st$upper - st$lower)), 3)))
}
```

A pointwise band covers the curve at each single point with the stated
probability. A simultaneous band covers the **whole curve at once**, and is
therefore wider. If the claim concerns the shape of the curve rather than its
value at one rate, the simultaneous band is the correct one.

## 11.4 The resampling unit

The bootstrap resamples the declared block by default, keeping whole blocks
together, because plots within a block are not exchangeable with plots in other
blocks. That is what declaring a block asserts.

## 11.5 Interpretation

State `B`, the seed, whether the band is pointwise or simultaneous, and what it
covers. `B = 99` is a speed device for a vignette; use at least 999 for anything
reported.

---

# Part V. More than one gradient, more than one curve

# 12. Response surfaces

```{r surface}
if (requireNamespace("mgcv", quietly = TRUE)) {
  set.seed(2006)
  surf <- expand.grid(
    N     = seq(0, 200, length.out = 15),
    water = seq(0.5, 1.0, length.out = 10)
  )
  surf$yield <- 5 + 0.065 * surf$N - 0.00018 * surf$N^2 +
    5 * surf$water + 0.015 * surf$N * surf$water + rnorm(nrow(surf), 0, .55)

  fit_surface <- agri_np_regression(yield ~ N + water, surf, method = "gam",
                                    gam_structure = "tensor")
  print(agri_np_plot(fit_surface, type = "surface",
                     surface_predictors = c("N", "water"), n = 50))
}
```

## 12.1 Additive or tensor?

| `gam_structure` | Fits | Assumes |
|---|---|---|
| `"additive"` | one smooth per predictor, added | the effect of N is the same at every water level |
| `"tensor"` | a joint smooth over both | nothing; the effects may interact |
| `"varying"` | one smooth per level of a factor | the shape may differ between levels |

The data above were generated **with** an N-by-water interaction, so the additive
structure would misrepresent them. Choose `"tensor"` when the two gradients are
expected to interact, which for water and nitrogen they nearly always do.

## 12.2 The cost

A tensor smooth spends many more degrees of freedom than an additive one. With
150 plots it is affordable; with 30 it is not, and the additive structure is the
honest simplification.

## 12.3 Interactive inspection

```{r interactive, eval = FALSE}
# Not run in the vignette: it opens a browser widget.
if (requireNamespace("plotly", quietly = TRUE)) {
  agri_np_interactive(fit_surface, type = "surface",
                      surface_predictors = c("N", "water"), n = 40)
}
```

---

# 13. One curve per cultivar

```{r grouped}
if (requireNamespace("mgcv", quietly = TRUE)) {
  set.seed(2007)
  gd <- expand.grid(cultivar = factor(c("A", "B", "C")),
                    dose = seq(0, 180, length.out = 18))
  gd$yield <- 5 + .06 * gd$dose - .00017 * gd$dose^2 +
    c(A = 0, B = 1.2, C = -0.6)[gd$cultivar] + rnorm(nrow(gd), 0, .5)
  gfit <- agri_np_regression(yield ~ dose + cultivar, gd, method = "gam")
  print(agri_np_plot(gfit, predictor = "dose", group = "cultivar"))
}
```

## 13.1 Additive means parallel

The model above adjusts for cultivar **additively**, so the three curves are
parallel by construction: same shape, different heights.

That is a modelling choice, not a finding. If the question is whether the
cultivars respond differently, the model must allow the shapes to differ:

```{r varying-fit}
if (requireNamespace("mgcv", quietly = TRUE)) {
  gvar <- agri_np_regression(yield ~ dose + cultivar, gd, method = "gam",
                             gam_structure = "varying")
  print(gvar$formula_used)
}
```

```{r varying-plot, fig.cap = "One smooth of dose per cultivar. The curves may now differ in shape, not only in height."}
if (exists("gvar")) print(agri_np_plot(gvar, predictor = "dose", group = "cultivar"))
```

This distinction has a consequence developed fully in *Optima, Quantiles, and How
the Block Enters the Model*: parallel curves share one optimum by construction,
so comparing optima across additively adjusted levels reports a difference of
exactly zero.

## 13.2 Levels and coefficients

```{r levels}
if (exists("gfit")) print(agri_np_levels(gfit, B = 99, seed = 1))
```

`agri_np_levels()` reports the response at each level of the qualitative
predictor with a bootstrap interval, which is the level-oriented companion to a
coefficient forest plot: coefficients state contrasts against a reference,
this states what the model predicts at each level itself.

---

# Part VI. Testing, and what the tests are for

# 14. Bootstrap significance for kernel fits

```{r significance}
if (exists("fit_kernel") && requireNamespace("np", quietly = TRUE)) {
  print(agri_np_significance(fit_kernel, variables = "cultivar",
                             B = 99, boot_type = "I"))
}
```

This tests whether a predictor contributes to a kernel fit, by bootstrap. It
answers an **inferential** question, which cross-validated predictive error does
not.

# 15. Is a prespecified equation too restrictive?

```{r specification}
if (requireNamespace("np", quietly = TRUE)) {
  candidate_linear <- lm(yield ~ dose, data = dose_data, x = TRUE, y = TRUE)
  candidate_quadratic <- lm(yield ~ dose + I(dose^2), data = dose_data,
                            x = TRUE, y = TRUE)

  print(agri_np_specification(candidate_linear, B = 99))
  print(agri_np_specification(candidate_quadratic, B = 99))
}
```

## 15.1 What this does, and what it does not license

It tests a **prespecified** parametric equation against a nonparametric
alternative. A rejection says the parametric form is too restrictive for these
data.

It does not license adopting the parametric form when it is not rejected.
Failure to reject is not evidence of adequacy, particularly with modest sample
sizes, and the package's position throughout is that a functional form should be
imposed only when the biology supplies it.

## 15.2 The legitimate use

Reporting that a linear response was rejected in favour of a nonparametric
alternative is a useful, checkable statement about the data. It is the honest
version of "we fitted a curve because a straight line did not describe the
response".

---

# Part VII. Missing values

# 16. Silence is the enemy

```{r missing-refused, error = TRUE}
miss <- dose_data
miss$yield[c(4, 19)] <- NA

agri_np_regression(yield ~ dose, miss, method = "smoothing_spline")
```

```{r missing-complete}
fit_complete <- agri_np_regression(
  yield ~ dose,
  miss,
  method    = "smoothing_spline",
  na_action = "complete"
)
c(original = fit_complete$n_original, omitted = fit_complete$n_omitted)
```

## 16.1 Why the default refuses

Silent row deletion is among the most common sources of irreproducible results,
and it is invisible in the printed output. A curve fitted to 53 of 55 plots looks
exactly like a curve fitted to 55.

The default `na_action = "fail"` stops. Requesting `"complete"` is an explicit
decision, and the object records how many rows were dropped so a reader can check.

## 16.2 Complete-case analysis is not imputation

It is not a missing-data model. It assumes the omitted rows are missing
completely at random, and where that is doubtful the missingness itself deserves
analysis. See *Repeated Measures and Missing Longitudinal Data*.

---

# Part VIII. Reporting

# 17. Tables and export

```{r report}
fit_report <- agri_np_regression(yield ~ dose, dose_data,
                                 method = "smoothing_spline")
agri_table(fit_report, "metrics", format = "data.frame")
```

```{r report-derivative}
head(agri_table(fit_report, "derivative", n = 40, format = "data.frame"))
agri_table(fit_report, "optimum", format = "data.frame")
```

```{r report-files, eval = FALSE}
agri_report(fit_report, "regression_analysis.md", format = "md")
export_results(fit_report, "regression_analysis.rds")
```

---

# Part IX. Common mistakes, and the function that prevents each

# 18. Analysing a gradient as unrelated categories

**The mistake.** Five nitrogen rates reported as five letters.

**Why it is wrong.** It leaves the shape of the response undescribed, which is
the agronomic content.

**What prevents it.** `agri_np_regression()`. Report both analyses; see section
4.

---

# 19. Fitting a quadratic because it is available

**The mistake.** `lm(y ~ x + I(x^2))` as the default dose-response model.

**Why it is wrong.** A quadratic is symmetric about its peak and falls as fast as
it rose. Almost no crop response does that, and the imposed symmetry moves the
estimated optimum.

**What prevents it.** The nonparametric engines, and `agri_np_specification()`
for the honest test of whether the parametric form is too restrictive. See
section 15.

---

# 20. Choosing the engine by the p-value it produces

**The mistake.** Fitting several smoothers and reporting the one whose treatment
effect reached significance.

**Why it is wrong.** The reported p-value is then the minimum of several and its
null distribution is not the one being quoted.

**What prevents it.** `agri_np_compare()` reports predictive error only, and
nothing in it feeds back into inference. See section 10.1.

---

# 21. Choosing a shape constraint from the scatter plot

**The mistake.** Looking at the data, seeing a decline, and imposing
`shape = "decreasing"`.

**Why it is wrong.** The data are used twice, and the stated uncertainty does not
cover the first use.

**What prevents it.** Nothing automatic. The constraint must come from biology.
See section 6.5.

---

# 22. Reporting a single R-squared beside a flexible fit

**The mistake.** "The smoothing spline explained 94% of the variation."

**Why it is wrong.** A sufficiently flexible smoother explains everything and
predicts nothing. See section 9.2.

**What prevents it.** `agri_np_diagnostics(cv = TRUE)` returns three indices and
the effective degrees of freedom together.

---

# 23. Quoting a bootstrap band as a prediction interval

**The mistake.** "Yields will fall in this band."

**Why it is wrong.** The band covers the fitted curve, not a plot. A plot varies
far more.

**What prevents it.** `agri_np_conformal()` in the uncertainty vignette. See
section 11.2.

---

# 24. Comparing optima across additively adjusted levels

**The mistake.** Fitting `y ~ dose + cultivar` and comparing the optimum per
cultivar.

**Why it is wrong.** The curves are parallel by construction, so the optima are
identical.

**What prevents it.** `gam_structure = "varying"`, and
`agri_np_optimum_test()` refuses the comparison outright. See section 13.1.

---

# 25. Letting rows disappear

**The mistake.** A missing value silently removing plots from the fit.

**Why it is wrong.** The reported n does not match the analysed n, and nobody can
tell.

**What prevents it.** `na_action = "fail"` is the default, and `"complete"`
records the count. See section 16.

---

# Part X. Compact selection guide

# 26. Choose the engine by the structure and the biology

| Your situation | Engine |
|---|---|
| smooth response, no shape known, no block | `smoothing_spline` or `loess` |
| smooth response, blocked design | `gam` |
| a known monotone direction | `isotonic` or `scam` |
| a known single peak | `unimodal_isotonic` or `umbrella` |
| mixed continuous and categorical predictors | `kernel` |
| a near-linear response with extreme plots | `theil_sen` or `siegel` |
| a question about the poor plots | `quantile` or `smooth_quantile` |
| two interacting gradients | `gam` with `gam_structure = "tensor"` |
| curves that may differ in shape between levels | `gam` with `gam_structure = "varying"` |
| whole-number treatments | `integer_grid`, see the integer vignette |

# 27. Choose the diagnostic by the question

| Question | Use |
|---|---|
| how well does it fit | `agri_np_diagnostics()`, all three indices |
| which engine predicts best | `agri_np_compare()` |
| is the shape an artefact of the smoother | `agri_np_curves()` |
| how uncertain is the curve | `agri_np_bootstrap()` |
| does a predictor contribute | `agri_np_significance()` |
| is a parametric form too restrictive | `agri_np_specification()` |

---

# Part XI. Minimum reporting checklist

# 28. What the methods section must contain

1. that the treatment was analysed as a gradient, and why;
2. the engine, named, and the reason it was admissible for this design;
3. any shape constraint, and the biological justification for it;
4. whether the declared block was retained;
5. that the engine was not selected on the basis of a response p-value;
6. `pseudo_r2` and `cv_r2` together, with the effective degrees of freedom;
7. for a resampling band: `B`, the seed, and whether it is pointwise or
   simultaneous;
8. the number of rows analysed against the number supplied;
9. the package version.

# 29. A worked methods paragraph

> The nitrogen response was modelled without assuming a functional form, using a
> penalized regression spline with the block retained as an adjustment term
> (`agri_np_regression(method = "gam")` from agriRank 0.14.0). The engine was
> chosen because the design was blocked and no shape was fixed by the biology;
> it was not selected on the basis of any response p-value. Predictive comparison
> across admissible engines is reported in Table S1 (`agri_np_compare()`,
> 5-fold cross-validation, seed 1). Explained variation was 0.87 on the fitted
> values and 0.81 out of fold, with 4.2 effective degrees of freedom. Uncertainty
> in the fitted curve is shown as a simultaneous resampling band
> (`agri_np_bootstrap()`, 999 replicates, whole blocks resampled, seed 2005); it
> covers the curve and not an individual plot.

---

# 30. Where to go next

| If you now want | Read |
|---|---|
| where the response stops changing | *Distribution-Free Uncertainty and Model Checking* |
| an interval covering the next plot | the same |
| a rate to recommend, with an interval on its location | *Optima, Quantiles, and How the Block Enters the Model* |
| whole-number treatments | *Integer-Support Nonparametric Regression* |
| the design the curve is fitted inside | *Design Foundations, CRD, and RCBD* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# Part XII. Glossary

# 31. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **gradient** | a quantitative treatment whose levels lie on a meaningful scale |
| **smoother** | an estimator that fits a curve without a functional form |
| **shape constraint** | a restriction such as monotone or single-peaked, imposed from biology |
| **effective degrees of freedom** | how much flexibility a penalized fit actually used |
| **pseudo R-squared** | variation reproduced by the fitted values |
| **cross-validated R-squared** | variation reproduced on held-out observations |
| **pointwise band** | covers the curve at each single point |
| **simultaneous band** | covers the whole curve at once, and is wider |
| **tensor smooth** | a joint smooth over two predictors, allowing interaction |
| **varying smooth** | one smooth per level of a qualitative predictor |
| **specification test** | a test of whether a parametric form is too restrictive |
| **complete-case analysis** | analysing only rows with no missing value |

---

# Selected methodological references

- Hastie, T., and Tibshirani, R. (1990). *Generalized Additive Models*. Chapman and Hall.
- Hayfield, T., and Racine, J. S. (2008). Nonparametric econometrics: the np package. *Journal of Statistical Software*, 27(5). https://doi.org/10.18637/jss.v027.i05
- Koenker, R. (2005). *Quantile Regression*. Cambridge University Press.
- Pya, N., and Wood, S. N. (2015). Shape constrained additive models. *Statistics and Computing*, 25, 543-559. https://doi.org/10.1007/s11222-013-9448-7
- Racine, J., and Li, Q. (2004). Nonparametric estimation of regression functions with both categorical and continuous data. *Journal of Econometrics*, 119, 99-130. https://doi.org/10.1016/S0304-4076(03)00157-X
- Sen, P. K. (1968). Estimates of the regression coefficient based on Kendall's tau. *Journal of the American Statistical Association*, 63, 1379-1389.
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R*, 2nd edition. Chapman and Hall/CRC.

The package also ships a verified RIS library under `inst/references/`.
