---
title: "Integer-Support Nonparametric Regression for Agronomy"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integer-Support Nonparametric Regression for Agronomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.8,
  fig.height = 4.4,
  dpi = 130
)
```

**Integer decision vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** treatments whose admissible values are whole numbers, and the decision
support that follows from that fact.

---

# 1. Why this vignette exists

A grower cannot plant 7.4 plants per hill, apply 2.6 sprays, install 3.7
irrigation events, or set 11.3 traps per hectare. The admissible values are
whole numbers, and there is nothing between them.

The habitual analysis fits a continuous curve, locates its maximum at 7.4, and
recommends 7. That is wrong at the level of the **estimand**, not the
presentation: the rounded value is the neighbour of an inadmissible one, and it
need not be the best admissible decision.

```{r rounding-trap}
# A response evaluated only where a decision can actually be taken.
x <- 1:12
fx <- c(24, 31, 37, 41.6, 44.8, 46.6, 47.0, 46.0, 43.6, 39.8, 34.6, 28)
data.frame(plants = x, fitted = fx)[6:8, ]
```

The continuous maximum of a curve through those points sits near 6.8. Rounding
gives 7, and here 7 happens to be right. But nothing guaranteed it: with a
slightly asymmetric response the continuous maximum can sit at 6.8 while the best
admissible value is 6, and rounding then recommends the wrong density.

> **Evaluate the response on the admissible support. Do not optimise a
> continuous curve and round afterwards.**

## 1.1 What follows from taking this seriously

Once the support is discrete, three familiar objects change:

| Object | Continuous | Integer support |
|---|---|---|
| the optimum | a point on a line | the best of a finite set |
| its uncertainty | an interval | a **probability mass** over decisions |
| the rate of change | a derivative | a finite difference |

The third is the one most often overlooked. A derivative is not an admissible
quantity on a discrete support, and the package refuses to compute one.

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. recognise an agronomic treatment whose support is integer;
2. explain why rounding a continuous optimum is an error of estimand;
3. declare an integer support, and choose among the three declaration modes;
4. choose among the four integer engines on structural grounds;
5. read a first and second finite difference agronomically;
6. compute response per unit of input and say when it is the right criterion;
7. apply the three threshold rules and match each to a decision context;
8. read a bootstrap probability mass over decisions, rather than an interval;
9. build and interpret a discrete confidence set;
10. explain why the package refuses fractional predictions and derivatives;
11. distinguish interpolation from extrapolation on a sparse integer support.

---

# 3. The integer module in one map

| Function | Answers |
|---|---|
| `agri_np_regression(predictor_support =)` | fit, restricted to admissible values |
| `agri_integer_predict()` | the response at every admissible decision |
| `agri_integer_optimum()` | the best admissible decision |
| `agri_integer_difference()` | what one more unit gains |
| `agri_integer_efficiency()` | response per unit of input |
| `agri_integer_threshold()` | the smallest decision meeting a stated rule |
| `agri_integer_bootstrap()` | probability mass over decisions |
| `agri_integer_confset()` | a discrete confidence set |

## 3.1 The four engines

```{r engines}
data.frame(
  engine = c("discrete_kernel", "unimodal_isotonic", "umbrella", "integer_grid"),
  fits = c("ordered-discrete kernel regression",
           "a single increase-then-decrease response",
           "constrained umbrella order with covariate adjustment",
           "a flexible latent curve projected onto the lattice"),
  keeps_block = c("no", "no", "yes", "via the base engine"),
  use_when = c("no shape assumption, moderate replication",
               "one peak is known, no block",
               "one peak is known, blocked design",
               "no shape assumption, and a familiar smoother is wanted")
)
```

## 3.2 The three declaration modes

```{r support-modes}
data.frame(
  predictor_support = c("observed_integer", "integer_range", "custom_integer"),
  admits = c("only the integers actually tested",
             "every integer between declared bounds",
             "an explicit set supplied by the analyst"),
  interpolates = c("no", "yes", "as declared"),
  use_when = c("untested densities are not recommendable",
               "the gap between tested values is agronomically meaningless",
               "only certain values are operationally possible")
)
```

The distinction between the first two is scientific. If the trial tested 1, 3,
5, 7 and 9 plants, is 4 a recommendation the experiment supports? Sometimes yes,
because nothing physical happens between 3 and 5. Sometimes no, because the
seeder cannot be set to 4. The declaration records which.

---

# Part I. The four engines

# 4. Ordered-discrete kernel regression

```{r discrete-kernel}
if (requireNamespace("np", quietly = TRUE)) {
  set.seed(4101)

  insects <- data.frame(insects = rep(0:10, each = 7))
  insects$damage <- 3.5 + 1.25 * insects$insects +
    0.08 * insects$insects^2 +
    rnorm(nrow(insects), 0, 1.4)

  fit_dk <- agri_np_regression(
    damage ~ insects,
    data              = insects,
    method            = "discrete_kernel",
    predictor_support = "observed_integer",
    integer_kernel    = "wangvanryzin"
  )
  print(fit_dk)
  print(agri_integer_predict(fit_dk))
}
```

```{r dk-plot, fig.cap = "An integer-support fit drawn as steps and crosses, not a continuous line. There is nothing between two admissible decisions, and the figure says so."}
if (exists("fit_dk")) print(agri_np_plot(fit_dk))
```

## 4.1 Why the figure is drawn this way

A continuous line through integer values suggests that a value exists between
them. Where the support is genuinely discrete, that suggestion is false, and it
is the visual form of the rounding error this vignette exists to prevent.

## 4.2 Fractional predictions are refused

```{r discrete-reject, error=TRUE}
if (exists("fit_dk")) {
  agri_np_predict(fit_dk, data.frame(insects = 4.5))
}
```

Once integer support has been declared, a prediction at 4.5 insects describes an
observation that cannot occur. The refusal is not pedantry: a fitted value there
would flow into an optimum, a threshold and eventually a recommendation.

## 4.3 The kernel choice

| `integer_kernel` | Suits |
|---|---|
| `"wangvanryzin"` | ordered discrete predictors, where 3 is between 2 and 4 |
| `"liracine"` | unordered discrete predictors, where the levels have no order |

For counts of plants, insects or sprays the ordering is real, so
`"wangvanryzin"` is the default choice.

---

# 5. Unimodal isotonic regression

```{r unimodal}
if (requireNamespace("Iso", quietly = TRUE)) {
  set.seed(4102)

  density <- data.frame(plants = rep(1:10, each = 6))
  density$yield <- 28 +
    8.5 * pmin(density$plants, 6) -
    5.2 * pmax(density$plants - 6, 0) +
    rnorm(nrow(density), 0, 2)

  fit_ui <- agri_np_regression(
    yield ~ plants,
    data              = density,
    method            = "unimodal_isotonic",
    predictor_support = "observed_integer"
  )
  print(fit_ui)
  print(agri_integer_predict(fit_ui))
  print(agri_integer_optimum(fit_ui))
}
```

## 5.1 What the constraint buys

A plant-density response rises to a peak and then falls: too few plants waste
land, too many compete. That shape is known from the agronomy, not discovered
from the data.

Imposing it means the estimate cannot show a spurious second peak produced by
noise, and it means the optimum is well defined by construction. The information
comes from the biology, and the data are spent on locating the peak rather than
on establishing that there is one.

## 5.2 When not to impose it

If the response might be monotone over the tested range, because the trial did
not reach the density at which competition begins, a unimodal constraint forces a
peak that may lie outside the data. The fitted optimum is then at the boundary
and means nothing.

Check with an unconstrained engine before imposing the constraint.

---

# 6. Umbrella regression with a block

```{r umbrella}
if (requireNamespace("cgam", quietly = TRUE)) {
  set.seed(4103)

  rcbd <- expand.grid(block = factor(1:5), plants = 1:9)
  rcbd$yield <- 32 +
    9 * pmin(rcbd$plants, 6) -
    5.5 * pmax(rcbd$plants - 6, 0) +
    as.numeric(rcbd$block) * 0.7 +
    rnorm(nrow(rcbd), 0, 1.8)

  fit_umb <- agri_np_regression(
    yield ~ plants,
    data              = rcbd,
    method            = "umbrella",
    block             = block,
    predictor_support = "observed_integer"
  )
  print(fit_umb)
  print(agri_integer_optimum(fit_umb))
}
```

```{r umbrella-plot, fig.cap = "Block-adjusted umbrella fit on an integer support."}
if (exists("fit_umb")) print(agri_np_plot(fit_umb))
```

Umbrella regression is the engine to reach for when the response has one peak
**and** the design was blocked. It is the only one of the four that carries a
block directly.

---

# 7. Integer grid projection

```{r integer-grid}
set.seed(4104)

ig <- data.frame(plants = rep(1:12, each = 6))
ig$yield <- 24 + 7.5 * ig$plants -
  0.48 * ig$plants^2 +
  rnorm(nrow(ig), 0, 1.4)

fit_ig <- agri_np_regression(
  yield ~ plants,
  data                = ig,
  method              = "integer_grid",
  integer_base_method = "smoothing_spline",
  predictor_support   = "integer_range",
  integer_range       = c(1, 12)
)

agri_integer_predict(fit_ig)
```

```{r ig-optimum}
agri_integer_optimum(fit_ig)
```

```{r ig-plot, fig.cap = "A flexible latent curve, evaluated only at admissible densities."}
agri_np_plot(fit_ig)
```

## 7.1 What projection means

A familiar smoother is fitted to the data, and every **public** prediction,
optimum and decision is then evaluated on the admissible lattice. The latent
curve is a computational device; nothing is ever reported off the lattice.

That is a different operation from fitting a curve and rounding the answer. The
optimum is the best value **among the admissible ones**, found by evaluation, not
the nearest integer to a continuous argmax.

## 7.2 Choosing the base engine

| `integer_base_method` | When |
|---|---|
| `"smoothing_spline"` | no block, smooth response |
| `"gam"` | blocked design |
| `"scam"` | blocked design and a known shape |
| `"quantile"` | the question concerns the poor plots |

---

# Part II. Decision support on a discrete support

# 8. Finite differences replace the derivative

```{r differences}
agri_integer_difference(fit_ig, order = 1)
```

## 8.1 The first difference is the agronomic quantity

On a discrete support the derivative is not admissible. What replaces it is the
difference between adjacent decisions: **what is gained by planting one more**.

That is also the quantity an economic threshold applies to. If one more plant per
hill gains 1.2 units of yield and costs 0.9 units of seed and handling, the
decision is arithmetic.

```{r differences-2}
agri_integer_difference(fit_ig, order = 2)
```

## 8.2 The second difference

The second difference measures whether the gains are shrinking. A response with
diminishing returns has negative second differences throughout, and the point
where the first difference crosses zero is the optimum.

## 8.3 The derivative is refused

```{r derivative-refused, error = TRUE}
agri_np_derivative(fit_ig)
```

```{r sizer-refused, error = TRUE, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_sizer(fit_ig)
```

Both refusals name the alternative. A derivative describes the rate of change of
a function defined on a continuum; there is no continuum here.

## 8.4 Interpretation

Report first differences beside the fitted values. A table of predicted yields
without them leaves the reader to compute the only quantity that supports a
decision.

---

# 9. Efficiency: response per unit of input

```{r efficiency}
eff <- agri_integer_efficiency(fit_ig)
eff
```

## 9.1 A different question from the optimum

The optimum maximises the **response**. Efficiency maximises the response **per
unit of input**, and the two rarely coincide.

```{r efficiency-compare}
e <- as.data.frame(eff)
names(e)
ecol <- grep("efficien|ratio|per_unit", names(e), value = TRUE)[1]
xcol <- names(e)[1]
data.frame(
  criterion = c("maximum response", "maximum response per unit of input"),
  decision  = c(agri_integer_optimum(fit_ig)$optimum,
                if (!is.na(ecol)) e[[xcol]][which.max(e[[ecol]])] else NA)
)
```

The efficiency optimum is almost always **lower** than the yield optimum,
because the first units of input are the most productive. Which criterion applies
depends on what is scarce.

| Scarce resource | Criterion |
|---|---|
| land | maximum response per unit area, the yield optimum |
| seed, water, fertilizer | maximum response per unit of that input |
| capital | maximum net return, which needs prices |

## 9.2 Interpretation

State the criterion explicitly. "The optimum density was 8 plants" is incomplete
without saying whether that maximises yield or yield per seed.

---

# 10. Threshold rules

Three rules are available, and each matches a different decision context.

## 10.1 A fraction of the maximum

```{r threshold95}
agri_integer_threshold(
  fit_ig,
  criterion = "fraction_of_maximum",
  value     = 0.95
)
```

The smallest decision reaching 95% of the maximum response. This is the rule for
a grower who wants nearly all the benefit at a lower input, and it is often far
below the optimum because the response is flat near its peak.

## 10.2 A gain over a baseline

```{r threshold-gain}
agri_integer_threshold(
  fit_ig,
  criterion = "gain_from_baseline",
  value     = 15,
  baseline  = 1
)
```

The smallest decision achieving a stated absolute gain over a reference. This is
the rule when the question is "is the change worth making at all", with the
baseline being current practice.

## 10.3 A marginal gain

```{r threshold-marginal}
agri_integer_threshold(
  fit_ig,
  criterion = "marginal_gain",
  value     = 1
)
```

The largest decision at which one more unit still gains at least the stated
amount. This is the economic rule: set `value` to the input cost expressed in
response units, and the answer is the profit-maximising decision.

## 10.4 Choosing among them

```{r threshold-compare}
data.frame(
  rule = c("fraction_of_maximum", "gain_from_baseline", "marginal_gain"),
  question = c("how little can I use and still get nearly all of it",
               "is the change worth making at all",
               "where does the next unit stop paying for itself"),
  needs = c("a fraction", "a baseline and a gain", "a cost in response units")
)
```

## 10.5 Interpretation

Report which rule was applied and the value used. A recommended density without
its decision rule is not reproducible, because three rules give three answers
from the same fit.

---

# Part III. Uncertainty on a discrete support

# 11. Probability mass, not an interval

```{r bootstrap-optimum}
boot_opt <- agri_integer_bootstrap(
  fit_ig,
  B    = 199,
  seed = 4105
)
boot_opt
```

## 11.1 Why a mass and not an interval

The optimum is a discrete quantity. Across resampled experiments it takes a
handful of values, and the natural summary is **how often each one was best**.

A symmetric interval would be misleading in two ways: it can include values that
were never optimal in any replicate, and it hides an asymmetric or bimodal
distribution, which is common when two adjacent densities are nearly tied.

```{r boot-plot, fig.cap = "Probability mass over admissible decisions. The height of each bar is the share of resampled experiments in which that density was best."}
plot(boot_opt)
```

## 11.2 Reading the mass

| Pattern | Reading |
|---|---|
| one tall bar | the decision is well determined |
| two adjacent bars of similar height | the two densities are practically tied; either is defensible |
| a wide flat spread | the experiment does not determine the decision |
| mass at an end of the range | the optimum may lie outside the tested densities |

The last row is the discrete counterpart of the boundary problem in the
continuous case, and it carries the same warning: the experiment may not have
tested far enough.

## 11.3 The resampling unit

The bootstrap resamples whole blocks when a block is declared, exactly as in the
continuous case, because plots within a block are not exchangeable with plots in
other blocks.

---

# 12. Discrete confidence sets

```{r confset}
agri_integer_confset(boot_opt, level = 0.95)
```

## 12.1 What it contains

Only admissible decisions, accumulated from the most probable downward until the
stated level is reached. A set of three densities is a more useful answer to a
grower than a single number with a symmetric interval that includes densities
nobody can plant.

```{r confset-plot, fig.cap = "The confidence set. Decisions outside it are faded rather than removed, so the reader can see what was excluded."}
plot(agri_integer_confset(boot_opt, level = 0.95))
```

## 12.2 Why excluded decisions are faded, not deleted

A figure that shows only the set gives no sense of how close the excluded
decisions were. Fading them keeps that information visible, which matters when
the set boundary is nearly arbitrary.

## 12.3 Interpretation

Report the set, not only the point. "Density 7, 95% set {6, 7, 8}" is a complete
statement; "density 7" is not.

---

# Part IV. Interpolation and the tested support

# 13. What a sparse trial can and cannot support

```{r support-comparison}
set.seed(4106)

sparse <- data.frame(plants = rep(c(1, 3, 5, 7, 9), each = 6))
sparse$yield <- 20 + 7 * sparse$plants -
  0.48 * sparse$plants^2 +
  rnorm(nrow(sparse), 0, 1.2)

fit_observed <- agri_np_regression(
  yield ~ plants, data = sparse, method = "integer_grid",
  integer_base_method = "smoothing_spline",
  predictor_support   = "observed_integer"
)

fit_full <- agri_np_regression(
  yield ~ plants, data = sparse, method = "integer_grid",
  integer_base_method = "smoothing_spline",
  predictor_support   = "integer_range",
  integer_range       = c(1, 9)
)

agri_integer_predict(fit_observed)
```

```{r support-full}
agri_integer_predict(fit_full)
```

## 13.1 The two tables answer different questions

`observed_integer` reports only the five densities actually tested. It makes no
claim about 2, 4, 6 or 8.

`integer_range` reports all nine. It **interpolates**, and that interpolation is
a modelling assumption: that nothing biologically distinct happens at the
untested densities.

## 13.2 When interpolation is defensible

| Situation | Interpolate |
|---|---|
| densities 1, 3, 5, 7, 9 in a smooth competition response | yes, usually |
| sprays 0, 2, 4 where 1 and 3 are operationally impossible | no; use `custom_integer` |
| irrigation events where an even number matters for scheduling | no; declare the admissible set |
| plant densities where the seeder has fixed settings | no; declare the settings |

## 13.3 The optimum can differ

```{r support-optima}
cat("observed only:\n"); print(agri_integer_optimum(fit_observed))
cat("\nfull range:\n");  print(agri_integer_optimum(fit_full))
```

If the two differ, the recommendation depends on an interpolation the experiment
did not test. Report which support was declared.

## 13.4 Interpretation

Declare the support explicitly and justify it. `observed_integer` is the
conservative default and should be preferred unless there is a reason to
interpolate.

---

# Part V. Reporting

# 14. Tables

```{r reporting}
agri_table(fit_ig, what = "integer_predictions", format = "data.frame")
```

```{r reporting-optimum}
agri_table(fit_ig, what = "integer_optimum", format = "data.frame")
```

```{r reporting-efficiency}
agri_table(fit_ig, what = "integer_efficiency", format = "data.frame")
```

# 15. Figures

```{r figures, fig.cap = "Gain from one more unit. Where it crosses zero is the optimum, and where it crosses the input cost is the economic decision."}
agri_np_plot(fit_ig, type = "difference")
```

```{r efficiency-fig, fig.cap = "Response per unit of input. Its maximum is usually at a lower density than the yield maximum."}
agri_np_plot(fit_ig, type = "efficiency")
```

---

# Part VI. Common mistakes, and the function that prevents each

# 16. Rounding a continuous optimum

**The mistake.** Fitting a curve over density, finding the maximum at 7.4, and
recommending 7.

**Why it is wrong.** The rounded value is the neighbour of an inadmissible one,
not the best admissible decision. They can differ. See section 1.

**What prevents it.** `predictor_support = "observed_integer"` and
`agri_integer_optimum()`, which evaluate on the lattice.

---

# 17. Predicting at a fractional value

**The mistake.** Reporting the fitted response at 4.5 insects.

**Why it is wrong.** The value describes an observation that cannot occur, and it
propagates into thresholds and recommendations.

**What prevents it.** `agri_np_predict()` refuses once the support is declared.
See section 4.2.

---

# 18. Reporting a derivative

**The mistake.** "The marginal response was 1.2 units per plant, from the
derivative of the fitted curve."

**Why it is wrong.** A derivative is not an admissible quantity on a discrete
support.

**What prevents it.** `agri_np_derivative()` and `agri_np_sizer()` both refuse
and name `agri_integer_difference()` instead. See section 8.3.

---

# 19. Drawing a continuous line through integer values

**The mistake.** A smooth curve through predicted yields at 1 to 12 plants.

**Why it is wrong.** It suggests a value exists between two admissible decisions.

**What prevents it.** `agri_np_plot()` draws steps and crosses for an
integer-support fit. See section 4.1.

---

# 20. Reporting a symmetric interval for a discrete optimum

**The mistake.** "Optimum 7, 95% CI 5.8 to 8.2."

**Why it is wrong.** The bounds are not admissible decisions, and the interval
hides an asymmetric or bimodal mass.

**What prevents it.** `agri_integer_bootstrap()` returns a probability mass and
`agri_integer_confset()` a set of admissible values. See sections 11 and 12.

---

# 21. Reporting an optimum without its criterion

**The mistake.** "The optimum density was 8 plants."

**Why it is wrong.** Yield-maximising, efficiency-maximising and
economically-optimal densities differ, often substantially.

**What prevents it.** `agri_integer_efficiency()` and
`agri_integer_threshold()` make the criterion an explicit argument. See sections
9 and 10.

---

# 22. Interpolating on a sparse support without saying so

**The mistake.** Testing 1, 3, 5, 7, 9 and recommending 6.

**Why it is wrong.** Density 6 was never tested. It may be a fine recommendation,
but it rests on an interpolation the reader cannot see.

**What prevents it.** `predictor_support = "observed_integer"` is the
conservative default, and `integer_range` records the decision to interpolate.
See section 13.

---

# 23. Imposing a unimodal constraint on a monotone range

**The mistake.** Using `unimodal_isotonic` when the trial never reached the
density at which competition begins.

**Why it is wrong.** The constraint forces a peak that may lie outside the data,
and the fitted optimum then sits at the boundary and means nothing.

**What prevents it.** Fitting an unconstrained engine first. See section 5.2.

---

# Part VII. Compact selection guide

# 24. Choose the engine

| Situation | Engine |
|---|---|
| no shape assumption, no block | `discrete_kernel` |
| one peak known, no block | `unimodal_isotonic` |
| one peak known, blocked design | `umbrella` |
| no shape assumption, blocked design | `integer_grid` with `gam` |
| a familiar smoother is wanted | `integer_grid` with your base engine |

# 25. Choose the support

| The untested values are | `predictor_support` |
|---|---|
| operationally impossible | `custom_integer` |
| biologically uninteresting but possible | `integer_range` |
| simply untested, and you will not recommend them | `observed_integer` |

# 26. Choose the decision rule

| The question | Function and argument |
|---|---|
| what maximises yield | `agri_integer_optimum()` |
| what maximises yield per unit of input | `agri_integer_efficiency()` |
| how little can I use for nearly all the benefit | `threshold("fraction_of_maximum")` |
| is the change worth making at all | `threshold("gain_from_baseline")` |
| where does the next unit stop paying | `threshold("marginal_gain")` |
| how sure am I | `agri_integer_bootstrap()`, `agri_integer_confset()` |

---

# Part VIII. Minimum reporting checklist

# 27. What the methods section must contain

1. that the treatment support is integer, and why;
2. the engine, and the reason it was admissible;
3. the declared support mode, and the justification for any interpolation;
4. any shape constraint, and its biological basis;
5. the decision criterion, named, with its parameter value;
6. the optimum, the probability mass over decisions, and the confidence set;
7. first differences beside the predicted values;
8. the number of bootstrap replicates and the seed;
9. the package version.

# 28. A worked methods paragraph

> Plant density took whole-number values only, so the response was evaluated on
> the admissible integer support rather than on a continuous scale
> (`agri_np_regression(predictor_support = "observed_integer")` from agriRank
> 0.14.0). A unimodal constraint was imposed because plant competition is known
> to produce a single peak. The optimum was defined as the best admissible
> density, obtained by evaluation and not by rounding a continuous maximum.
> Uncertainty is reported as the probability mass over admissible densities from
> a block-resampling bootstrap (`agri_integer_bootstrap()`, 999 replicates, seed
> 4105) and as a 95% discrete confidence set. First differences, the yield gained
> from one additional plant, are given in Table 2, and the economically optimal
> density under a seed cost of 1.0 yield units per plant was obtained with
> `agri_integer_threshold(criterion = "marginal_gain")`.

---

# 29. Where to go next

| If you now want | Read |
|---|---|
| continuous gradients | *Nonparametric and Shape-Aware Regression* |
| where a continuous response stops changing | *Distribution-Free Uncertainty and Model Checking* |
| a continuous rate with an interval on its location | *Optima, Quantiles, and How the Block Enters the Model* |
| the design the fit sits inside | *Design Foundations, CRD, and RCBD* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# Part IX. Glossary

# 30. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **integer support** | the set of whole-number values a treatment may take |
| **admissible decision** | a value the grower can actually implement |
| **estimand** | the quantity being estimated, as distinct from the estimator |
| **projection** | evaluating a latent continuous fit on the admissible lattice |
| **first difference** | the change in response from one more unit |
| **second difference** | whether those gains are shrinking |
| **efficiency** | response per unit of input |
| **threshold rule** | a stated criterion for choosing a decision |
| **marginal gain** | what one more unit adds, compared with what it costs |
| **probability mass** | the share of resampled experiments in which each decision was best |
| **discrete confidence set** | a set of admissible decisions covering the stated level |
| **interpolation** | claiming a value for an untested admissible level |

---

# Selected methodological references

- Barlow, R. E., Bartholomew, D. J., Bremner, J. M., and Brunk, H. D. (1972). *Statistical Inference Under Order Restrictions*. Wiley.
- Li, Q., and Racine, J. S. (2007). *Nonparametric Econometrics: Theory and Practice*. Princeton University Press.
- Meyer, M. C. (2013). Semi-parametric additive constrained regression. *Journal of Nonparametric Statistics*, 25(3), 715-730.
- Racine, J., and Li, Q. (2004). Nonparametric estimation of regression functions with both categorical and continuous data. *Journal of Econometrics*, 119, 99-130. https://doi.org/10.1016/S0304-4076(03)00157-X
- Turner, T. R., and Wollan, P. C. (1997). Locating a maximum using isotonic regression. *Computational Statistics and Data Analysis*, 25, 305-320. https://doi.org/10.1016/S0167-9473(97)00009-1
- Wang, M. C., and van Ryzin, J. (1981). A class of smooth estimators for discrete distributions. *Biometrika*, 68(1), 301-309.

The package also ships a verified RIS library under `inst/references/`.
