---
title: "Rank-based repeated-measures inference"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rank-based repeated-measures inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

Repeated observations on the same experimental unit are dependent. The correct analysis must distinguish between-subject treatment structure from within-subject time or occasion structure.



## State of the art

`nparLD` provides rank-based relative effects and ANOVA-/Wald-type inference for longitudinal factorial designs. `MANOVA.RM` provides resampling-based multivariate and repeated-measures inference. Wild-bootstrap rank procedures improve finite-sample calibration for general repeated-measures/split-plot settings. `permuco` offers permutation ANOVA machinery that can represent repeated-error strata. These approaches target different estimands and are not interchangeable merely because all avoid a simple Gaussian repeated-measures ANOVA.

## `np_repeated()`: three examples

```r
x <- simulate_agri("repeated", seed = 71, n = 8)
np_repeated(height ~ treatment * time, x, subject = subject, within = time, method = "auto")
if (requireNamespace("nparLD", quietly = TRUE))
  np_repeated(height ~ treatment * time, x, subject = subject, within = time, method = "nparLD")
xb <- transform(x, block = factor(rep(1:4, length.out = nrow(x))))
if (requireNamespace("permuco", quietly = TRUE))
  np_repeated(height ~ treatment * time, xb, subject = subject, within = time, block = block, method = "permuco")
```

## `agri_repeated()`: three backend examples

```r
d <- agri_design(height ~ treatment * time, x, design = "repeated", subject = subject, within = time)
if (requireNamespace("nparLD", quietly = TRUE)) agri_repeated(d, backend = "nparLD")
if (requireNamespace("MANOVA.RM", quietly = TRUE)) agri_repeated(d, backend = "MANOVA.RM", iter = 499)

xm <- simulate_agri("repeated_missing", seed = 72, n = 8, missing_rate = .12)
dm <- agri_design(height ~ treatment * time, xm, design = "repeated", subject = subject, within = time)
agri_repeated(dm, backend = "native_wild", B = 499, missing_assumption = "MCAR")
```

## Backend choice

| Backend | Primary target | Missing response values | Agronomic block in current adapter | Typical role |
|---|---|---:|---:|---|
| `nparLD` | relative effects / rank-based ATS-WTS | specialized complete/longitudinal structures | no | complete unblocked repeated design |
| `MANOVA.RM` | resampling-based multivariate/repeated inference | depends on model/backend | no | semiparametric sensitivity/alternative |
| `permuco` | permutation model with error strata | not routed as all-available incomplete method | yes, complete data | blocked repeated designs |
| `native_wild` | relative marginal effects with wild ATS/WTS/MATS | yes | no | incomplete unblocked repeated designs |

`agriRank` deliberately refuses to ignore a declared agronomic block to make an incompatible repeated-measures engine run.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.
