---
title: "Missing data in repeated-measures experiments"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data in repeated-measures experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

Missing repeated observations change both the available information and the covariance estimation problem. The package characterizes missingness first and never silently imputes values or deletes entire subjects.



## State of the art

Amro, Konietschke, and Pauly developed quadratic-form rank procedures for incompletely observed nonparametric factorial repeated-measures designs, allowing ordinal outcomes and singular covariance matrices. The native `agriRank` engine implements the corresponding all-available rank construction with wild bootstrap and is marked experimental until numerical benchmarking against reference implementations is complete. The principal theoretical justification is under MCAR; observed data alone cannot prove MCAR, MAR, or MNAR.

## `agri_missing_report()`: three examples

```r
# 1. Sparse missingness
x1 <- simulate_agri("repeated_missing", seed = 81, n = 10, missing_rate = .10)
d1 <- agri_design(height ~ treatment * time, x1, design = "repeated", subject = subject, within = time)
agri_missing_report(d1)

# 2. Monotone dropout created deliberately
x2 <- simulate_agri("repeated", seed = 82, n = 6)
x2$height[x2$subject %in% levels(x2$subject)[1:2] & as.numeric(x2$time) >= 3] <- NA
agri_missing_report(x2, response = "height", subject = "subject", within = "time")

# 3. Higher missingness
x3 <- simulate_agri("repeated_missing", seed = 83, n = 10, missing_rate = .25)
d3 <- agri_design(height ~ treatment * time, x3, design = "repeated", subject = subject, within = time)
agri_missing_report(d3)
```

## `incomplete_wild_rank_test()`: three examples

```r
incomplete_wild_rank_test(d1, B = 499, statistic = "ATS", weights = "rademacher", missing_assumption = "MCAR")
incomplete_wild_rank_test(d1, B = 499, statistic = "WTS", weights = "mammen", missing_assumption = "MCAR")
incomplete_wild_rank_test(d1, B = 499, statistic = "MATS", weights = "normal", missing_assumption = "MAR-sensitivity")
```

## `agri_missing_sensitivity()`: three examples

```r
agri_missing_sensitivity(d1, B = 299, seed = 84)
agri_missing_sensitivity(d3, B = 299, seed = 85, statistic = "ATS")
d_complete <- agri_design(height ~ treatment * time, simulate_agri("repeated", seed = 86),
                          design = "repeated", subject = subject, within = time)
agri_missing_sensitivity(d_complete)
```

## What this analysis can and cannot establish

A missingness report is descriptive. A discrepancy between all-available and complete-subject analyses is a sensitivity signal, not proof of a missingness mechanism. Scientific knowledge about why measurements are absent remains essential. For blocked repeated experiments with missing observations, the current native engine stops rather than discarding the block stratum.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.
