---
title: "Integrated Agronomic Case Study: Design to Report"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated Agronomic Case Study: Design to Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.8,
  fig.height = 4.4,
  dpi = 130
)
```

**Integrated instructional vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Purpose:** a practical starting point for students, researchers, reviewers and
analysts who need to move from a field notebook to a defensible agronomic
recommendation, without losing the logic of the experimental design along the
way.

---

# 1. Why this vignette exists

Nonparametric analysis becomes difficult when several decisions arrive at the
same time: which unit was randomized, whether a factor is qualitative or a
gradient, whether the response is a measurement at all, which comparisons are
admissible, how much uncertainty a recommendation carries, and what to report
when the data cannot answer the question that was asked.

`agriRank` is organised so that those decisions can be taken in a defensible
order. The central rule is short:

> **Declare the randomization first. Then choose the analysis the design
> permits. Check the fit before interpreting it. Report every estimate with the
> uncertainty it actually carries, and report nothing the data do not contain.**

That last clause does more work than it appears to. A large part of this package
consists of functions that decline to answer, by name and with a reason, rather
than return a number that looks like an answer. This vignette shows the whole
sequence on one experiment, and every refusal it produces is treated as a
result, not as an obstacle.

The other vignettes each own one analytical block. This one connects them.

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. identify the experimental unit and the randomization before choosing any
   test, and recognise when a design declaration has been made incorrectly;
2. represent a randomized complete block factorial explicitly, and understand
   what the declaration then forbids;
3. distinguish an omnibus rank test from the comparisons that may follow it;
4. read relative marginal effects rather than treatment means when the response
   is analysed by ranks;
5. produce compact letter displays that are computed inside the correct stratum,
   and explain why a display computed across strata is not interpretable;
6. decide whether a quantitative treatment should be analysed as a factor or as
   a gradient, and defend the decision on design grounds rather than on fit;
7. fit a nonparametric regression without selecting the engine by p-value;
8. separate the three explained-variation indices and say what each one answers;
9. report where a response is still changing, rather than where a fitted curve
   happens to peak;
10. attach a confidence interval to the **location** of an optimum, and detect
    when no interior optimum is identified;
11. distinguish an interval for a fitted curve from an interval for the next
    plot, and produce the second when the recommendation concerns a plot;
12. check a model without assuming a distribution, and know which of the checks
    has power;
13. analyse a treatment whose admissible values are whole numbers, without
    rounding a continuous optimum;
14. analyse experiments whose datum is a censored time or an order rather than
    a measurement;
15. run a sensitivity analysis and a missing-data sensitivity analysis as
    routine parts of the workflow rather than as afterthoughts;
16. assemble tables, figures and a reproducible report from the fitted objects;
17. recognise the recurring mistakes this package is built to prevent, and name
    the function that prevents each one.

---

# 3. The package in one map

## 3.1 Instructional stages

| Stage | What decision it settles | Main functions |
|---|---|---|
| Design | which unit was randomized, and what that permits | `agri_design()`, `validate_agri_design()`, `design_summary()`, `simulate_agri()` |
| Omnibus | is there an effect at all, in the declared structure | `agri_rank()`, `np_crd()`, `np_rcbd()`, `np_factorial()`, `np_splitplot()`, `np_splitsplit()`, `np_stripplot()` |
| Effects | how large, on a scale ranks support | `agri_effects()`, `agri_format_ci()` |
| Comparisons | which pairs differ, with multiplicity controlled | `agri_pairs()`, `agri_conover()`, `agri_cld()`, `agri_contrast()` |
| Gradients | what shape the response has over a quantitative treatment | `agri_np_regression()`, `agri_np_predict()`, `agri_np_plot()`, `agri_np_compare()`, `agri_np_diagnostics()` |
| Recommendation | which rate to recommend, for whom, and with what interval | `agri_np_sizer()`, `agri_np_significant_slope()`, `agri_np_optimum_test()`, `agri_np_quantile_curves()` |
| Uncertainty | what covers the curve, and what covers the next plot | `agri_np_bootstrap()`, `agri_np_conformal()`, `agri_np_coverage()` |
| Model checking | does the fit describe the data, without a distribution | `agri_np_simdiag()` |
| Integer decisions | treatments that can only be whole numbers | `agri_integer_predict()`, `agri_integer_optimum()`, `agri_integer_difference()`, `agri_integer_confset()` |
| Repeated and missing | subject dependence, and incompleteness | `agri_repeated()`, `np_repeated()`, `agri_missing_report()`, `agri_missing_sensitivity()` |
| Several responses or sites | multivariate and multi-environment structure | `agri_multivariate()`, `agri_multienv()` |
| Non-measurements | censored times, and orders | `agri_np_timetoevent()`, `agri_rankings()` |
| Sensitivity | does the conclusion depend on one analytical choice | `agri_sensitivity()`, `agri_batch()` |
| Communication | tables, figures, reports, export | `agri_table()`, `agri_plot()`, `agri_report()`, `agri_dashboard()`, `agri_theme()`, `agri_save_figure()`, `export_results()` |

The progression is deliberate. A reader should not begin with conformal
prediction because it carries a guarantee, or with quantile curves because they
produce an attractive figure. Each extension should answer a question that the
experiment actually raises.

## 3.2 Inspect the package programmatically

```{r map}
# Which methods does the package know about, and what does each assume?
head(agri_methods(), 10)
```

The methods table is the honest starting point on a teaching computer: several
engines are optional, and a method that is listed but whose backend is absent
will say so by name rather than fall back silently to something else.

---

# 4. The experiment

## 4.1 Scientific problem

A researcher evaluates **three bean cultivars** under **four salinity levels**
in a randomized complete block design with **five blocks**. Two outcomes are
recorded:

1. final dry biomass, measured once at harvest;
2. SPAD, measured repeatedly at four times during the cycle.

The questions, in the order a researcher actually asks them:

| # | Question | Where it is answered |
|---|---|---|
| Q1 | Do cultivars differ in biomass? | Part II |
| Q2 | Does salinity change biomass? | Part II |
| Q3 | Is there a cultivar-by-salinity interaction? | Part II |
| Q4 | Which cultivars differ from which, and by how much? | Part III |
| Q5 | Should salinity be a factor or a gradient? | Part IV |
| Q6 | Up to which salinity is biomass still falling? | Part V |
| Q7 | Is there an identifiable salinity optimum, and where? | Part V |
| Q8 | What happens to the poor plots, not the average plot? | Part V |
| Q9 | What interval covers the **next** plot? | Part VI |
| Q10 | Does the model describe the data? | Part VI |
| Q11 | How does SPAD change over time, and does that depend on cultivar? | Part VII |
| Q12 | What if some repeated SPAD values are missing? | Part VII |
| Q13 | Does the conclusion survive a different analytical choice? | Part VIII |

## 4.2 Simulate the teaching dataset

The dataset is frozen with `set.seed()` so that every number in this vignette is
reproducible. It is realistic enough for instruction but it is **not field
evidence**, and no agronomic recommendation should be drawn from it.

```{r sim}
set.seed(1501)

base <- expand.grid(
  block    = factor(1:5),
  cultivar = factor(c("C1", "C2", "C3")),
  salinity = c(0, 2, 4, 6)
)
base$salinity_f <- factor(base$salinity)

cult_eff <- c(C1 = 0.0, C2 = 1.2, C3 = 0.6)

# Biomass falls with salinity, and C2 tolerates it better than the others.
base$biomass <-
  18 +
  cult_eff[as.character(base$cultivar)] +
  as.numeric(base$block) * 0.25 -
  1.15 * base$salinity +
  0.06 * base$salinity^2 +
  ifelse(base$cultivar == "C2", 0.22 * base$salinity, 0) +
  rnorm(nrow(base), 0, 0.9)

str(base)
```

## 4.3 Look at the data before modelling anything

```{r descriptives}
obs <- aggregate(biomass ~ cultivar + salinity, base, function(v)
  c(mean = mean(v), sd = stats::sd(v)))
tab <- data.frame(obs[, c("cultivar", "salinity")],
                  mean = round(obs$biomass[, "mean"], 2),
                  sd   = round(obs$biomass[, "sd"], 2))
tab
```

These are **observed summaries, not model estimates**. They already suggest a
decline with salinity and a difference between cultivars. They say nothing about
the block structure, about uncertainty, or about whether the decline is the same
for every cultivar.

Look at them as a figure before deciding anything:

```{r raw-figure, fig.cap = "Observed biomass. Every point is one plot. Nothing has been modelled yet."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(
    ggplot2::ggplot(base, ggplot2::aes(x = salinity, y = biomass,
                                       colour = cultivar)) +
      ggplot2::geom_point(alpha = 0.7, size = 1.8) +
      ggplot2::stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
      ggplot2::labs(x = "Salinity (dS/m)", y = "Dry biomass (g)",
                    colour = "Cultivar",
                    caption = "Lines join observed means. They are not fitted curves.") +
      agri_theme()
  )
}
```

Two features are visible and both matter later: the decline is not a straight
line, and the cultivars do not decline in parallel.

---

# Part I. Design before analysis

# 5. Start with the randomization, not the test

## 5.1 The declaration

```{r design}
des <- agri_design(
  biomass ~ cultivar * salinity_f,
  data   = base,
  design = "rcbd",
  block  = block
)
des
```

The declaration is not bookkeeping. It is a statement about how the experiment
was carried out, and it decides what may legitimately follow.

```{r design-summary}
design_summary(des)
```

## 5.2 Validate it

```{r validate}
validate_agri_design(des)
```

A design audit answers questions that no model-selection criterion can repair
afterwards:

- Are all intended treatment combinations represented?
- Which factor was randomized to which unit?
- Which observations share an experimental unit?
- Is a repeated measurement being counted as an independent replicate?
- Are there empty cells?
- Does the declared structure match the randomization that actually happened?

If any of those answers is uncertain, the next step is not analysis. It is
clarification of the design.

## 5.3 What the declaration then forbids

This is where `agriRank` differs from a toolbox of tests. Once a block is
declared, methods that cannot carry it are refused **by name**, rather than
applied to the data with the block quietly dropped:

```{r forbidden, error = TRUE}
agri_rank(des, method = "kruskal")
```

The refusal is the point. A Kruskal-Wallis test on blocked data does not fail
loudly on its own; it simply answers a different question, about a completely
randomized experiment that was not the one performed. Discarding the block
inflates the residual variation with between-block differences that the design
was built to remove.

## 5.4 The same rule in the other direction

An analysis may also be refused because the declaration is missing information
the method needs:

```{r forbidden2, error = TRUE}
np_repeated(biomass ~ cultivar, base)
```

Repeated measurements on one plant are not exchangeable with measurements on
different plants, so the subject has to be declared. The message names the
missing argument and the scientific reason for it.

## 5.5 Interpretation

The design declaration has three consequences that persist through the rest of
the analysis:

1. it fixes the **stratum** in which every subsequent comparison is made;
2. it determines which engines are admissible, and which will be refused;
3. it becomes part of the reported methods, so a reader can reconstruct what was
   assumed.

Nothing later in this vignette overrides it.

---

# Part II. The omnibus analysis

# 6. Rank-based inference for the declared design

## 6.1 Fit

```{r fit}
fit <- agri_rank(des)
fit
```

The engine was chosen from the **structure** of the design, not from the
response. No p-value was consulted in making the choice, and none should be: an
engine selected because it produced the smallest p-value is no longer testing
the hypothesis it appears to test.

```{r omnibus}
fit$omnibus
```

## 6.2 Reading the omnibus table

Three effects are reported and they answer three different questions:

| Effect | Question |
|---|---|
| `cultivar` | averaged over salinity, do cultivars differ? |
| `salinity_f` | averaged over cultivars, does salinity change biomass? |
| `cultivar:salinity_f` | does the salinity effect depend on the cultivar? |

The interaction is the one to read first. If it is present, the two main effects
are averages over conditions in which the behaviour differs, and quoting them as
if they described every cultivar is misleading.

```{r anova}
anova(fit)
```

## 6.3 A result worth pausing on

The two main effects are overwhelming. The interaction is **not detected**.

That deserves attention, because the raw figure of section 4.3 looked as though
the cultivars declined at different rates, and the data were in fact generated
with a real interaction: C2 was given a genuine advantage that grows with
salinity.

So the experiment contains an effect that it cannot resolve. Five blocks are not
enough to separate an interaction of that size from noise. Nothing is wrong with
the analysis; the design simply lacks the power.

This is the single most useful thing a worked example can demonstrate, and it is
why section 35 of this vignette exists. **A non-significant interaction is not a
parallel response.** It is a statement that the experiment could not distinguish
one from the other.

## 6.4 What follows from a non-detected interaction

Two readings are defensible and they should be reported together:

1. **Marginally.** With no evidence of an interaction, cultivar effects may be
   quoted averaged over salinity, and salinity effects averaged over cultivars.
   This is the simpler report and it is what most manuscripts would present.
2. **With the caveat.** The absence of evidence is not evidence of parallelism,
   and the confidence intervals of Part III will show that an agronomically
   meaningful interaction remains compatible with these data.

The gradient analysis of Part IV pursues the second reading further, because a
curve estimated per cultivar can show a difference in shape that a factorial
interaction test, spending six degrees of freedom, cannot detect.

## 6.5 What an omnibus test does not tell you

An omnibus test answers one question: whether the treatment labels carry any
information at all about the ranks. It does not say which treatments differ, in
what direction, or by how much. Those are separate questions with separate
multiplicity problems, and they are the subject of Part III.

A significant omnibus test is a licence to look further. It is not itself a
finding worth reporting alone.

## 6.6 Interpretation

For this experiment the reading is:

- salinity changes biomass, decisively;
- cultivars differ, decisively;
- the experiment cannot say whether they respond to salinity differently.

---

# 7. Effects on a scale that ranks support

## 7.1 Cell summaries and mean ranks

```{r effects}
eff <- agri_effects(fit)
eff
```

Two quantities are reported per treatment cell, and the choice is deliberate.

`median` is a **location on the response scale**, in grams. It is the quantity a
reader wants for agronomic magnitude, and it is the robust counterpart of a
treatment mean: one extreme plot moves it very little.

`mean_rank` is the quantity the **analysis actually used**. The omnibus test of
section 6 and the comparisons of Part III both operate on within-block ranks and
never look at the grams again. Reporting the mean rank beside the median lets a
reader see the analysis, not merely its conclusion.

## 7.2 Why not a treatment mean

A rank-based analysis does not estimate a treatment mean, and printing one beside
a rank test is a category error that appears constantly in the applied
literature. The two objects answer different questions and can order the
treatments differently.

| Property | Treatment mean | Median and mean rank |
|---|---|---|
| Requires a scale with equal intervals | yes | no |
| Affected by a single extreme plot | strongly | weakly |
| Defined for ordinal data | no | yes |
| Invariant to monotone transformation of the response | no | yes for the rank |
| Matches what the test computed | no | yes |

The last row is the one that matters for a manuscript. If the test ranked and the
table means, a reader cannot check the one against the other.

## 7.3 Reading the cell table

```{r effects-read}
e <- as.data.frame(eff)
e$cultivar <- sub("::.*$", "", e$cell)
e$salinity <- as.numeric(sub("^.*::", "", e$cell))
with(e, tapply(median, list(cultivar, salinity), identity))
```

Laid out as a grid, the decline with salinity is visible in every cultivar, and
C2 sits above the others at every level. Whether the **gap** widens with salinity
is exactly the interaction the design could not resolve.

## 7.4 An estimate travels with its interval

```{r format-ci}
# The helper exists so that an estimate and its interval reach a manuscript
# table together, whatever produced them.
agri_format_ci(estimate = 17.9, lower = 16.4, upper = 19.3, digits = 2)
```

Every estimate this package reports carries an interval where one is defined,
and `agri_format_ci()` formats the pair as a single reportable string. An
estimate quoted without its uncertainty is not a result.

## 7.5 Interpretation

Report the cell medians for magnitude and the omnibus test for evidence. Do not
report either alone: the median without a test is anecdote, and the p-value
without the median is a decision with no agronomic content.

---

# Part III. Comparisons that the design permits

# 8. Pairwise comparisons and compact letters

## 8.1 Design-aware Conover comparisons

```{r conover}
cmp <- agri_conover(fit)
head(as.data.frame(cmp), 8)
```

The Conover procedure operates on the same within-block ranks that the omnibus
test used. That continuity matters: a post-hoc procedure computed on a different
quantity from the omnibus test can contradict it, and the reader has no way to
tell which to believe.

## 8.2 Compact letter displays, computed inside the stratum

```{r cld}
cld <- agri_cld(cmp)
cld
```

A compact letter display is the most-read and least-understood object in
agronomic papers. Two rules govern it here.

**First, letters are computed within strata.** In a factorial with an
interaction, a display that pools across the other factor is not interpretable,
because two treatments may share a letter at one salinity and not at another.
`agri_cld()` computes the display inside the stratum in which the comparisons
were made.

**Second, an incomplete family of comparisons is refused.** If some pairs are
missing, the letters cannot be constructed consistently, and a partial display
invites exactly the wrong reading. The function says so instead of producing
something plausible-looking.

## 8.3 What sharing a letter does and does not mean

Sharing a letter means the comparison did not reach the chosen significance
level. It does **not** mean the treatments are equal. With five blocks, the
experiment has limited power, and a real difference of agronomic size can easily
share a letter with zero.

The honest report pairs the letters with the effect sizes of section 7. Letters
alone answer "was it detected"; effects answer "how large", which is the
question that matters agronomically.

## 8.4 Planned contrasts, and an honest limitation

Comparing every pair is rarely what the science asks. A planned contrast asks one
specific question and pays a much smaller multiplicity price. The coefficients
sum to zero, which is what makes it a contrast rather than a weighted average,
and declaring it **before** looking at the data is what preserves its
interpretation. A contrast chosen after seeing which cells are far apart is a
pairwise comparison in disguise, with the multiplicity hidden.

```{r contrast, error = TRUE}
C <- rbind("C2 vs the other two" = c(-0.5, 1, -0.5))
agri_contrast(fit, C)
```

The refusal is worth reading carefully, because it is a limitation of this
version rather than a statement about the data. General user-defined contrasts
are currently implemented only for the native repeated wild-rank engine. For
every other engine the message says so and points to the backend object, instead
of silently computing a contrast on a quantity for which it has not been
validated.

Where a planned contrast is essential and the engine does not support it, the
defensible route is the pairwise machinery of section 8.1 with an appropriate
multiplicity adjustment, accepting the loss of power that comparing more pairs
entails.

## 8.5 Interpretation

The layers answer different questions and should be reported together:

| Layer | Question | Function |
|---|---|---|
| Omnibus | is there anything here | `agri_rank()` |
| Effects | how large, on the response scale and on the rank scale | `agri_effects()` |
| Comparisons | which pairs, with multiplicity controlled | `agri_conover()`, `agri_cld()` |
| Contrast | the one question the study was designed to answer | `agri_contrast()`, where supported |

---

# Part IV. From factor to gradient

# 9. Should salinity be a factor or a curve?

## 9.1 The decision is about the treatment, not about the fit

Salinity was applied at 0, 2, 4 and 6 dS/m. Those are levels of a **quantitative**
treatment, and they can be analysed two ways:

| As a factor | As a gradient |
|---|---|
| makes no assumption about the shape between levels | assumes the response varies smoothly between levels |
| answers: do these four levels differ | answers: what is the shape, and where is it changing |
| cannot interpolate | can interpolate, within the tested range |
| spends 3 degrees of freedom | spends fewer, and gains precision if the shape is real |

The decision should be made on scientific grounds. If the four levels were
chosen as representative points on a continuum, and interpolation between them
is meaningful, the gradient analysis answers the real question. If they are
distinct managements that happen to be labelled with numbers, they are a factor.

Salinity is a continuum. It is analysed as a gradient from here on, with the
factorial analysis of Part II retained as the design-faithful companion.

## 9.2 Both analyses, side by side

```{r factor-vs-gradient}
data.frame(
  analysis = c("factor", "gradient"),
  question = c("do the four levels differ",
               "what shape does biomass follow, and where does it change"),
  reported_in = c("Part II", "Parts IV to VI")
)
```

Reporting both is not indecision. They answer different questions, and a
manuscript that reports only the factorial analysis of a quantitative treatment
has left the shape of the response undescribed.

---

# 10. Fitting the gradient

## 10.1 A first fit

```{r np-fit}
npfit <- agri_np_regression(biomass ~ salinity, base, method = "gam",
                            block = block)
npfit
```

The block is retained as an adjustment term. An engine that cannot carry it is
refused, exactly as in Part I:

```{r np-forbidden, error = TRUE}
agri_np_regression(biomass ~ salinity, base, method = "loess", block = block)
```

## 10.2 Comparing engines without p-value shopping

```{r np-compare}
cmp_eng <- agri_np_compare(
  biomass ~ salinity, base,
  methods = c("gam", "kernel", "quantile"),
  block = block, kfold = 5, seed = 1
)
cmp_eng
```

Note the `failures` column. An engine that cannot carry the declared block is
recorded as a failure rather than silently dropped from the comparison, so the
table cannot mislead by omission.

Read this table as **predictive** comparison, not as model selection for
inference. Cross-validated error says which engine predicts a held-out plot
best. It does not say which engine gives the correct p-value, and choosing the
engine that produced the most convenient p-value is the single most common way
to invalidate a nonparametric analysis.

The package therefore separates the two operations deliberately:
`agri_np_compare()` reports predictive error, and nothing in it feeds back into
the choice of an inferential engine.

## 10.3 The fitted curve

```{r np-plot, fig.cap = "Fitted salinity response with observed plots. The band is the analytic interval for the curve, not for a plot."}
agri_np_plot(npfit, points = TRUE)
```

## 10.4 Explained variation, and what each index answers

```{r np-diag}
agri_np_diagnostics(npfit, cv = TRUE, seed = 1)$r2
```

Three indices are reported side by side because they answer three questions and
routinely disagree:

| Index | Question | Fails to notice |
|---|---|---|
| `pseudo_r2` | how much variation does the fitted curve reproduce | overfitting: it always rises with flexibility |
| `cv_r2` | how much would it reproduce on a plot it has not seen | nothing, but it is noisy in small experiments |
| `spearman_r2` | how much of the **ordering** does it get right | the size of the departures |

A flexible engine typically shows a larger `pseudo_r2`, a larger
`effective_df` and a **smaller** `cv_r2` than a rigid one on the same data. That
pattern is the signature of a curve that is following noise, and it is visible
only because the three are reported together.

## 10.5 Interpretation

Report `pseudo_r2` and `cv_r2` together, always. A single R-squared-like number
beside a flexible nonparametric fit is uninformative at best, because the reader
cannot tell whether it describes the response or the sampling noise.

---

# Part V. From a curve to a recommendation

# 11. Where is the response still changing?

## 11.1 The trap of the fitted extreme

```{r optimum-naive}
agri_np_optimum(npfit, objective = "min")
```

That is a number with no uncertainty attached, and it may well sit on the edge
of the tested range. An extremum has to land somewhere; if the response does not
turn over inside the range, it lands on a boundary, and reading the boundary as
a recommendation is an artefact of the fitting, not a finding about the crop.

## 11.2 The experiment decides which analysis is available

The natural next step is a SiZer map, which classifies the derivative of the
fitted curve across a whole column of bandwidths. On this experiment it is
refused:

```{r sizer-refused, error = TRUE, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_sizer(npfit)
```

Salinity was applied at **four** levels. A map of the derivative across
bandwidths needs more positions than that, because with four points every
bandwidth sees essentially the same three gaps and the column carries no
information about robustness to smoothing.

This is a **design** limitation, not a software one, and no analytical choice
repairs it. A four-level design supports the factorial comparison of Part III
and a fitted curve for description. It does not support a claim about where the
response stops changing.

The lesson is worth stating in its general form: **if a recommendation curve is
the objective, the experiment has to be designed for it.** Six to eight levels
with adequate replication is the usual minimum, and that decision is made before
the field is laid out, not after the data arrive.

## 11.3 A companion experiment that was designed for a gradient

The package ships a nitrogen response with eight rates in five complete blocks,
which is what the rest of Part V requires.

```{r dose-data}
data(agri_dose)
c(levels = length(unique(agri_dose$dose)), blocks = nlevels(agri_dose$block),
  plots = nrow(agri_dose))

gfit <- agri_np_regression(yield ~ dose, agri_dose, method = "gam",
                           block = block)
agri_np_optimum(gfit)
```

The fitted maximum sits at the top of the tested range and `at_boundary` says so.
That is the trap of section 11.1 in its commonest form: this response plateaus,
and a maximum has to land somewhere.

## 11.4 An answer that survives the smoothing choice

```{r sizer, eval = requireNamespace("SiZer", quietly = TRUE)}
sz <- agri_np_sizer(gfit)
sz
```

```{r sizer-map, eval = requireNamespace("SiZer", quietly = TRUE), fig.cap = "SiZer map. Rows are amounts of smoothing, columns positions on the nitrogen axis. Read it vertically: a position classified the same way in every row does not depend on the smoothing choice."}
plot(sz, type = "map")
```

```{r sizer-statement, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_significant_slope(sz, stability = 0.8)
```

This is the sentence a manuscript can carry: the interval over which yield is
still rising, at a stated level of agreement across bandwidths. It is a different
and far more defensible claim than the boundary maximum of section 11.3.

## 11.5 Interpretation

When the two disagree, the SiZer statement is the one to report. It describes the
data; the fitted extreme describes the fit.

---

# 12. The location of an optimum, and its interval

## 12.1 A recommended level is a location

`agri_np_bootstrap()` resamples the **height** of the curve.
`agri_np_optimum_test()` resamples its **position**. Those are different and
unequally hard quantities: a curve can be estimated precisely while the position
of its extreme wanders widely from one resampled experiment to the next, and a
response that flattens is exactly the shape that makes it wander.

```{r opt-test}
ot <- agri_np_optimum_test(gfit, B = 199, seed = 1, n = 120, external = FALSE)
ot
```

Three columns carry the result.

`optimum` is the point estimate. `lower` and `upper` bound its **location**, not
its height. `p_boundary` is the share of resampled experiments whose extreme
landed on an end of the searched range, and `identified` turns `FALSE` when that
share reaches one half.

```{r opt-dist, fig.cap = "Where the extreme lands across resampled experiments. Mass piled against one end of the range is the visual form of an unidentified optimum."}
plot(ot, type = "distribution")
```

## 12.2 The resampling respects the blocks

Note the header line of the printed object: the resampling unit is the declared
block, not the individual plot. Whole blocks are resampled together, because
plots within a block are not exchangeable with plots in other blocks. That is
what declaring a block asserts, and the bootstrap honours it.

`B = 199` keeps this vignette fast. Use `B >= 999` for anything reported.

## 12.3 Comparing the response of two cultivars

Part II could not resolve the interaction. A curve fitted per cultivar is a
second, more economical way to ask the same question, and to ask it the model
must allow the curve **shapes** to differ, not merely their heights.

```{r additive-refused, error = TRUE}
fit_add <- agri_np_regression(biomass ~ salinity + cultivar, base,
                              method = "gam", block = block)
agri_np_optimum_test(fit_add, by = cultivar, B = 49, objective = "min",
                     external = FALSE)
```

The refusal is arithmetic, not statistical. An additive adjustment shifts one
curve above the other **without changing its shape**, so parallel curves have
their extremes at the same place by construction. A comparison would report a
difference of exactly zero and a p-value of one, describing the model rather
than the experiment.

```{r varying}
fit_var <- agri_np_regression(biomass ~ salinity + cultivar, base,
                              method = "gam", block = block,
                              gam_structure = "varying")
fit_var$formula_used
```

One smooth of salinity is now fitted per cultivar:

```{r opt-by}
ot2 <- agri_np_optimum_test(fit_var, by = cultivar, objective = "min",
                            B = 199, seed = 1, n = 120, external = FALSE)
ot2$optimum
```

```{r opt-curve, fig.cap = "Fitted response per cultivar with the resampled interval for the location of the extreme. The horizontal bar is uncertainty along the salinity axis, not along the biomass axis."}
plot(ot2, type = "curve")
```

## 12.4 Interpretation

Report the interval for the location, and report `p_boundary`. When
`identified` is `FALSE`, do not report a level: report the SiZer statement of
section 11 instead, which describes where the response is still changing without
inventing a turning point.

---

# 13. The typical plot and the poor plot

## 13.1 Why a mean curve is not enough

Every curve so far describes a central tendency. That is a strong restriction on
the agronomic question it can answer.

A cultivar can hold up well in good plots and collapse in poor ones. The average
response then looks acceptable, and a recommendation built on it will disappoint
exactly the growers whose fields resemble the poor plots. Conditional quantiles
ask the question directly: the median describes the typical plot, the tenth
percentile is the **exposure curve**, what a grower meets in a bad year.

```{r quantiles, eval = requireNamespace("qgam", quietly = TRUE)}
q10 <- agri_np_regression(biomass ~ salinity, base,
                          method = "smooth_quantile", tau = 0.10, block = block)
q50 <- agri_np_regression(biomass ~ salinity, base,
                          method = "smooth_quantile", tau = 0.50, block = block)

nd <- data.frame(salinity = c(0, 2, 4, 6),
                 block = factor("3", levels = levels(base$block)))
data.frame(
  salinity = nd$salinity,
  q10 = round(as.numeric(agri_np_predict(q10, nd)), 2),
  q50 = round(as.numeric(agri_np_predict(q50, nd)), 2),
  gap = round(as.numeric(agri_np_predict(q50, nd)) -
              as.numeric(agri_np_predict(q10, nd)), 2)
)
```

Each curve is fitted by the pinball loss, which defines the quantile directly.
Nothing is assumed about the shape of the response distribution, only about the
smoothness of each curve.

## 13.2 The fan, and the replication it needs

A tail quantile needs observations in that tail. Sixty plots is thin for a
tenth percentile, and the function says so rather than fitting one from three
values.

```{r fan, eval = requireNamespace("qgam", quietly = TRUE)}
qc <- suppressWarnings(
  agri_np_quantile_curves(biomass ~ salinity, base, block = block,
                          quantiles = c(0.25, 0.5, 0.75), n = 50))
qc$summary
```

Read `deviation` before anything else. It is the gap between the share of plots
falling below each fitted curve and the quantile that curve claims to be,
measured on the very data it was fitted to. A quantile that misses its own
target here is not a tail this experiment can resolve, and `tracking` flags it.

```{r fan-plot, eval = requireNamespace("qgam", quietly = TRUE), fig.cap = "Smooth conditional quantiles. Curves that fan out mean the treatment changes the spread of biomass, not only its level."}
plot(qc, type = "fan")
```

```{r spread-plot, eval = requireNamespace("qgam", quietly = TRUE), fig.cap = "Distance between the outer quantiles along the gradient."}
plot(qc, type = "spread")
```

## 13.3 Interpretation

If the spread widens with salinity, the stress raises variability as well as
lowering the level, and those are different agronomic problems. A recommendation
that averages over them is not the same recommendation a risk-averse grower
would make.

---

# Part VI. Uncertainty that survives the assumptions

# 14. Three kinds of interval, and which to quote

## 14.1 They answer different questions

| Tool | What the interval covers | What it rests on |
|---|---|---|
| `agri_np_predict(interval = "confidence")` | the fitted curve | the asymptotic theory of that engine |
| `agri_np_bootstrap()` | the fitted curve | the legitimacy of resampling experimental units |
| `agri_np_conformal()` | **a future plot** | exchangeability alone, in finite samples |

The first two describe how well the **average** response is known. Only the
third describes where an individual plot will fall, which is what a grower is
asking when a level is recommended.

```{r three-intervals}
nd2 <- data.frame(salinity = c(0, 3, 6),
                  block = factor("3", levels = levels(base$block)))

an <- as.data.frame(agri_np_predict(npfit, nd2, interval = "confidence"))
bo <- as.data.frame(agri_np_bootstrap(npfit, newdata = nd2, B = 199, seed = 2))
co <- as.data.frame(agri_np_conformal(npfit, newdata = nd2, level = 0.95,
                                      seed = 1))

data.frame(
  salinity  = nd2$salinity,
  analytic  = round(an$upper - an$lower, 3),
  bootstrap = round(bo$upper - bo$lower, 3),
  conformal = round(co$upper - co$lower, 3)
)
```

The conformal interval is several times wider, and correctly so. Quoting a
confidence band for the curve as if it described a plot understates the risk a
grower carries by a large factor. This is one of the commonest and most
consequential misreadings in applied agronomy.

## 14.2 The guarantee, and where the design enters

Split conformal prediction refits the engine on part of the data, measures
absolute residuals on the held-out part, and adds the appropriate empirical
quantile of those residuals. With the finite-sample correction the resulting
interval satisfies

$$P(Y_{\text{new}} \in \Gamma) \ge 1 - \alpha$$

in finite samples, for any engine and any response distribution.

The condition is **exchangeability**, and that is exactly where a declared
design stops being decoration. Plots inside a block were randomized and are
exchangeable. Plots in different blocks are not, because asserting that they
differ is what declaring a block means.

```{r conformal-scopes}
cw <- agri_np_conformal(npfit, newdata = base, level = 0.90, seed = 1,
                        scope = "within_block")
cn <- agri_np_conformal(npfit, newdata = base, level = 0.90, seed = 1,
                        scope = "new_block")

data.frame(
  question = c("a plot in a block we observed",
               "a plot in a field we did not"),
  mean_width = round(c(mean(cw$upper - cw$lower),
                       mean(cn$upper - cn$lower)), 3)
)
```

Predicting into a **new** block is a stronger claim, so the interval is wider: it
now carries between-block variation. It also requires dropping the block-specific
term, because a block effect is not estimable in a block that was never observed.

## 14.3 Holding the method to its promise

```{r coverage}
cv <- agri_np_coverage(cw, data = base)
data.frame(target = cv$target, empirical = round(cv$empirical, 3),
           mean_width = round(cv$mean_width, 3), n = cv$n)
cv$by_block
```

Coverage measured on the fitting data is optimistic, since those rows helped
build the interval. Its role here is teaching and diagnosis.

Note also that the guarantee is **marginal**, averaged over the gradient. It does
not promise the stated coverage separately at every salinity.

## 14.4 The model-based route to a new field, and the assumption-free one

There are two ways to say something about a field that was never observed, and
they rest on different things. They should be reported together.

```{r shrunk}
f_shrunk <- agri_np_regression(biomass ~ salinity, base, method = "gam",
                               block = block, block_effect = "shrunk")
agri_np_block_effects(f_shrunk)
```

A block that looks extreme is pulled back towards the common mean, because part
of its apparent difference is noise. The amount of travel is the amount of
between-block variation the data attribute to noise, and it is read off the
table rather than assumed.

```{r shrunk-plot, fig.cap = "How far each block is pulled towards the common mean. Filled: one free effect per block. Open: penalized, partly borrowed from the other blocks."}
plot(agri_np_block_effects(f_shrunk))
```

The shrunk block term is the **model-based** route to a new field, at the price
of a working assumption about how blocks vary. The conformal interval with
`scope = "new_block"` is the **assumption-free** route. When the two disagree,
the assumption is doing work the data do not support, and the wider of the two is
the one to quote.

---

# 15. Model checking without a distribution

## 15.1 Why not a normal QQ-plot

A normal QQ-plot asks whether residuals look Gaussian. For a package built to
avoid assuming a distribution, that is the wrong question asked of the wrong
quantity.

Simulation-based quantile residuals ask a better one: given the fitted model,
where does each observation fall inside its own predictive distribution? Under a
correct model those positions are uniform on the unit interval, whatever the
response distribution is.

```{r simdiag}
sd_fit <- agri_np_simdiag(npfit, nsim = 200, seed = 1)
sd_fit
```

```{r simdiag-plots, fig.cap = "Left question: are the scaled residuals uniform? Right question: is the fit equally good along the whole gradient?"}
plot(sd_fit, type = "uniform_qq")
plot(sd_fit, type = "residual_predictor")
```

## 15.2 The check that actually has power

The three checks do not have equal power, and it is worth demonstrating rather
than asserting. Below, the same data are fitted well by a block-adjusted GAM and
badly by a straight line, which cannot follow a curved decline.

```{r simdiag-power}
fit_line <- agri_np_regression(biomass ~ salinity, base, method = "theil_sen")

good <- agri_np_simdiag(npfit, nsim = 300, seed = 1)$checks
bad  <- agri_np_simdiag(fit_line, nsim = 300, seed = 1)$checks

data.frame(check = good$check,
           p_gam = round(good$p_value, 4),
           p_straight_line = round(bad$p_value, 4))
```

The overall uniformity check pools every observation, so a fit that is too low at
the ends and too high in the middle still produces residuals that are uniform on
average. The **location check along the gradient** is the row that moves, because
it compares residual positions across bins of the predictor and therefore detects
a non-monotone pattern that a rank correlation would miss entirely.

The practical lesson: read the location row, not the headline uniformity row.

## 15.3 These are descriptions, not decision rules

Nothing here selects a method. A departure tells you that the fitted mean or the
dispersion does not describe the data, which sends you back to the science, not
to a different p-value.

## 15.4 A model can pass every check and still be useless

Diagnostics answer "is the fit wrong". The explained-variation indices of
section 10.4 and the SiZer statement of section 11 answer "is the fit worth
anything". A model can pass every diagnostic while explaining almost nothing,
and reporting only the first is a common and consequential omission.

---

# Part VII. Treatments and responses that are not continuous

# 16. When the treatment can only be a whole number

## 16.1 The estimand, not the printing

Some agronomic treatments cannot take fractional values: plants per hill,
irrigation events, sprays per season, traps per hectare. The usual practice is
to fit a continuous curve, find its maximum, and round. That is wrong at the
level of the estimand rather than the presentation: the rounded value is not the
best admissible decision, it is the neighbour of an inadmissible one.

`agriRank` evaluates the fitted response **directly on the admissible integer
support** and defines the optimum as the best of those values.

```{r integer-fit}
data(agri_density)
fi <- agri_np_regression(yield ~ plants, agri_density, method = "integer_grid",
                         integer_base_method = "smoothing_spline",
                         predictor_support = "observed_integer")
fi
```

```{r integer-predict}
agri_integer_predict(fi)
```

## 16.2 Fractional predictions are refused

```{r integer-refused, error = TRUE}
agri_np_predict(fi, data.frame(plants = 7.5))
```

Once integer support is declared, a prediction at 7.5 plants describes an
experiment that cannot be carried out. The function refuses rather than
interpolating.

## 16.3 The decision, its uncertainty, and its confidence set

```{r integer-optimum}
agri_integer_optimum(fi)
```

```{r integer-boot}
ib <- agri_integer_bootstrap(fi, B = 199, seed = 1)
ib
```

The bootstrap does not produce an interval here, because the quantity is
discrete. It produces a **probability mass over admissible decisions**, which is
the honest object: it says how often each whole number was best across resampled
experiments.

```{r integer-confset}
agri_integer_confset(ib, level = 0.90)
```

The confidence set contains only admissible decisions. A set of three plant
densities is a more useful answer to a grower than a single number with a
symmetric interval that includes densities that cannot be planted.

## 16.4 Differences between adjacent decisions

```{r integer-diff}
agri_integer_difference(fi, order = 1L)
```

On a discrete support the derivative is not an admissible quantity. What replaces
it is the finite difference between adjacent admissible values: what is gained by
planting one more. That is also the quantity an economic threshold is applied to.

```{r sizer-integer, error = TRUE, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_sizer(fi)
```

## 16.5 Interpretation

For an integer treatment, report the optimum, the probability mass over
decisions, and the confidence set. Do not report a derivative, and do not report
a rounded continuous optimum.

---

# 17. Repeated measurements

## 17.1 The design changes, so the analysis changes

SPAD was measured four times on the same plants. Repeated measurements on one
plant are not exchangeable with measurements on different plants, and the
subject has to be declared.

```{r spad-sim}
set.seed(1502)
times <- c(20, 35, 50, 65)

spad <- do.call(rbind, lapply(times, function(tt) {
  z <- base[, c("block", "cultivar", "salinity", "salinity_f")]
  z$subject <- factor(paste(z$block, z$cultivar, z$salinity, sep = "_"))
  z$time <- factor(tt, levels = times)
  z$spad <- 42 +
    c(C1 = 0, C2 = 2.4, C3 = 1.1)[as.character(z$cultivar)] -
    0.9 * z$salinity -
    0.05 * (tt - 20) -
    0.004 * z$salinity * (tt - 20) +
    rnorm(nrow(z), 0, 1.3)
  z
}))
str(spad)
```

```{r spad-design}
des_r <- agri_design(spad ~ cultivar * time, spad, design = "repeated",
                     subject = subject, within = time)
des_r
```

```{r spad-fit}
fit_r <- agri_rank(des_r)
fit_r$omnibus
```

## 17.2 Reading a repeated-measures rank analysis

The three effects answer:

| Effect | Question |
|---|---|
| `cultivar` | averaged over time, do cultivars differ in SPAD |
| `time` | averaged over cultivars, does SPAD change through the cycle |
| `cultivar:time` | do the cultivars follow different trajectories |

As in Part II, the interaction governs the reading of the main effects.

```{r spad-plot, fig.cap = "SPAD trajectories. Each line is one cultivar averaged over blocks and salinity levels."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  print(
    ggplot2::ggplot(spad, ggplot2::aes(x = time, y = spad,
                                       colour = cultivar,
                                       group = cultivar)) +
      ggplot2::stat_summary(fun = mean, geom = "line", linewidth = 0.9) +
      ggplot2::stat_summary(fun = mean, geom = "point", size = 2) +
      ggplot2::labs(x = "Days after sowing", y = "SPAD", colour = "Cultivar") +
      agri_theme()
  )
}
```

## 17.3 Interpretation

A repeated-measures analysis answers questions about **trajectories**, not about
final values. If the scientific question concerns the endpoint, the endpoint
should be analysed as a single response in its own design, which is what Part II
did for biomass.

---

# 18. Missing repeated observations

## 18.1 Describe the missingness before modelling it

```{r missing-sim}
set.seed(1503)
spad_miss <- spad
drop <- sample(which(spad_miss$time != times[1]), 30)
spad_miss$spad[drop] <- NA
```

```{r missing-report}
des_m <- agri_design(spad ~ cultivar * time, spad_miss, design = "repeated",
                     subject = subject, within = time)
agri_missing_report(des_m)
```

The report answers three separate questions: how much is missing, where it is
concentrated, and whether the pattern is related to the design. Those are not the
same question, and a single overall percentage answers none of them well.

## 18.2 Sensitivity, not imputation

```{r missing-sens}
ms <- agri_missing_sensitivity(des_m, B = 199, seed = 1)
ms
```

This is a **sensitivity analysis**, not an imputation procedure. It asks whether
the conclusion changes under different assumptions about why the values are
missing. If it does, the honest report says so; no amount of imputation
machinery converts an unverifiable assumption into evidence.

## 18.3 The experimental engine, and its label

```{r wild, eval = FALSE}
# Not run in the vignette: this engine is experimental and its calibration
# study is not complete. The label stays until it is.
fit_w <- agri_repeated(des_m, backend = "native_wild", B = 1999,
                       missing_assumption = "MCAR")
fit_w$omnibus
```

The native incomplete repeated-measures engine implements rank-based
wild-bootstrap inference. It is labelled **experimental** in the package
documentation and it will keep that label until the validation plan in
`VALIDATION_PLAN.md` has been completed. Using an unlabelled experimental engine
in a manuscript is a risk that belongs to the analyst, and the package makes the
label impossible to miss.

## 18.4 Interpretation

Report the missingness pattern, the sensitivity analysis, and the assumption
under which the primary analysis was run. A missing-data assumption is part of
the methods, not a technical detail.

---

# 19. When the datum is not a measurement at all

## 19.1 Germination is counted inside intervals

A germination trial produces counts inside intervals, not measurements. A seed
that germinated between two inspections is known only to have done so somewhere
inside that interval, and a seed that never germinates is not a missing value: it
is an observation, censored at the end of the trial.

```{r tte, eval = requireNamespace("drcte", quietly = TRUE)}
data(verbascum, package = "drcte")
tte <- agri_np_timetoevent(nSeeds ~ timeBef + timeAf, verbascum,
                           by = Species, units = Dish, B = 99, seed = 1)
tte$summary[, c("level", "subjects", "responded", "t50_responders", "t50_lot")]
```

Two properties of a seed lot are routinely collapsed into one number and should
not be. **Capacity** is `responded`, the share that germinates at all.
**Speed** is the quantiles among those that do.

The whole-lot median is `NA` for a lot that never reaches half. That `NA` is the
result: such a lot has no median germination time, and reporting one would
require inventing a germination date for seeds that never germinated.

```{r tte-plot, eval = requireNamespace("drcte", quietly = TRUE), fig.cap = "Estimated time-to-event distributions. The step function is the estimate itself, not a smooth curve fitted through cumulative percentages."}
plot(tte, type = "cdf")
```

## 19.2 On-farm trials produce an order

```{r rankings}
r <- agri_rankings(biomass ~ cultivar, base, block = block)
r$summary
r$completeness
```

Every rank-based test in this package already works on within-block ranks.
`agri_rankings()` shows them. Here the design is complete, so rank sums are
comparable and the Conover machinery of Part III applies.

In a tricot trial, where each farmer ranks three varieties out of many, the
design is incomplete and rank sums are **not** comparable, because a variety
allocated to favourable farms collects flattering ranks for a reason that has
nothing to do with the variety. The function detects that, withholds the rank
sums, and keeps the pairwise record, which is made inside blocks.

```{r rankings-plot, fig.cap = "Share of blocks in which the row item was placed above the column item. Each cell uses only the blocks containing both."}
plot(r, type = "pairwise")
```

## 19.3 Interpretation

Both modules follow the same rule as the rest of the package: report the quantity
the data contain, and decline the one they do not.

---

# Part VIII. Sensitivity and communication

# 20. Does the conclusion depend on one analytical choice?

## 20.1 Sensitivity across admissible engines

```{r sensitivity}
sens <- agri_sensitivity(fit, methods = c("primary", "permuco"))
sens$table
sens$interpretation
```

Every analysis contains choices that are defensible but not unique. A sensitivity
analysis runs the admissible alternatives and reports whether the conclusion
survives them.

Read it as follows. If the alternatives agree, the conclusion does not rest on
the choice, and that is worth one sentence in the results. If they disagree, the
disagreement **is** the finding, and reporting only the convenient one is
misconduct rather than simplification.

## 20.2 Several responses at once

```{r batch}
bt <- agri_batch(des, responses = "biomass")
bt$summary
```

`agri_batch()` repeats one declared design across several responses without
changing the right-hand side. That constancy is the point: the temptation in a
multi-response study is to adjust the model per response until each one is
significant, and the batch interface makes that impossible by construction.

## 20.3 Interpretation

Sensitivity analysis is not a robustness ritual. It is the part of the workflow
that distinguishes a conclusion about the crop from a conclusion about the
software.

---

# 21. Publication-ready tables

## 21.1 From a fitted object to a table

```{r table-omnibus}
agri_table(fit, what = "omnibus", format = "data.frame")
```

```{r table-effects}
agri_table(fit, what = "effects", format = "data.frame")
```

`format = "auto"` returns a `gt` object when that package is installed, which is
what a manuscript wants. `format = "data.frame"` is used here so the vignette
prints plainly.

Every table this package produces carries the uncertainty of the estimates it
reports. A table of point estimates without intervals is not a result; it is a
summary of one realisation of the experiment.

## 21.2 Regression tables

```{r table-reg}
agri_np_diagnostics(npfit, cv = TRUE, seed = 1)$metrics
```

## 21.3 Table principles

1. report the estimate and its interval in the same cell or in adjacent columns,
   never on different pages;
2. state the stratum in which the comparison was made;
3. state the number of resampling replicates and the seed;
4. do not report a compact letter display without the effect sizes beside it;
5. do not report a p-value to more digits than the resampling supports: with
   `B = 999` the smallest attainable value is `2/1000`.

---

# 22. Publication-ready figures

## 22.1 The observed data

```{r fig-raw, fig.cap = "Observed biomass by cultivar and salinity."}
agri_plot(fit, type = "interaction")
```

## 22.2 The fitted response

```{r fig-fit, fig.cap = "Fitted salinity response with observed plots."}
agri_np_plot(npfit, points = TRUE, x_unit = "dS/m", y_unit = "g")
```

## 22.3 Colour that survives the reader

```{r fig-grey, fig.cap = "One curve per cultivar in grey tones, safe for black-and-white print."}
agri_np_plot(fit_var, group = "cultivar", palette = "grey",
             x_unit = "dS/m", y_unit = "g")
```

Group colours default to a colour-blind-safe palette. Grey tones are available
for journals that print in black and white, and the choice is an argument rather
than a manual override, so it survives a rebuild.

## 22.4 Export at journal widths

```{r fig-export, eval = FALSE}
p <- agri_np_plot(npfit, points = TRUE)

# One column, TIFF with LZW compression, as most agronomy journals require.
agri_save_figure(p, "figure_2.tiff", width = "one_column", dpi = 600)

# The same figure as an editable vector for a thesis.
agri_save_figure(p, "figure_2.pdf", width = "one_column")
```

`agri_save_figure()` writes TIFF, PDF, SVG, EPS or PNG at preset journal widths
and keeps text and lines editable in the vector formats. A figure exported as a
bitmap at screen resolution cannot be repaired later.

## 22.5 Figure principles

1. show the observed data under the fitted curve, always;
2. say in the caption what the band covers: the curve, or a plot;
3. do not draw a continuous line through an integer support;
4. label the axis with its unit;
5. use one figure to make one point.

---

# 23. A reproducible report

```{r report, eval = FALSE}
agri_report(fit, "salinity_case_study.md")
agri_dashboard(fit)
export_results(fit, "salinity_case_study.rds")
```

`agri_report()` writes a methods-and-results skeleton from the fitted object
itself, so the reported design, engine, replication and seeds cannot drift from
what was actually run. `export_results()` stores the object for a later session
or for a reviewer.

---

# Part IX. Two integrated workflows

# 24. Basic workflow: a factorial in blocks

The minimal defensible sequence, in ten steps.

### Step 1: data and design

```{r wf1-1}
d1 <- base
des1 <- agri_design(biomass ~ cultivar * salinity_f, d1, design = "rcbd",
                    block = block)
```

### Step 2: validate the declaration

```{r wf1-2}
validate_agri_design(des1)
```

### Step 3: omnibus inference

```{r wf1-3}
f1 <- agri_rank(des1)
f1$omnibus
```

### Step 4: read the interaction first

```{r wf1-4}
ip <- f1$omnibus[grep(":", f1$omnibus$effect), ]
ip[, c("effect", "Pr(>F)")]
cat(if (ip[["Pr(>F)"]] < 0.10)
      "Interaction detected: report the cultivar effect within salinity levels.\n"
    else
      paste("Interaction not detected. Main effects may be read marginally,",
            "but absence of evidence is not evidence of parallelism.\n"))
```

### Step 5: effect sizes on the response scale

```{r wf1-5}
head(as.data.frame(agri_effects(f1)), 6)
```

### Step 6: comparisons with compact letters

```{r wf1-6}
agri_cld(agri_conover(f1))
```

### Step 7: sensitivity across admissible engines

```{r wf1-7}
agri_sensitivity(f1, methods = c("primary", "permuco"))$table
```

### Step 8: the design figure

```{r wf1-8, fig.cap = "Interaction figure for the factorial analysis."}
agri_plot(f1, type = "interaction")
```

### Step 9: the manuscript table

```{r wf1-9}
agri_table(f1, what = "omnibus", format = "data.frame")
```

### Step 10: report

```{r wf1-10, eval = FALSE}
agri_report(f1, "basic_workflow.md")
```

### Basic interpretation template

> Biomass was analysed by design-aware rank-based inference in a randomized
> complete block design with five blocks (`agri_design`, `agri_rank`). The
> cultivar-by-salinity interaction was [significant / not significant]
> (p = [value]), so cultivar effects are reported [within each salinity level /
> marginally]. Relative marginal effects with 95% intervals are given in Table 1.
> Pairwise comparisons used the design-aware Conover procedure with [method]
> multiplicity control, and compact letters were computed within [stratum].
> Conclusions were unchanged under [alternative engines], reported in the
> supplementary material.

---

# 25. Advanced workflow: a quantitative gradient to a recommendation

The sequence that turns a fitted curve into something a grower can act on.

The gradient-designed nitrogen experiment is used here, because Stage E requires
more than four levels.

### Stage A: fit the gradient in the declared design

```{r wf2-a}
gA <- agri_np_regression(yield ~ dose, agri_dose, method = "gam", block = block)
```

### Stage B: compare engines predictively, and do not select on p

```{r wf2-b}
agri_np_compare(yield ~ dose, agri_dose, methods = c("gam", "kernel"),
                block = block, kfold = 5, seed = 1)
```

### Stage C: explained variation, all three indices

```{r wf2-c}
agri_np_diagnostics(gA, cv = TRUE, seed = 1)$r2
```

### Stage D: check the fit without a distribution

```{r wf2-d}
agri_np_simdiag(gA, nsim = 200, seed = 1)$checks
```

### Stage E: where is the response still changing

```{r wf2-e, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_significant_slope(agri_np_sizer(gA), stability = 0.8)
```

### Stage F: the location of the extreme, with its interval

```{r wf2-f}
agri_np_optimum_test(gA, B = 199, seed = 1, n = 80, external = FALSE)$optimum
```

### Stage G: an interval for the next plot

```{r wf2-g}
cfA <- agri_np_conformal(gA, newdata = agri_dose, level = 0.90, seed = 1)
agri_np_coverage(cfA, data = agri_dose)$empirical
```

### Stage H: the poor plot as well as the typical one

```{r wf2-h, eval = requireNamespace("qgam", quietly = TRUE)}
suppressWarnings(
  agri_np_quantile_curves(yield ~ dose, agri_dose, block = block,
                          quantiles = c(0.25, 0.5, 0.75), n = 40))$summary
```

### Advanced interpretation template

> The salinity response was modelled without assuming a functional form
> (`agri_np_regression`, GAM engine, block retained as an adjustment term).
> Predictive comparison across admissible engines is reported in Table S1; the
> engine was not selected on the basis of any p-value. Explained variation was
> [pseudo_r2] on the fitted values and [cv_r2] out of fold. Simulation-based
> quantile residuals showed [no departure / a departure] in the location check
> along the gradient (p = [value]).
>
> Biomass declined significantly up to [value] dS/m, with no evidence of further
> change beyond it, at agreement across 80% of bandwidths (`agri_np_sizer`). The
> location of the fitted extreme [was / was not] identified
> (p_boundary = [value]); [where identified, its 95% interval was [a, b]].
>
> A split-conformal interval covering a **future plot** at [level] had mean width
> [value] within an observed block and [value] for an unobserved field. Coverage
> was verified at [value].
>
> All resampling used B = [value] replicates and seed [value].

---

# Part X. Common mistakes, and the function that prevents each

# 26. Dropping the block because the test is simpler

**The mistake.** Running a Kruskal-Wallis test on data from a blocked design,
because the block "was not of interest".

**Why it is wrong.** The block was removed from the treatment comparison by the
randomization. Discarding it in the analysis puts the between-block variation
back into the residual, which inflates it and reduces power. The test is valid
for an experiment that was not performed.

**What prevents it.** `agri_design()` records the block, and `agri_rank()`
refuses an engine that cannot carry it. See section 5.3.

---

# 27. Treating a quantitative treatment as a factor without saying so

**The mistake.** Analysing four salinity levels as four unrelated categories, and
reporting only compact letters.

**Why it is wrong.** The shape of the response is the agronomic content. Letters
say which of the four tested levels differ, and nothing about what happens
between them, which is where a recommendation lives.

**What prevents it.** `agri_np_regression()` and Part IV. Reporting both
analyses is the answer, not choosing one.

---

# 28. Selecting the engine by the p-value it produces

**The mistake.** Fitting several smoothers and reporting the one whose treatment
effect reached significance.

**Why it is wrong.** The reported p-value is then the minimum of several, and its
distribution under the null is no longer the one being quoted.

**What prevents it.** `agri_np_compare()` reports cross-validated predictive
error only, and nothing in it feeds back into the inferential engine. See
section 10.2.

---

# 29. Reporting the fitted extreme as a recommendation

**The mistake.** Reading the maximum or minimum of a fitted curve as the
recommended level, without checking whether it is interior.

**Why it is wrong.** An extreme has to land somewhere. If the response flattens
rather than turning over, it lands on the boundary of the tested range, and the
boundary is a property of the experimental design, not of the crop.

**What prevents it.** `agri_np_optimum_test()` reports `p_boundary` and
`identified`, and `agri_np_significant_slope()` provides the defensible
alternative statement. See sections 11 and 12.

---

# 30. Comparing optima across curves the model forced to be parallel

**The mistake.** Fitting `y ~ x + cultivar` and comparing the optimum of each
cultivar.

**Why it is wrong.** An additive adjustment shifts curves without changing their
shape. Parallel curves share one optimum by construction, so the comparison
describes the model, not the experiment.

**What prevents it.** `agri_np_optimum_test()` detects parallel fitted curves and
refuses, pointing to `gam_structure = "varying"`. See section 12.3.

---

# 31. Quoting a confidence band as if it covered a plot

**The mistake.** Reporting the band around a fitted curve as the range in which
a grower's yield will fall.

**Why it is wrong.** A confidence band covers the average response. An individual
plot varies far more, typically by a factor of several.

**What prevents it.** `agri_np_conformal()` produces the interval that covers a
future plot, with a finite-sample guarantee. The three widths are contrasted in
section 14.1.

---

# 32. Rounding a continuous optimum to a whole number

**The mistake.** Fitting a continuous curve over plant density, finding the
maximum at 7.4 plants, and recommending 7.

**Why it is wrong.** The rounded value is the neighbour of an inadmissible one,
not the best admissible decision. They can differ.

**What prevents it.** `predictor_support = "observed_integer"` and
`agri_integer_optimum()`, which evaluate the response on the admissible support.
See section 16.

---

# 33. Deleting the seeds that never germinated

**The mistake.** Removing rows with no event before analysis, or forcing the
germination curve up to 100%.

**Why it is wrong.** A seed that never germinates is an observation, not a
missing value. Deleting it overstates the capacity of the lot, sometimes
grossly.

**What prevents it.** `agri_np_timetoevent()` requires `end = Inf` for censored
subjects, warns when no row has it, and reports capacity separately from speed.
See section 19.1.

---

# 34. Summing ranks across an incomplete design

**The mistake.** Ranking varieties within farms in an on-farm trial where each
farmer saw only three of thirty, and comparing the rank sums.

**Why it is wrong.** A variety allocated to favourable farms collects flattering
ranks for a reason unrelated to the variety. The comparison measures allocation
as much as genetics.

**What prevents it.** `agri_rankings()` detects incompleteness, withholds the
rank sums, and keeps the pairwise record, which is made inside blocks. See
section 19.2.

---

# 35. Reading non-significance as equality

**The mistake.** "The cultivars did not differ" after a test that did not reach
the chosen level.

**Why it is wrong.** Failing to detect a difference is not evidence of its
absence, particularly with five blocks. The statement conflates a decision with a
finding.

**What prevents it.** `agri_effects()` reports the effect size with its interval,
so a reader can see whether an agronomically important difference remains
compatible with the data. See section 8.3.

---

# 36. Treating a diagnostic as a model-selection rule

**The mistake.** Fitting several models and keeping the one whose residual
diagnostics look best.

**Why it is wrong.** Diagnostics answer whether a fit describes the data. Used
repeatedly as a filter, they become a selection procedure whose properties are
unknown, and the final p-value no longer means what it says.

**What prevents it.** `agri_np_simdiag()` returns descriptions with the question
each answers, and the documentation states that nothing in it selects a method.
See section 15.3.

---

# 37. Using an experimental engine without its label

**The mistake.** Using the native incomplete repeated-measures engine in a
manuscript because it ran without error.

**Why it is wrong.** Its calibration study is not complete. Running without error
is not evidence of correct Type-I error.

**What prevents it.** The engine is labelled experimental in its own
documentation, in `VALIDATION_PLAN.md`, and in this vignette. See section 18.3.

---

# Part XI. Compact function-selection guide

# 38. Choose by the question, not by the name

| Your situation | Start here | Then |
|---|---|---|
| I have a field notebook and no analysis yet | `agri_design()` | `validate_agri_design()`, `design_summary()` |
| One factor, no blocks | `np_crd()` | `agri_effects()`, `agri_conover()` |
| One factor in blocks | `np_rcbd()` | `agri_conover()`, `agri_cld()` |
| Two or more crossed factors | `np_factorial()` | read the interaction first |
| Whole plots and subplots | `np_splitplot()` | `np_splitsplit()`, `np_stripplot()` |
| Repeated measurements on the same unit | `np_repeated()` | `agri_repeated()`, `agri_missing_report()` |
| Several responses on the same units | `agri_multivariate()` | `agri_table()` |
| Several sites or years | `agri_multienv()` | check the GxE term |
| A quantitative treatment, and I want the shape | `agri_np_regression()` | `agri_np_plot()`, `agri_np_diagnostics()` |
| I need to know where the response stops changing | `agri_np_sizer()` | `agri_np_significant_slope()` |
| I need to recommend a level, with an interval | `agri_np_optimum_test()` | check `p_boundary` |
| The treatment can only be a whole number | `agri_np_regression(predictor_support=)` | `agri_integer_optimum()`, `agri_integer_confset()` |
| I need an interval for the next plot | `agri_np_conformal()` | `agri_np_coverage()` |
| I care about the poor plots, not the average | `agri_np_quantile_curves()` | `plot(type = "spread")` |
| I want to know whether the model describes the data | `agri_np_simdiag()` | read the location row |
| My data are germination counts | `agri_np_timetoevent()` | report capacity and speed apart |
| My data are farmer rankings | `agri_rankings()` | check `completeness` first |
| I want to know if the conclusion is fragile | `agri_sensitivity()` | `agri_batch()` |
| I need a manuscript table | `agri_table()` | `agri_format_ci()` |
| I need a journal figure | `agri_np_plot()`, `agri_plot()` | `agri_theme()`, `agri_save_figure()` |
| I need a reproducible record | `agri_report()` | `export_results()` |

# 39. Choose by what you must not do

| Temptation | What to do instead |
|---|---|
| drop the block because the test is simpler | declare it and accept the refusal |
| pick the engine with the best p-value | `agri_np_compare()` for prediction only |
| round a continuous optimum to an integer | declare the integer support |
| quote the fitted extreme as a recommendation | `agri_np_optimum_test()`, read `p_boundary` |
| quote a confidence band as a plot interval | `agri_np_conformal()` |
| report letters without effect sizes | report both |
| delete the non-germinated seeds | code them as `end = Inf` |
| sum ranks in an incomplete design | report the pairwise record |
| call non-significance equality | report the interval |

---

# Part XII. Minimum reporting checklist

# 40. What a manuscript using this package should contain

**Design**

1. the design as declared, including the blocking factor and the experimental
   unit for each treatment;
2. the number of blocks, replicates and plots, and any missing cells;
3. the randomization actually used, if it differed from the intended one.

**Analysis**

4. the engine, named, and the reason it was admissible for this design;
5. that the engine was not selected on the basis of a response p-value;
6. the multiplicity control used for any set of comparisons;
7. the stratum in which comparisons and compact letters were computed;
8. the number of resampling replicates and the random seed.

**Results**

9. the omnibus result and the interaction, read before the main effects;
10. effect sizes with intervals, not only p-values or letters;
11. for a quantitative treatment: the fitted curve, the explained-variation
    indices, and the interval over which the response is still changing;
12. for a recommendation: the interval for the **location** of the optimum, and
    the value of `p_boundary`; if not identified, say so and report the SiZer
    statement instead;
13. for a plot-level recommendation: a conformal interval, with its scope stated;
14. for an integer treatment: the optimum, the probability mass over decisions,
    and the confidence set;
15. for germination data: capacity and speed reported separately, with the
    censored subjects included;
16. for ranking data: whether the design was complete.

**Assurance**

17. the model check, reported whatever its result;
18. the sensitivity analysis across admissible engines;
19. the missing-data assumption, if any values were missing;
20. the package version and `sessionInfo()`.

# 41. A note on what not to report

Do not report a treatment mean beside a rank-based analysis; the analysis did not
estimate one. Do not report a compact letter display as if it established
equality. Do not report an optimum without its uncertainty. Do not report an
R-squared-like index alone beside a flexible fit. Do not report a p-value to more
digits than the resampling supports.

---

# Part XIII. Appendices

# Appendix A. Exported function registry

## A.1 Design

| Function | Purpose |
|---|---|
| `agri_design()` | declare the randomization structure explicitly |
| `validate_agri_design()` | audit the declaration against the data |
| `design_summary()` | describe the declared design |
| `simulate_agri()` | generate teaching data for a named design |
| `agri_methods()` | list the available engines and what each assumes |

## A.2 Design-aware inference

| Function | Purpose |
|---|---|
| `agri_rank()` | omnibus inference for a declared design |
| `np_crd()` | one factor, completely randomized |
| `np_rcbd()` | one factor in complete blocks |
| `np_factorial()` | crossed factors |
| `np_splitplot()` | whole plots and subplots |
| `np_splitsplit()` | three randomization levels |
| `np_stripplot()` | two strips crossed within blocks |
| `np_repeated()` | repeated measurements on the same unit |
| `agri_repeated()` | repeated measures with backend selection |
| `incomplete_wild_rank_test()` | experimental wild-bootstrap engine |

## A.3 Effects, comparisons and contrasts

| Function | Purpose |
|---|---|
| `agri_effects()` | relative marginal effects with intervals |
| `agri_pairs()` | pairwise comparisons |
| `agri_conover()` | design-aware Conover comparisons |
| `agri_cld()` | compact letters, computed within strata |
| `agri_contrast()` | planned contrasts |
| `agri_format_ci()` | estimate and interval as reportable text |

## A.4 Regression over a gradient

| Function | Purpose |
|---|---|
| `agri_np_regression()` | fit a nonparametric or shape-aware curve |
| `agri_np_predict()` | predict, with an analytic interval where defined |
| `agri_np_plot()` | the standard regression figures |
| `agri_np_curves()` | overlay the curves of several engines |
| `agri_np_derivative()` | derivative of the fitted curve |
| `agri_np_optimum()` | the fitted extreme, as a point |
| `agri_np_bootstrap()` | resampling intervals and bands for the curve |
| `agri_np_compare()` | predictive comparison of engines |
| `agri_np_diagnostics()` | explained variation and predictive error |
| `agri_np_significance()` | bootstrap tests for kernel fits |
| `agri_np_specification()` | is a prespecified equation too restrictive |
| `agri_np_interactive()` | interactive layer, when plotly is present |
| `agri_np_forest()` | forest plot of coefficient intervals |
| `agri_np_levels()` | the response at each level of a factor |

## A.5 From a curve to a recommendation

| Function | Purpose |
|---|---|
| `agri_np_sizer()` | where the derivative is significant, across bandwidths |
| `agri_np_significant_slope()` | the same, as a reportable sentence |
| `agri_np_optimum_test()` | interval and tests for the **location** of an optimum |
| `agri_np_quantile_curves()` | conditional quantiles and their spread |
| `agri_np_block_effects()` | fixed and shrunk block effects side by side |

## A.6 Distribution-free uncertainty and checking

| Function | Purpose |
|---|---|
| `agri_np_conformal()` | interval covering a future plot, with a guarantee |
| `agri_np_coverage()` | empirical coverage of that interval |
| `agri_np_simdiag()` | simulation-based quantile residuals |

## A.7 Integer decision support

| Function | Purpose |
|---|---|
| `agri_integer_predict()` | prediction restricted to admissible integers |
| `agri_integer_optimum()` | best admissible whole-number decision |
| `agri_integer_difference()` | gain from one more unit |
| `agri_integer_threshold()` | economic or agronomic threshold rules |
| `agri_integer_efficiency()` | response per unit of input |
| `agri_integer_bootstrap()` | probability mass over admissible decisions |
| `agri_integer_confset()` | discrete confidence set |

## A.8 Missing data, several responses, several sites

| Function | Purpose |
|---|---|
| `agri_missing_report()` | describe the missingness pattern |
| `agri_missing_sensitivity()` | does the conclusion survive the assumption |
| `agri_multivariate()` | several responses on the same units |
| `agri_multienv()` | several sites or years, and GxE |
| `agri_sensitivity()` | the conclusion across admissible engines |
| `agri_batch()` | one design, several responses |

## A.9 Data that are not measurements

| Function | Purpose |
|---|---|
| `agri_np_timetoevent()` | interval-censored germination and emergence |
| `agri_rankings()` | within-block rankings and on-farm trials |

## A.10 Trend, covariance and power

| Function | Purpose |
|---|---|
| `agri_trend()` | trend across ordered levels |
| `agri_ancova()` | rank-based covariance adjustment |
| `agri_power()` | power for a declared design |

## A.11 Communication

| Function | Purpose |
|---|---|
| `agri_table()` | manuscript tables from a fitted object |
| `agri_plot()` | the standard design figures |
| `agri_interactive()` | interactive layer for design figures |
| `agri_theme()` | the shared journal theme |
| `agri_save_figure()` | export at preset journal widths |
| `agri_report()` | reproducible methods-and-results skeleton |
| `agri_dashboard()` | assembled view of one analysis |
| `export_results()` | store the fitted object |

## A.12 Example data

| Object | Content |
|---|---|
| `agri_dose` | nitrogen response in five complete blocks, with a plateau |
| `agri_density` | plant density on an integer support |
| `agri_surface` | two quantitative gradients, for a response surface |

---

# Appendix B. A single starter script

Everything a first analysis needs, in one block.

```{r starter, eval = FALSE}
library(agriRank)

# 1. Declare the randomization.
des <- agri_design(yield ~ treatment, data = mydata, design = "rcbd",
                   block = block)
validate_agri_design(des)

# 2. Omnibus inference in that structure.
fit <- agri_rank(des)
fit$omnibus

# 3. Effect sizes with intervals, then comparisons with letters.
agri_effects(fit)
agri_cld(agri_conover(fit))

# 4. If the treatment is a quantitative gradient, describe its shape.
np <- agri_np_regression(yield ~ dose, mydata, method = "gam", block = block)
agri_np_diagnostics(np, cv = TRUE, seed = 1)$r2
agri_np_plot(np, points = TRUE)

# 5. Say where the response is still changing, not where the fit peaks.
agri_np_significant_slope(agri_np_sizer(np), stability = 0.8)

# 6. If a level is to be recommended, give the interval for its location.
agri_np_optimum_test(np, B = 999, seed = 1)

# 7. If the recommendation concerns a plot, give a conformal interval.
cf <- agri_np_conformal(np, level = 0.90, seed = 1)
agri_np_coverage(cf, data = mydata)

# 8. Check the fit without assuming a distribution.
agri_np_simdiag(np, nsim = 500, seed = 1)

# 9. Ask whether the conclusion survives another admissible engine.
agri_sensitivity(fit, methods = c("primary", "permuco"))$table

# 10. Record everything.
agri_report(fit, "analysis.md")
export_results(fit, "analysis.rds")
sessionInfo()
```

---

# Appendix C. Where to go next

| If you want | Read |
|---|---|
| the design declaration in detail | *Design Foundations, CRD, and RCBD* |
| comparisons and compact letters in detail | *Effects, Conover, Contrasts, and Factorial Inference* |
| split-plot, strip-plot, trend, ANCOVA, power | *Hierarchical Plot Designs, Trends, ANCOVA, and Power* |
| repeated measures and missing data | *Repeated Measures and Missing Longitudinal Data* |
| several responses or sites | *Multivariate, Multi-Environment, Batch, and Sensitivity Workflows* |
| the regression engines | *Nonparametric and Shape-Aware Regression for Agronomic Gradients* |
| integer decisions | *Integer-Support Nonparametric Regression for Agronomy* |
| figures, tables and reports | *Graphics, Tables, Reports, and Reproducibility* |
| the theory and the pitfalls | *Methodological Foundations, State of the Art, and Common Mistakes* |
| conformal prediction and model checking | *Distribution-Free Uncertainty and Model Checking* |
| optima, quantiles and block structure | *Optima, Quantiles, and How the Block Enters the Model* |
| germination and ranking data | *Time-to-Event and Ranking Data* |

---

# Final perspective

The functions in this package are ordinary. What is not ordinary is the number
of places where they stop.

A design declaration that refuses an engine. A comparison of optima that refuses
parallel curves. A quantile that refuses a tail the replication cannot resolve. A
whole-lot median that returns `NA` because the lot never reaches half. A rank sum
withheld because the design is incomplete. A derivative refused on an integer
support.

Each of those refusals replaces a number that would have looked like an answer.
That is the entire design philosophy, and it is worth stating plainly: **the most
useful thing a statistical package can do for an agronomist is to be clear about
what the experiment cannot support.**

The remaining scientific limitation of this version is documented rather than
hidden. The permutation backend shows a calibration failure in split-split-plot
and strip-plot strata, reported in `VALIDATION_PLAN.md`, and the native
incomplete repeated-measures engine carries an experimental label until its
calibration study is complete. Neither is repaired by test coverage, and neither
should be used in a manuscript without reading those documents first.

```{r session, eval = FALSE}
sessionInfo()
```
