---
title: "Quantitative treatments, covariate adjustment and power"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantitative treatments, covariate adjustment and power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

Ordered doses, baseline covariates, and design-stage power require methods that preserve the scientific structure rather than converting every numerical treatment to unrelated categories.



## `agri_trend()`: three examples

```r
x <- simulate_agri("rcbd", seed = 91, n = 6)
x$dose <- rep(c(0, 50, 100, 150), times = 6)
d <- agri_design(yield ~ dose, x, design = "rcbd", block = block, quantitative = dose)
agri_trend(d, B = 499)
agri_trend(d, treatment = dose, B = 499, seed = 92)
agri_trend(d, treatment = dose, scores = c(`0`=0, `50`=1, `100`=3, `150`=6), B = 499)
```

## `agri_ancova()`: three examples

The current implementation is a Freedman-Lane permutation ANCOVA adapter through `permuco`. It must not be described as the 2026 resampling NANCOVA method of Thiel et al.; that method remains a future validated engine.

```r
if (requireNamespace("permuco", quietly = TRUE)) {
  z <- simulate_agri("crd", seed = 93)
  z$baseline <- rnorm(nrow(z), 10, 2)
  agri_ancova(yield ~ treatment, z, covariates = baseline, np = 499)
  agri_ancova(yield ~ treatment, z, covariates = baseline, np = 499, rank_response = FALSE)
  z$block <- factor(rep(1:6, length.out = nrow(z)))
  agri_ancova(yield ~ treatment, z, covariates = baseline, block = block, np = 499)
}
```

## `agri_power()`: three examples

```r
G <- function(i) simulate_agri("crd", seed = 3000 + i, n = 6)
A <- function(z) np_crd(yield ~ treatment, z)$omnibus$p_value[1]
agri_power(G, A, nsim = 50, alpha = .05, seed = 94)
agri_power(G, A, nsim = 50, alpha = .01, seed = 95)
sapply(c(4, 6, 8), function(n) {
  Gn <- function(i) simulate_agri("crd", seed = 4000 + i, n = n)
  agri_power(Gn, A, nsim = 30)$power
})
```

Simulation-based power estimates the power of the **actual analysis workflow**, which is preferable when closed-form formulas do not match the intended rank/permutation procedure.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.
