---
title: "Multivariate responses and multi-environment experiments"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multivariate responses and multi-environment experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

Agronomic experiments frequently collect several correlated responses and repeat genotype evaluations across environments. These settings require explicit attention to within-unit multivariate dependence and genotype-by-environment structure.



## `agri_multivariate()`: integrated result objects

`agri_multivariate()` now routes among three MANOVA.RM modes while returning the same `agri_multivariate_fit` interface. Multiple responses in wide form use `MANOVA.wide`; long-form multivariate vectors use `MANOVA`; multiple responses combined with a repeated within-subject factor use `multRM`.

```r
if (requireNamespace("MANOVA.RM", quietly = TRUE)) {
  set.seed(101)
  mv <- data.frame(trt = factor(rep(LETTERS[1:3], each = 12)))
  mv$biomass <- rgamma(36, 5, 1) + as.numeric(mv$trt)
  mv$height <- rnorm(36, 40 + 2 * as.numeric(mv$trt), 4)
  mv$spad <- rnorm(36, 35 + as.numeric(mv$trt), 3)

  fit_mv <- agri_multivariate(cbind(biomass, height, spad) ~ trt, mv, iter = 499)
  fit_mv
  agri_table(fit_mv)
  agri_report(fit_mv, tempfile(fileext = ".md"), format = "md")
}
```

An optional block can be added as an explicit adjustment factor. MANOVA.RM treats this as a model factor rather than a random block effect; that distinction is reported and should be respected in interpretation.

For multivariate repeated measures, `within=` is not metadata only. `agriRank` enforces the within-subject factor in the fitted MANOVA.RM formula after the between-subject factors. Thus, a call such as `cbind(y1, y2) ~ treatment` with `within = time` is canonicalized to the full treatment-by-time structure required by `multRM()`.

```r
if (requireNamespace("MANOVA.RM", quietly = TRUE)) {
  set.seed(103)
  mvr <- expand.grid(subject = factor(1:12), time = factor(c("T1", "T2")))
  mvr$treatment <- factor(ifelse(as.integer(mvr$subject) <= 6, "control", "treated"))
  b <- rnorm(12)
  mvr$biomass <- 8 + b[as.integer(mvr$subject)] +
    0.8 * (mvr$treatment == "treated") * (mvr$time == "T2") + rnorm(nrow(mvr), 0, 0.3)
  mvr$spad <- 35 + 0.5 * b[as.integer(mvr$subject)] +
    1.2 * (mvr$time == "T2") + rnorm(nrow(mvr), 0, 0.5)

  fit_mvr <- agri_multivariate(
    cbind(biomass, spad) ~ treatment,
    mvr, subject = subject, within = time, iter = 199, seed = 103
  )
  fit_mvr$formula
  summary(fit_mvr)
}
```

## `agri_multienv()`: environment is enforced

A multi-environment declaration may no longer omit environment silently. Direct use of `agri_design(..., design = "multienv")` requires the environment in the formula. The convenience wrapper can inject it. With the default `environment_interaction = TRUE`,

```r
yield ~ genotype
```

is expanded conceptually to

```r
yield ~ genotype * environment
```

when `environment` was not already supplied. When `environment_interaction = FALSE`, the additive environment model is routed through `permuco` because ARTool requires a factorial fixed-effect specification containing the relevant interactions.

```r
me <- simulate_agri("multienv", seed = 102, n = 5)
if (requireNamespace("ARTool", quietly = TRUE) || requireNamespace("permuco", quietly = TRUE)) {
  fit_auto <- agri_multienv(yield ~ genotype, me, environment = environment, block = block)
  fit_auto$design$formula
}
if (requireNamespace("ARTool", quietly = TRUE))
  agri_multienv(yield ~ genotype * environment, me, environment = environment, block = block, method = "ART")
if (requireNamespace("permuco", quietly = TRUE))
  agri_multienv(yield ~ genotype, me, environment = environment, block = block,
                environment_interaction = FALSE, method = "permuco", np = 199)
```

When block labels are reused across environments, block-aware engines construct environment-specific block units before fitting.

## Interpretation

A multivariate global test should not be followed automatically by a collection of unadjusted univariate tests. Likewise, a multi-environment analysis should distinguish genotype performance, environment effects, and G×E rather than rank genotypes after pooling environments without justification.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.
