---
title: "Distribution-Free Uncertainty and Model Checking for Agronomic Regression"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distribution-Free Uncertainty and Model Checking for Agronomic Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 150
)
```

**Uncertainty vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** where a response is really changing, what interval covers the next
plot, and whether a fit describes the data, all without assuming a distribution.

---

# 1. Why this vignette exists

A fitted curve always has a shape. That is a property of the fitting, not of the
crop, and three of the commonest errors in applied agronomy follow from
forgetting it.

The first is reading the maximum of a fitted curve as a recommendation, when the
response never turned over and the maximum landed on the boundary of the tested
range. The second is quoting a confidence band for the curve as if it described
a grower's plot, which understates the risk by a large factor. The third is
checking a model with a normal QQ-plot, in an analysis whose whole premise is to
avoid assuming a distribution.

Each has a distribution-free answer, and this vignette provides all three.

> **Report what the data contain. Where they contain nothing, say so, rather
> than reporting the number the fitting produced anyway.**

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. explain why the maximum of a fitted curve is not a recommendation;
2. read a SiZer map vertically, and say what a conclusion that survives every
   bandwidth is worth;
3. produce the reportable sentence about where a response is still changing;
4. distinguish three kinds of interval and say what each covers;
5. explain what exchangeability means in a blocked design, and why it fixes the
   conformal split;
6. choose between `within_block` and `new_block` scope, and justify the choice;
7. verify empirical coverage, and explain why the guarantee is marginal;
8. explain why a normal QQ-plot is the wrong check for this package;
9. read simulation-based quantile residuals, and know which of the three checks
   has power;
10. recognise that a model can pass every diagnostic and still be useless;
11. apply the whole sequence to a real experiment and accept a null result.

---

# 3. The uncertainty module in one map

| Function | Answers | Rests on |
|---|---|---|
| `agri_np_sizer()` | where is the derivative significant, at every bandwidth | a distribution-free classification |
| `agri_np_significant_slope()` | the same, as one sentence | agreement across bandwidths |
| `agri_np_predict(interval=)` | an interval for the **curve** | the engine's asymptotic theory |
| `agri_np_bootstrap()` | an interval for the **curve** | resampling experimental units |
| `agri_np_conformal()` | an interval for the **next plot** | exchangeability alone, finite sample |
| `agri_np_coverage()` | did the interval keep its promise | the observed data |
| `agri_np_simdiag()` | does the model describe the data | simulation, no distribution assumed |

## 3.1 The order to use them in

```{r module-order}
data.frame(
  step = 1:5,
  ask = c("does the fit describe the data",
          "how much does it explain",
          "where is the response still changing",
          "what covers the curve",
          "what covers the next plot"),
  fn = c("agri_np_simdiag()", "agri_np_diagnostics()", "agri_np_sizer()",
         "agri_np_bootstrap()", "agri_np_conformal()")
)
```

---

# 4. Scope and navigation

This vignette owns the **uncertainty and model-checking block** of the
regression module. It answers three questions that the fitting vignettes
deliberately leave open:

1. Where along the gradient is the response *actually* changing, in a way that
   does not depend on how much we smoothed?
2. What interval will contain the yield of the **next plot**, with a guarantee
   that survives without assuming a distribution?
3. Does the fitted model describe these data, checked without a normality
   assumption we have refused everywhere else?

It does not own model fitting, which belongs to
*Nonparametric and Shape-Aware Regression for Agronomic Gradients*, nor integer
decisions, which belong to *Integer-Support Nonparametric Regression*.

Every tool here is distribution free. That is not a stylistic preference: it is
the only way the answers remain valid for the skewed, bounded and
heteroscedastic responses that field experiments routinely produce.

```{r data}
data(agri_dose)
str(agri_dose)
```

`agri_dose` is a nitrogen response in five complete blocks, with a plateau built
in at 200 kg ha^-1^. The plateau matters throughout this vignette: it is the
feature that separates a defensible recommendation from an artefact.

```{r base-fit}
fit <- agri_np_regression(yield ~ dose, agri_dose, method = "gam", block = block)
fit
```

---

# 1. Where is the response still rising?

## 1.1 Why the fitted optimum is not enough

`agri_np_optimum()` reports the maximum of one particular smooth. That is a
narrow question, and its answer moves when the smoothing moves.

```{r optimum-trap}
agri_np_optimum(fit)
```

The optimum lands on the upper edge of the tested range, flagged by
`at_boundary`. Reading that as "apply the highest rate" would be an artefact:
the data plateau well before it, and a maximum has to land somewhere.

The agronomic question is different. It is not *where is the maximum*, it is
**from which rate on is there no longer evidence that yield still rises**.

## 1.2 SiZer: an answer that survives the smoothing choice

`agri_np_sizer()` classifies the sign of the derivative at every position of
the gradient and for a whole column of bandwidths. A conclusion that holds
across the column does not depend on the analyst's smoothing choice.

```{r sizer, eval = requireNamespace("SiZer", quietly = TRUE)}
sz <- agri_np_sizer(fit)
sz
```

The table reads directly: over the lower part of the range the slope is
significantly positive; beyond it there is no evidence of change. Bandwidths
are printed on the scale of the predictor, in kg ha^-1^, so they can be judged
agronomically rather than treated as a tuning artefact.

```{r sizer-map, eval = requireNamespace("SiZer", quietly = TRUE), fig.cap = "SiZer map. Each row is one amount of smoothing, each column one position on the nitrogen axis. The four states are: significantly increasing, significantly decreasing, not distinguishable from flat, and sparse, meaning too few observations in that window to decide."}
plot(sz, type = "map")
```

Read the map **vertically**. A position classified as increasing in every row
supports the claim whatever the smoothing; a position that is increasing in only
a few rows does not. The sparse cells are equally informative: they mark where
the design does not carry enough plots to answer the question at that resolution,
which is a statement about the experiment rather than about the crop.

```{r sizer-stability, eval = requireNamespace("SiZer", quietly = TRUE), fig.cap = "Share of bandwidths supporting each state. The reading is robust where one curve is near one."}
plot(sz, type = "stability")
```

## 1.3 The statement to put in a paper

```{r sizer-statement, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_significant_slope(sz, stability = 0.8)
```

This is the sentence the figure supports: yield increases up to the reported
rate and there is no evidence of further increase beyond it, at a threshold of
agreement across 80% of the bandwidths. Compare it with the boundary optimum
above. The two disagree, and the SiZer statement is the defensible one.

## 1.4 Curvature

Setting `derivative = 2` classifies the second derivative, which is where a
change in the *rate of return* lives.

```{r sizer-curvature, eval = requireNamespace("SiZer", quietly = TRUE)}
summary(agri_np_sizer(fit, derivative = 2))
```

Here the whole range is classified as flat, which says the data do not support
locating a change in curvature. That is the expected outcome with five blocks:
second derivatives are estimated far less precisely than first ones, so a
curvature claim needs considerably more replication than a slope claim. Reading
the flat row as "there is no curvature" would be a mistake; it means "these data
cannot resolve it".

## 1.5 What SiZer does not do

It describes a continuous gradient. For an integer decision the derivative is
not an admissible quantity, and the function says so instead of producing a
number:

```{r sizer-integer, error = TRUE, eval = requireNamespace("SiZer", quietly = TRUE)}
data(agri_density)
fi <- agri_np_regression(yield ~ plants, agri_density, method = "integer_grid",
                         integer_base_method = "smoothing_spline",
                         predictor_support = "observed_integer")
agri_np_sizer(fi)
```

---

# 2. An interval for the next plot

## 2.1 Three questions that are routinely confused

| Tool | What the interval covers | What it assumes |
|---|---|---|
| `agri_np_predict(interval = "confidence")` | the fitted curve | the asymptotic theory of that engine |
| `agri_np_bootstrap()` | the fitted curve | that resampling the experimental units is legitimate |
| `agri_np_conformal()` | **a future plot** | exchangeability only, finite sample |

The first two describe how well we know the *average* response. Only the third
describes where an individual plot will fall, which is what a grower is asking
when a rate is recommended.

```{r three-uncertainties}
nd <- data.frame(dose = c(80, 160, 240),
                 block = factor("B3", levels = levels(agri_dose$block)))

an <- as.data.frame(agri_np_predict(fit, nd, interval = "confidence"))
bo <- as.data.frame(agri_np_bootstrap(fit, newdata = nd, B = 199, seed = 2))
co <- as.data.frame(agri_np_conformal(fit, newdata = nd, level = 0.95, seed = 1))

data.frame(
  dose       = nd$dose,
  analytic   = round(an$upper - an$lower, 3),
  bootstrap  = round(bo$upper - bo$lower, 3),
  conformal  = round(co$upper - co$lower, 3)
)
```

The conformal interval is several times wider, and correctly so. Quoting a
confidence band for the curve as if it described a plot understates the risk a
farmer carries by a large factor.

`B = 199` keeps the vignette fast. Use `B >= 999` for anything reported.

## 2.2 The guarantee, and where the design enters

Split conformal prediction refits the engine on part of the data, measures
absolute residuals on the held-out part, and adds the appropriate empirical
quantile of those residuals to the prediction. With the finite-sample
correction

$$\hat q = \frac{\lceil (n_{\text{cal}} + 1)(1 - \alpha) \rceil}{n_{\text{cal}}}\text{-quantile of the calibration residuals},$$

the resulting interval satisfies $P(Y_{\text{new}} \in \Gamma) \ge 1 - \alpha$
in finite samples, for any engine and any response distribution.

The condition is **exchangeability**, and this is exactly where a declared
design stops being decoration. Plots inside a block were randomized and are
exchangeable. Plots in different blocks are not: asserting that they differ is
what declaring a block means.

`agri_np_conformal()` therefore splits by block, and offers two scopes for two
different scientific questions.

```{r conformal-scopes}
cw <- agri_np_conformal(fit, newdata = agri_dose, level = 0.90, seed = 1,
                        scope = "within_block")
cn <- agri_np_conformal(fit, newdata = agri_dose, level = 0.90, seed = 1,
                        scope = "new_block")

data.frame(
  scope = c("future plot in an observed block", "future plot in a new block"),
  mean_width = round(c(mean(cw$upper - cw$lower), mean(cn$upper - cn$lower)), 3)
)
```

Predicting into a **new** block is a stronger claim, so the interval is wider:
it now carries between-block variation. It also requires dropping the
block-specific term, because a block effect is not estimable in a block that was
never observed. The function handles that and records it.

```{r conformal-print}
cn
```

```{r conformal-plot, fig.cap = "Split-conformal interval for a future plot. The band covers an individual plot, not the average response."}
plot(cw)
```

## 2.3 Holding the method to its promise

A method that promises 90% coverage should be checked.

```{r coverage}
cv <- agri_np_coverage(cw, data = agri_dose)
data.frame(target = cv$target,
           empirical = round(cv$empirical, 3),
           mean_width = round(cv$mean_width, 3),
           n = cv$n)
cv$by_block
```

Coverage on the fitting data is optimistic, since those rows helped build the
interval. Its role here is teaching and diagnosis. An honest assessment needs
the simulation study under `inst/calibration`.

Note also that the guarantee is **marginal**, averaged over the gradient. It
does not promise 90% separately at every rate. `normalize = TRUE` redistributes
the width, wider where the response is noisier, without changing what is
guaranteed.

```{r conformal-normalized}
cnorm <- agri_np_conformal(fit, newdata = agri_dose, level = 0.90, seed = 1,
                           normalize = TRUE)
data.frame(
  version = c("constant width", "locally scaled"),
  min_width = round(c(min(cw$upper - cw$lower), min(cnorm$upper - cnorm$lower)), 3),
  max_width = round(c(max(cw$upper - cw$lower), max(cnorm$upper - cnorm$lower)), 3),
  coverage  = round(c(agri_np_coverage(cw, data = agri_dose)$empirical,
                      agri_np_coverage(cnorm, data = agri_dose)$empirical), 3)
)
```

---

# 3. Checking the model without assuming a distribution

## 3.1 Why not a normal QQ-plot

A normal QQ-plot asks whether residuals look Gaussian. For a package built to
avoid assuming a distribution, that is the wrong question asked of the wrong
quantity.

Quantile residuals ask a better one: given the fitted model, where does each
observation fall inside its own predictive distribution? Under a correct model
those positions are uniform on the unit interval, whatever the response
distribution is.

```{r simdiag}
sd_fit <- agri_np_simdiag(fit, nsim = 200, seed = 1)
sd_fit
```

`agriRank` supplies the simulations by resampling the fitted residuals, which
keeps the reference distribution empirical. When `DHARMa` is installed, its
scaling machinery is applied to those same simulations.

```{r simdiag-plots, fig.cap = "Left question: are the scaled residuals uniform? Right question: is the fit equally good along the whole gradient?"}
plot(sd_fit, type = "uniform_qq")
plot(sd_fit, type = "residual_predictor")
```

## 3.2 The check that actually has power

The three checks answer different questions, and they do not have the same
power. This is worth demonstrating rather than asserting. Below, the same data
are fitted well by a block-adjusted GAM and badly by a straight line, which
cannot follow a plateau.

```{r simdiag-power}
fit_line <- agri_np_regression(yield ~ dose, agri_dose, method = "theil_sen")

good <- agri_np_simdiag(fit, nsim = 300, seed = 1)$checks
bad  <- agri_np_simdiag(fit_line, nsim = 300, seed = 1)$checks

data.frame(check = good$check,
           p_gam = round(good$p_value, 4),
           p_straight_line = round(bad$p_value, 4))
```

The overall uniformity check does not separate them at all: both sit near one.
This is the honest limitation of a marginal uniformity test. It pools every
observation, so a fit that is too low at the ends and too high in the middle
still produces residuals that are uniform *on average*.

The **location check along the gradient** is the row that moves, by an order of
magnitude, and it is the only one that flags the straight line at the 10% level.
It compares residual positions across bins of the predictor, which detects the
non-monotone pattern that a rank correlation would miss entirely.

The practical lesson: read the location row, not the headline uniformity row.
And read it as a flag, not a verdict. A p-value near 0.08 on 40 observations is
a reason to look at the fitted curve against the data, which is exactly what
section 1 already showed: the response plateaus, and a line cannot plateau.

## 3.3 These are descriptions, not decision rules

Nothing here selects a method. A departure tells you the fitted mean or the
dispersion does not describe the data, which sends you back to the science, not
to a different p-value.

---

# 4. A real experiment

Simulated data make points cleanly. Real data are the test of whether the tools
survive contact with a field.

`agridat::lasrosas.corn` holds a precision-agriculture maize trial from
Argentina: nitrogen rates applied on a grid, with topographic classes.

```{r agridat, eval = requireNamespace("agridat", quietly = TRUE)}
corn <- agridat::lasrosas.corn
corn <- corn[corn$year == 2001, c("yield", "nitro", "topo", "rep")]
corn$rep <- factor(corn$rep)
str(corn)
table(corn$topo)
```

There are 1705 plots, the nitrogen rates are not a tidy sequence, and the
response is noisy. All three are normal for on-farm data.

```{r agridat-fit, eval = requireNamespace("agridat", quietly = TRUE)}
fit_corn <- agri_np_regression(yield ~ nitro, corn, method = "gam", block = rep)
agri_np_diagnostics(fit_corn, cv = TRUE, seed = 1)$r2
```

The explained variation is close to zero, and the out-of-fold value is closer to
zero still. With 1705 plots that is not a shortage of data. Nitrogen rate simply
does not order yield in this field: topography and soil do.

```{r agridat-sizer, eval = requireNamespace("agridat", quietly = TRUE) && requireNamespace("SiZer", quietly = TRUE)}
sz_corn <- agri_np_sizer(fit_corn)
agri_np_significant_slope(sz_corn, stability = 0.8)
```

Every field is `NA`. This is the function working correctly, not failing: there
is no position on the nitrogen axis where the slope is significantly positive
across 80% of the bandwidths, so there is no rate to report and none is
invented. Contrast this with any procedure that would have returned an optimum
regardless, and note that the fitted curve below still *looks* like it has
shape.

```{r agridat-curve, eval = requireNamespace("agridat", quietly = TRUE), fig.cap = "Fitted nitrogen response in a real maize field. The curve has visible shape; the SiZer analysis shows that none of it is supported."}
agri_np_plot(fit_corn, points = TRUE)
```

```{r agridat-sizer-map, eval = requireNamespace("agridat", quietly = TRUE) && requireNamespace("SiZer", quietly = TRUE), fig.cap = "SiZer map for a real maize nitrogen trial. Nothing is classified as increasing at any bandwidth."}
plot(sz_corn, type = "map")
```

The gap between the two figures is the reason this vignette exists. A fitted
curve always has a shape. Only the second figure says whether that shape is
information.

```{r agridat-conformal, eval = requireNamespace("agridat", quietly = TRUE)}
cf_corn <- agri_np_conformal(fit_corn, newdata = corn, level = 0.90, seed = 1)
cvc <- agri_np_coverage(cf_corn, data = corn)
data.frame(target = cvc$target,
           empirical = round(cvc$empirical, 3),
           mean_width = round(cvc$mean_width, 2))
```

```{r agridat-check, eval = requireNamespace("agridat", quietly = TRUE)}
agri_np_simdiag(fit_corn, nsim = 200, seed = 1)$checks
```

Two things are worth noticing in these last two blocks.

The conformal interval is very wide relative to the yields, and its empirical
coverage lands essentially on target. Both are correct. When a predictor carries
almost no information, an honest prediction interval for a future plot must be
close to the spread of the response itself, and the conformal machinery
delivers that without being told.

The diagnostics, meanwhile, are clean. That is not a contradiction. The model
describes the data adequately; the data simply have almost nothing to say about
nitrogen. **A model can pass every diagnostic and still be useless.** Diagnostics
answer "is the fit wrong", the explained-variation indices and SiZer answer "is
the fit worth anything". Reporting only the first is a common and consequential
omission.

---

# 5. Reporting checklist

For a manuscript reporting a fertilizer response, the defensible set is:

1. the fitted curve with observed points, from `agri_np_plot()`;
2. the **SiZer statement**, not the boundary optimum, for where the response
   stops rising;
3. a **conformal interval** when the recommendation concerns a plot, with the
   scope stated: observed block, or a new field;
4. the **location check** from `agri_np_simdiag()`, reported whatever its
   result;
5. the **explained variation**, `pseudo_r2` and `cv_r2` together, so that a
   well-behaved but uninformative fit is visible as such;
6. the resampling counts and seeds, so the numbers can be reproduced.

If the SiZer statement comes back empty, that is the result. Reporting "no rate
at which the response is significantly increasing" is a finding, and it is the
one the maize field above supports.

Every figure here is a `ggplot` and every table a data frame, so both can be
restyled with `agri_theme()` or exported with `agri_save_figure()` as described
in *Graphics, Tables, Reports, and Reproducibility*.

---

# Part VI. Common mistakes, and the function that prevents each

# 6A. Reporting the fitted extreme as a recommendation

**The mistake.** "Yield was maximised at 280 kg N/ha."

**Why it is wrong.** A maximum has to land somewhere. If the response plateaus
rather than turning over, it lands on the boundary of the tested range, which is
a property of the experimental design, not of the crop.

**What prevents it.** `agri_np_optimum()` reports `at_boundary`,
`agri_np_optimum_test()` reports `p_boundary`, and
`agri_np_significant_slope()` supplies the defensible alternative sentence. See
sections 1.1 and 1.3.

---

# 6B. Reading a SiZer map horizontally

**The mistake.** Looking along one row of the map and reporting what that
bandwidth says.

**Why it is wrong.** One row is one smoothing choice. The whole point of the map
is that a conclusion holding in every row does not depend on the analyst's
choice, and one holding in a single row does.

**What prevents it.** `agri_np_significant_slope(stability =)` requires
agreement across a stated share of bandwidths. See section 1.2.

---

# 6C. Ignoring the sparse cells

**The mistake.** Treating a grey cell in the map as "no effect".

**Why it is wrong.** Sparse means the design does not carry enough plots in that
window to decide at that resolution. That is a statement about the experiment,
not about the crop.

**What prevents it.** The four states are reported separately and named in the
figure caption.

---

# 6D. Quoting a confidence band as a plot interval

**The mistake.** "A grower at 160 kg N/ha can expect between 4.9 and 5.2 t/ha."

**Why it is wrong.** That band covers the average response. An individual plot
varies several times more, as the three-way comparison in section 2.1 shows.

**What prevents it.** `agri_np_conformal()`, which covers a future plot with a
finite-sample guarantee.

---

# 6E. Using the wrong conformal scope

**The mistake.** Reporting a `within_block` interval as applying to a new farm.

**Why it is wrong.** `within_block` covers a plot in a block that was observed.
Predicting into a new field is a stronger claim and requires the wider
`new_block` interval, which carries between-block variation.

**What prevents it.** The scope is printed with the object and recorded as an
attribute. See section 2.2.

---

# 6F. Treating marginal coverage as conditional

**The mistake.** "The interval covers 90% of plots at every nitrogen rate."

**Why it is wrong.** The guarantee is **marginal**, averaged over the gradient.
It does not promise the stated coverage separately at every rate.

**What prevents it.** The printed output says so, and `normalize = TRUE`
redistributes the width for those who want local behaviour, without changing what
is guaranteed. See section 2.3.

---

# 6G. Checking a distribution-free model with a normal QQ-plot

**The mistake.** Plotting residuals against normal quantiles after a
nonparametric fit.

**Why it is wrong.** It asks whether the residuals are Gaussian, which the
analysis never assumed and does not need.

**What prevents it.** `agri_np_simdiag()` asks the right question instead: where
does each observation fall inside its own predictive distribution. See section
3.1.

---

# 6H. Reading the headline uniformity row

**The mistake.** "The residuals were uniform (p = 0.99), so the model fits."

**Why it is wrong.** The uniformity check pools every observation, so a fit that
is too low at the ends and too high in the middle still produces residuals that
are uniform on average. Section 3.2 demonstrates that it fails to distinguish a
plateau-following fit from a straight line.

**What prevents it.** The location check along the gradient, which is reported
in the same table and does have power.

---

# 6I. Using a diagnostic to choose a model

**The mistake.** Fitting several engines and keeping the one whose residual
checks look best.

**Why it is wrong.** Repeated use as a filter turns a description into a
selection procedure with unknown properties, and the final p-value no longer
means what it says.

**What prevents it.** The documentation states that nothing in `agri_np_simdiag()`
selects a method. See section 3.3.

---

# 6J. Confusing "not wrong" with "useful"

**The mistake.** Reporting clean diagnostics as evidence that the model is
informative.

**Why it is wrong.** The maize example in section 4 passes every diagnostic while
explaining essentially nothing. Diagnostics answer "is the fit wrong"; the
explained-variation indices and the SiZer statement answer "is it worth
anything".

**What prevents it.** Reporting both, as the checklist in section 5 requires.

---

# Part VII. Compact selection guide

# 6K. Choose by the question

| Your question | Use |
|---|---|
| where is the response still changing | `agri_np_sizer()` then `agri_np_significant_slope()` |
| how uncertain is the fitted curve | `agri_np_predict(interval=)` or `agri_np_bootstrap()` |
| what will the next plot yield | `agri_np_conformal(scope = "within_block")` |
| what will a plot on a new farm yield | `agri_np_conformal(scope = "new_block")` |
| did the interval keep its promise | `agri_np_coverage()` |
| does the model describe the data | `agri_np_simdiag()`, read the location row |
| is the model worth anything | `agri_np_diagnostics(cv = TRUE)` |
| the treatment is a whole number | the integer vignette; derivatives are refused |

# 6L. Choose the interval by what you are claiming

| The claim is about | Interval |
|---|---|
| the average response at a rate | analytic or bootstrap, on the curve |
| the shape of the whole curve | a **simultaneous** bootstrap band |
| one future plot in a field like these | conformal, `within_block` |
| one future plot on a new farm | conformal, `new_block` |
| the spread of plots at a rate | quantile curves, in the optima vignette |

---

# Part VIII. Glossary

# 6M. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **bandwidth** | how much of the gradient a local estimate averages over |
| **SiZer map** | the sign of the derivative at every position and bandwidth |
| **stability** | the share of bandwidths agreeing on a classification |
| **sparse** | too few observations in a window to classify at that resolution |
| **exchangeability** | any reordering of the observations is equally likely |
| **split conformal** | fit on one part, calibrate residuals on another |
| **calibration set** | the held-out part used to compute the interval width |
| **marginal coverage** | averaged over the gradient, not guaranteed at each point |
| **scope** | whether the future plot is in an observed block or a new one |
| **quantile residual** | where an observation falls inside its own predictive distribution |
| **binned Kruskal-Wallis** | comparing residual positions across bins of the predictor |
| **cured fraction** | in a related sense, the mass a model leaves on "never" |

---

# Part IX. Where to go next

| If you now want | Read |
|---|---|
| the engines that produced the curve | *Nonparametric and Shape-Aware Regression* |
| a rate to recommend, with an interval on its location | *Optima, Quantiles, and How the Block Enters the Model* |
| whole-number treatments | *Integer-Support Nonparametric Regression* |
| the design the fit sits inside | *Design Foundations, CRD, and RCBD* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# References

Chaudhuri, P. and Marron, J. S. (1999). SiZer for exploration of structures in
curves. *Journal of the American Statistical Association*, 94(447), 807-823.
DOI: 10.1080/01621459.1999.10474186.

Dunn, P. K. and Smyth, G. K. (1996). Randomized quantile residuals. *Journal of
Computational and Graphical Statistics*, 5(3), 236-244.
DOI: 10.1080/10618600.1996.10474708.

Lei, J., G'Sell, M., Rinaldo, A., Tibshirani, R. J. and Wasserman, L. (2018).
Distribution-free predictive inference for regression. *Journal of the American
Statistical Association*, 113(523), 1094-1111.
DOI: 10.1080/01621459.2017.1307116.

Vovk, V., Gammerman, A. and Shafer, G. (2005). *Algorithmic Learning in a Random
World*. Springer.
