---
title: "Scientific graphics and interactive exploration"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scientific graphics and interactive exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

Static figures should expose observed data and uncertainty whenever possible. `agriRank` returns real `ggplot2` objects, which remain editable with the standard grammar of graphics.



## `agri_plot()`: three examples

```{r plot-data}
x <- simulate_agri("crd", seed = 121)
d <- agri_design(yield ~ treatment, x, design = "crd")
fit <- agri_rank(d)
agri_plot(d, type = "data")
```

```{r plot-effects}
agri_plot(fit, type = "effects")
```

```{r plot-interaction}
xf <- simulate_agri("factorial", seed = 122)
df <- agri_design(yield ~ A * B, xf, design = "factorial")
agri_plot(df, type = "interaction")
```

Because the output is a `ggplot`, it can be extended:

```{r plot-extended}
agri_plot(fit) + ggplot2::labs(y = "Yield (Mg ha-1)") + ggplot2::theme_bw()
```

## `agri_interactive()`: three examples

The chunk below builds the three widgets and reports their classes. The widgets
themselves are not embedded: each one carries several megabytes of JavaScript,
which would dominate the installed size of the package for no statistical gain.
Print any of them in an interactive session to explore the graphic.

```{r interactive-widgets}
if (requireNamespace("plotly", quietly = TRUE)) {
  w_data        <- agri_interactive(fit, "data")
  w_effects     <- agri_interactive(fit, "effects")
  w_interaction <- agri_interactive(df, "interaction")
  vapply(list(w_data, w_effects, w_interaction), function(w) class(w)[1], character(1))
}
```

## Figure design principles

Avoid isolated bars for continuous outcomes when individual observations can be displayed. Show the observed distribution, the inferential estimand, and uncertainty as distinct visual layers. Interactive Plotly graphics are intended for exploration and teaching; the publication-quality static object remains the primary scientific figure.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.
