---
title: "Optima, Quantiles, and How the Block Enters the Model"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optima, Quantiles, and How the Block Enters the Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 150
)
```

**Recommendation vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** turning a fitted curve into something a grower can act on. What rate,
for whom, with what interval, and how the declared block enters.

---

# 1. Why this vignette exists

A fitted curve is a description. A recommendation is a decision, and the gap
between them is where most agronomic advice goes wrong.

Three questions have to be answered before a curve becomes advice, and none of
them is answered by the curve itself:

| Question | The curve alone gives | What is needed |
|---|---|---|
| what rate | a point with no uncertainty | an interval on the **location** |
| for whom | the average plot | the poor plot as well |
| where | the blocks that were observed | a route to a field that was not |

> **A recommended rate is a location, and locations have intervals. If the
> location is not identified, say so rather than reporting the point.**

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. explain why resampling the argmax is harder than resampling the curve;
2. read `p_boundary` and `identified`, and act on them;
3. explain why comparing optima across additively adjusted levels is arithmetic
   nonsense rather than a null result;
4. fit one smooth per level with `gam_structure = "varying"`;
5. interpret a resampling p-value with the Davison and Hinkley correction;
6. explain what a conditional quantile curve answers that a mean curve cannot;
7. read `coverage`, `deviation` and `tracking`, and abandon a quantile the
   experiment cannot resolve;
8. read the spread between outer quantiles as a statement about risk;
9. distinguish a fixed block effect from a shrunk one, and say what each permits;
10. report the model-based and assumption-free routes to a new field together,
    and know what disagreement between them means.

---

# 3. The recommendation module in one map

| Function | Answers |
|---|---|
| `agri_np_optimum()` | the fitted extreme, as a bare point |
| `agri_np_optimum_test()` | its **location**, with an interval and a boundary check |
| `agri_np_quantile_curves()` | what the poor plots do, not only the average one |
| `agri_np_regression(method = "smooth_quantile")` | one quantile at a time |
| `agri_np_regression(block_effect = "shrunk")` | a block term that can speak about a new field |
| `agri_np_block_effects()` | how much of the between-block spread is noise |
| `agri_np_conformal(scope = "new_block")` | the assumption-free route to the same claim |

---

# 4. Scope and navigation

This vignette owns the **recommendation block** of the regression module. Three
questions belong here, and each one exposes a place where a fitted curve is
routinely over-read:

1. A recommended rate is a *location*, so what is its interval, and is there a
   rate to recommend at all?
2. A curve through the middle of the data describes the typical plot. What
   describes the plot a grower actually gets in a bad year?
3. A declared block can enter the model in two ways, and the choice decides
   whether the model can say anything about a field it never saw.

Fitting engines belong to *Nonparametric and Shape-Aware Regression for
Agronomic Gradients*. Distribution-free uncertainty for the curve and for a
future plot belongs to *Distribution-Free Uncertainty and Model Checking*, which
this vignette leans on repeatedly. Integer decisions belong to *Integer-Support
Nonparametric Regression*.

Everything here is nonparametric. No response function is assumed, no plateau
model is fitted, and no distribution is claimed for the response.

```{r data}
data(agri_dose)
fit <- agri_np_regression(yield ~ dose, agri_dose, method = "gam", block = block)
fit
```

---

# 1. A rate is a location, and locations have intervals

## 1.1 The point estimate is not the answer

```{r optimum-point}
agri_np_optimum(fit)
```

That is a number with no uncertainty attached, and it sits on the upper edge of
the tested range. Both facts matter, and neither is visible in the number
itself.

## 1.2 Resampling the location, not the curve

`agri_np_bootstrap()` resamples the *height* of the curve. `agri_np_optimum_test()`
resamples its *argmax*, which is a different and harder quantity: a curve can be
estimated precisely while the position of its maximum wanders widely, and a
plateau is exactly the shape that makes this happen.

```{r optimum-test}
ot <- agri_np_optimum_test(fit, B = 199, seed = 1, n = 120, external = FALSE)
ot
```

`p_boundary` is the share of resampled experiments whose optimum lands on an end
of the searched range. Here it is one: in every replicate the maximum is at the
boundary, so there is no interior optimum to report and `identified` is `FALSE`.

The function says so rather than printing the boundary value as a recommendation.
Compare that with the bare point estimate above, which offers 280 with no
qualification at all.

```{r optimum-dist, fig.cap = "The location of the optimum across resampled experiments. Mass piled against one end is the visual form of an unidentified optimum."}
plot(ot, type = "distribution")
```

The honest report in this situation is the SiZer statement, which describes the
interval over which the response is still rising rather than inventing a turning
point.

```{r sizer-companion, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_significant_slope(agri_np_sizer(fit), stability = 0.8)
```

`B = 199` keeps the vignette fast. Use `B >= 999` for anything reported.

## 1.3 An experiment where the optimum does exist

To see the machinery work, the response has to turn over inside the tested
range. Below, one cultivar does and the other does not.

```{r two-cultivars}
set.seed(7)
a <- agri_dose; a$cultivar <- "late"
b <- agri_dose; b$cultivar <- "early"
b$yield <- b$yield - 0.000045 * (b$dose - 160)^2 + rnorm(nrow(b), 0, 0.15)
d2 <- rbind(a, b)
d2$cultivar <- factor(d2$cultivar, levels = c("early", "late"))
```

## 1.4 Parallel curves cannot have different optima

The obvious model adjusts for cultivar additively. It is also the wrong one for
this question, and the reason is arithmetic rather than statistical.

```{r additive-refused, error = TRUE}
fit_add <- agri_np_regression(yield ~ dose + cultivar, d2, method = "gam",
                              block = block)
agri_np_optimum_test(fit_add, by = cultivar, B = 49)
```

An additive adjustment shifts one curve above the other without changing its
shape. Parallel curves have their maxima at the same place by construction, so a
comparison of optima would report a difference of exactly zero and a p-value of
one, describing the model rather than the experiment. The function refuses.

`gam_structure = "varying"` fits one smooth of the rate per cultivar, so the
shapes may differ:

```{r varying}
fit2 <- agri_np_regression(yield ~ dose + cultivar, d2, method = "gam",
                           block = block, gam_structure = "varying")
fit2$formula_used
```

```{r optimum-test-2}
ot2 <- agri_np_optimum_test(fit2, by = cultivar, B = 199, seed = 1, n = 120,
                            external = requireNamespace("npregfast", quietly = TRUE))
ot2
```

Read the table row by row. `early` has an interior optimum with a usable
interval. `late` does not, and `p_boundary` says so. The contrast between them is
decisive, and the printed note explains what that contrast may and may not be
called: the two cultivars behave differently, but one side of the difference is
a boundary, not a rate.

When `npregfast` is installed the print also carries an independent check. It
uses different machinery and resamples rows rather than blocks, so it is a second
opinion rather than the interval to quote; agreement is evidence and
disagreement is a reason to look closer.

```{r optimum-curve, fig.cap = "Fitted response per cultivar with the resampled interval for the location of the optimum. The horizontal bar is uncertainty along the rate axis, not along the yield axis."}
plot(ot2, type = "curve")
```

---

# 2. The typical plot and the poor plot

## 2.1 Why a mean curve is not enough

A treatment can raise the good plots without raising the poor ones. The mean
response rises either way, and a recommendation built on it will disappoint
exactly the growers whose fields resemble the poor plots.

Conditional quantiles ask the question directly. The median describes the
typical plot; the tenth percentile is the exposure curve, what a grower meets in
a bad year.

```{r single-quantiles}
q10 <- agri_np_regression(yield ~ dose, agri_dose, method = "smooth_quantile",
                          tau = 0.10, block = block)
q50 <- agri_np_regression(yield ~ dose, agri_dose, method = "smooth_quantile",
                          tau = 0.50, block = block)

nd <- data.frame(dose = c(80, 160, 240),
                 block = factor("B3", levels = levels(agri_dose$block)))
data.frame(
  dose = nd$dose,
  q10  = round(as.numeric(agri_np_predict(q10, nd)), 3),
  q50  = round(as.numeric(agri_np_predict(q50, nd)), 3),
  gap  = round(as.numeric(agri_np_predict(q50, nd)) -
               as.numeric(agri_np_predict(q10, nd)), 3)
)
```

Each curve is fitted by the pinball loss, which defines the quantile directly.
Nothing is assumed about the shape of the response distribution, only about the
smoothness of the curve.

## 2.2 The fan, and the replication it needs

```{r fan-data}
set.seed(11)
d <- do.call(rbind, lapply(1:12, function(bk) {
  z <- agri_dose[agri_dose$block == "B1", c("dose", "yield")]
  z$block <- factor(paste0("B", bk), levels = paste0("B", 1:12))
  z$yield <- z$yield + rnorm(nrow(z), 0, 0.10 + 0.006 * z$dose)
  z
}))
nrow(d)
```

Twelve blocks rather than five, deliberately. A tail quantile needs observations
in that tail, and a 40-plot trial does not have them.

```{r fan}
qc <- agri_np_quantile_curves(yield ~ dose, d, block = block, n = 60)
qc
```

Read `deviation` before anything else. It is the gap between the share of plots
below each fitted curve and the quantile that curve claims to be, measured on
the very data the curve was fitted to. A quantile that misses its own target
here is not a tail the experiment can resolve, and `tracking` flags it.

Crossings are reported for the same reason. The quantiles are fitted
independently, so a crossing is evidence that they are not separately identified
at that point, and repairing it silently would hide the evidence.

```{r fan-plot, fig.cap = "Smooth conditional quantiles. Curves that fan out mean the treatment changes the spread of yield, not only its level."}
plot(qc, type = "fan")
```

## 2.3 The quantity a grower is exposed to

```{r spread, fig.cap = "Distance between the outer quantiles along the gradient."}
plot(qc, type = "spread")
```

```{r spread-table}
s <- qc$spread
data.frame(
  dose = c(min(s$dose), s$dose[which.min(abs(s$dose - 140))], max(s$dose)),
  spread = round(c(s$spread[1], s$spread[which.min(abs(s$dose - 140))],
                   s$spread[nrow(s)]), 3)
)
```

The spread widens as the rate rises. That is the finding: this treatment raises
yield and raises variability together. An average gain bought with variability
is a different recommendation from an average gain that lifts every plot, and
only the quantile view distinguishes them.

## 2.4 How this relates to a prediction interval

Section 2 and the conformal interval of the previous vignette answer adjacent
questions and should not be confused.

| Tool | What it describes |
|---|---|
| `agri_np_quantile_curves()` | where the 10th, 50th, 90th percentile plot sits at each rate |
| `agri_np_conformal()` | an interval that covers one future plot, with a coverage guarantee |

The quantile curves describe the shape of the conditional distribution and carry
no finite-sample guarantee. The conformal interval carries the guarantee but
says nothing about the shape. Reported together they are complementary.

---

# 3. How the block enters the model

## 3.1 Two choices, one consequence

```{r block-formulas}
f_fixed  <- agri_np_regression(yield ~ dose, agri_dose, method = "gam",
                               block = block)
f_shrunk <- agri_np_regression(yield ~ dose, agri_dose, method = "gam",
                               block = block, block_effect = "shrunk")

data.frame(
  block_effect = c("fixed", "shrunk"),
  formula = c(deparse1(f_fixed$formula_used), deparse1(f_shrunk$formula_used))
)
```

Fixed estimates one free effect per block and assumes nothing about how blocks
relate to each other. That is the classical choice, and its limitation is that
those effects exist only for the blocks that were observed.

Shrunk replaces them with a penalized term. A block that looks extreme is pulled
back towards the common mean, because part of its apparent difference is noise.
This is what makes a prediction for an unobserved field possible at all.

The response curve is nonparametric under both. This argument concerns the
nuisance structure, not the shape of the response.

## 3.2 Making the shrinkage visible

```{r block-effects}
agri_np_block_effects(f_fixed)
```

Almost nothing moves. These blocks differ by more than noise, so the data
decline to borrow between them, which is the correct behaviour and not a failure
of the method.

Contrast an experiment where the blocks are noisy and their true differences
small:

```{r block-effects-noisy}
set.seed(3)
dn <- agri_dose
dn$yield <- dn$yield - 0.9 * (as.numeric(dn$block) - 3) * 0.35 +
  rnorm(nrow(dn), 0, 0.9)
fn <- agri_np_regression(yield ~ dose, dn, method = "gam", block = block)
agri_np_block_effects(fn)
```

```{r block-plot, fig.cap = "How far each block is pulled towards the common mean. Filled: one free effect per block. Open: penalized, partly borrowed from the other blocks."}
plot(agri_np_block_effects(fn))
```

The amount of travel is the amount of between-block variation the data attribute
to noise. It is a property of the experiment, read off the figure rather than
assumed.

## 3.3 Two routes to a new field, and they should be reported together

Predicting into a block that was never observed is the point where the choice
bites. There are two routes, and they rest on different things.

```{r new-block}
mb <- agri_np_conformal(f_shrunk, newdata = agri_dose, level = 0.90, seed = 1,
                        scope = "new_block")
wb <- agri_np_conformal(f_shrunk, newdata = agri_dose, level = 0.90, seed = 1,
                        scope = "within_block")

data.frame(
  question = c("a plot in a block we observed", "a plot in a field we did not"),
  mean_width = round(c(mean(wb$upper - wb$lower), mean(mb$upper - mb$lower)), 3)
)
```

The shrunk block term is the **model-based** route: it borrows across blocks and
so has something to say about a new one, at the price of a working assumption
about how blocks vary. The conformal interval with `scope = "new_block"` is the
**assumption-free** route: it holds out whole blocks and asks only for
exchangeability at the block level.

They should agree. When they do not, the assumption is doing work the data do
not support, and the wider of the two is the one to quote.

---

# 4. Reporting checklist

For a manuscript recommending a rate:

1. the fitted curve with observed points;
2. `agri_np_optimum_test()`, reporting the interval for the **location** and the
   value of `p_boundary`, never the bare point estimate;
3. if `identified` is `FALSE`, the `agri_np_significant_slope()` statement
   instead of a rate;
4. when comparing levels, confirmation that the model lets the curve shapes
   differ, since parallel curves share one optimum by construction;
5. the quantile fan, or at least the median and the low quantile, whenever the
   recommendation concerns risk rather than the average plot;
6. whether the block was fixed or shrunk, and for a prediction outside the
   observed blocks, both routes of section 3.3;
7. all resampling counts and seeds.

Every figure is a `ggplot` and every table a data frame, so both can be restyled
with `agri_theme()` or exported with `agri_save_figure()`.

---

# Part V. Common mistakes, and the function that prevents each

# 5A. Reporting the bare point estimate

**The mistake.** "The optimum nitrogen rate was 280 kg/ha."

**Why it is wrong.** It has no uncertainty attached, and in this experiment it
sits on the boundary in every resampled replicate.

**What prevents it.** `agri_np_optimum_test()` reports the interval on the
location, `p_boundary`, and `identified`. See section 1.

---

# 5B. Ignoring `p_boundary`

**The mistake.** Reading the interval and skipping the boundary column.

**Why it is wrong.** An interval of [280, 280] looks precise. It means the
extreme was pinned to the edge in every replicate, which is the opposite of
precision.

**What prevents it.** The printed output raises an explicit note when
`identified` is `FALSE`, and names the alternative.

---

# 5C. Comparing optima across parallel curves

**The mistake.** Fitting `y ~ x + cultivar` and comparing the optimum of each
cultivar.

**Why it is wrong.** An additive adjustment shifts curves without changing their
shape, so the optima are identical by construction. The comparison would report a
difference of exactly zero and a p-value of one, describing the model rather than
the experiment.

**What prevents it.** `agri_np_optimum_test()` detects parallel fitted curves and
refuses, pointing to `gam_structure = "varying"`. See section 1.4.

---

# 5D. Reporting a resampling p-value of zero

**The mistake.** "p < 0.001" from 199 bootstrap replicates.

**Why it is wrong.** With B replicates the smallest attainable two-sided
resampling p-value is 2/(B+1). Quoting anything smaller claims precision the
resampling cannot deliver.

**What prevents it.** The Davison and Hinkley correction is applied
automatically. See section 1.

---

# 5E. Recommending from the mean curve when risk is the question

**The mistake.** A single fitted curve, and a recommendation drawn from it.

**Why it is wrong.** A treatment can lift the good plots without lifting the poor
ones. The mean rises either way, and the recommendation disappoints exactly the
growers whose fields resemble the poor plots.

**What prevents it.** `agri_np_quantile_curves()`. See section 2.

---

# 5F. Reporting a tail quantile the experiment cannot resolve

**The mistake.** A 10th-percentile curve from 40 plots.

**Why it is wrong.** Four observations sit in that tail. The curve is
indicative at best.

**What prevents it.** `coverage`, `deviation` and `tracking` check each curve
against the share of plots actually below it, and an extreme quantile is refused
outright when the tail is too thin. See section 2.2.

---

# 5G. Repairing crossing quantile curves

**The mistake.** Reordering or monotonising crossed quantile curves before
plotting.

**Why it is wrong.** A crossing is evidence that the quantiles are not separately
identified at that point. Repairing it hides the evidence.

**What prevents it.** Crossings are counted and reported, never repaired.

---

# 5H. Choosing shrunk blocks to make a table tidier

**The mistake.** Setting `block_effect = "shrunk"` because five free block
effects clutter the output.

**Why it is wrong.** Shrinking imposes a working assumption about how blocks
vary. It is justified when a prediction for an unobserved field is needed, not
when the table is long.

**What prevents it.** `agri_np_block_effects()` makes the amount of shrinkage
visible, so the assumption is at least inspectable. See section 3.2.

---

# 5I. Reporting only one route to a new field

**The mistake.** Reporting the shrunk-block prediction and omitting the conformal
one, or the reverse.

**Why it is wrong.** They rest on different things. When they disagree, the
assumption is doing work the data do not support, and only reporting both reveals
it.

**What prevents it.** Section 3.3, and the checklist.

---

# Part VI. Compact selection guide

# 5J. Choose by the decision you are supporting

| The decision | Use |
|---|---|
| what rate to apply | `agri_np_optimum_test()`, read `identified` first |
| whether two cultivars need different rates | `gam_structure = "varying"` then `by =` |
| whether the gain is worth the risk | `agri_np_quantile_curves()`, read `spread` |
| what a grower gets in a bad year | the low quantile curve |
| what a plot on this farm will yield | `agri_np_conformal(scope = "within_block")` |
| what a plot on a new farm will yield | both routes of section 3.3 |
| how much of the block spread is noise | `agri_np_block_effects()` |

# 5K. Choose the block treatment by what you must predict

| You need to predict for | `block_effect` | Also report |
|---|---|---|
| the blocks you observed | `"fixed"` | nothing extra |
| a field you did not observe | `"shrunk"` | the conformal `new_block` interval too |
| nothing; the block is a nuisance | `"fixed"` | it is the assumption-free default |

---

# Part VII. Glossary

# 5L. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **argmax** | the position at which a function attains its maximum |
| **location of the optimum** | the rate, as opposed to the yield at that rate |
| **p_boundary** | share of resampled experiments whose extreme fell on an end of the range |
| **identified** | `FALSE` when that share reaches one half |
| **parallel curves** | curves differing only in height, from an additive adjustment |
| **varying smooth** | one smooth per level, allowing the shapes to differ |
| **resampling p-value** | a p-value computed from the resampling distribution |
| **Davison and Hinkley correction** | the added one that floors the p-value at 2/(B+1) |
| **pinball loss** | the loss function that defines a conditional quantile |
| **exposure curve** | the low quantile, describing a bad plot |
| **spread** | the distance between the outer quantiles at a given rate |
| **tracking** | whether a fitted quantile matches the share of plots below it |
| **shrinkage** | how far a penalized block effect is pulled towards the common mean |
| **crossing** | two quantile curves that swap order, indicating non-identification |

---

# Part VIII. Where to go next

| If you now want | Read |
|---|---|
| the engines that produced the curve | *Nonparametric and Shape-Aware Regression* |
| where the response is still changing | *Distribution-Free Uncertainty and Model Checking* |
| whole-number treatments | *Integer-Support Nonparametric Regression* |
| the design the fit sits inside | *Design Foundations, CRD, and RCBD* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# References

Davison, A. C. and Hinkley, D. V. (1997). *Bootstrap Methods and their
Application*. Cambridge University Press.

Fasiolo, M., Wood, S. N., Zaffran, M., Nedellec, R. and Goude, Y. (2021). Fast
calibrated additive quantile regression. *Journal of the American Statistical
Association*, 116(535), 1402-1412. DOI: 10.1080/01621459.2020.1725521.

Koenker, R. (2005). *Quantile Regression*. Cambridge University Press.

Roca-Pardinas, J., Ordonez, C. and Sperlich, S. (2017). npregfast, an R package
for nonparametric estimation and inference in life sciences. *Journal of
Statistical Software*.

Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R*, 2nd
edition. Chapman and Hall/CRC.
