---
title: "Time-to-Event and Ranking Data"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-to-Event and Ranking Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(agriRank)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 150
)
```

**Non-measurement vignette**
**Package:** `agriRank`
**Version targeted:** `0.14.0`
**Owns:** the two agronomic experiments whose datum is not a measurement.
Germination counted inside intervals, and on-farm trials returning an order.

---

# 1. Why this vignette exists

Two kinds of agronomic experiment routinely have their data forced into methods
built for measurements, and in both cases the result is a number the data do not
contain.

A germination trial produces **counts inside intervals**, with some subjects
never responding. The habitual analysis converts them to cumulative percentages
and fits a curve, which asserts that the event time was observed, that successive
cumulative values are independent, and that the lot eventually reaches 100%. All
three are false.

An on-farm trial produces an **order**. Nothing was weighed. The habitual
analysis assigns scores to the ranks and treats them as measurements, which
imports an interval scale the data never had.

> **Report the quantity the data contain. Where a lot never reaches half, it has
> no median; where a design is incomplete, rank sums are not comparable. Say so.**

---

# 2. Learning objectives

After working through this vignette, the reader should be able to:

1. recognise interval-censored data and explain why the intervals matter;
2. explain why a subject that never responds is an observation, not a gap;
3. list the three assumptions the cumulative-percentage approach makes, and why
   each fails;
4. separate the capacity of a seed lot from its speed, and report both;
5. explain why a whole-lot median can be `NA`, and why that is the answer;
6. choose the correct experimental unit for a germination permutation test;
7. detect and correct the omission of censored rows;
8. read within-block rankings as the analysis rather than as a summary;
9. explain why completeness decides which ranking methods are admissible;
10. read a pairwise record, including its `blocks` column;
11. place a Plackett-Luce worth estimate correctly, as a model-based companion.

---

# 3. The module in one map

| Function | Answers | Requires |
|---|---|---|
| `agri_np_timetoevent()` | when, and how many, responded | `count ~ start + end`, `end = Inf` for censored |
| `plot(type = "cdf")` | the estimated distribution, as steps | a fitted object |
| `plot(type = "capacity")` | how much of each lot responds at all | a fitted object |
| `agri_rankings()` | who was placed above whom, inside blocks | `block` |
| `plot(type = "items")` | mean rank per item | a rankings object |
| `plot(type = "pairwise")` | the dominance matrix | at least two items |

## 3.1 What replaces what

```{r replacement-map}
data.frame(
  habitual = c("curve through cumulative percentages",
               "t50 from that curve",
               "drop the ungerminated seeds",
               "scores assigned to ranks",
               "rank sums across an incomplete design"),
  correct = c("nonparametric MLE under interval censoring",
              "capacity and speed reported separately",
              "code them as end = Inf",
              "the ranks themselves",
              "the pairwise record, made inside blocks")
)
```

---

# 4. Scope and navigation

Every other vignette in this package analyses a **measurement**: a yield, a
height, a SPAD reading. This one covers the two agronomic experiments that do
not produce one.

A germination trial produces **counts inside intervals**. A tray is inspected on
day 3, day 5, day 7. A seed that germinated between two inspections is known
only to have done so somewhere inside that interval, and a seed that never
germinates is not a missing value: it is an observation, censored at the end of
the trial.

An on-farm or tricot trial produces an **order**. Each farmer receives three of
thirty varieties and returns a ranking. Nothing was weighed, so there is no
response to analyse; the order is the datum.

Both are routinely forced into methods built for measurements, and in both cases
the damage is the same kind: a quantity gets reported that the data do not
contain. This vignette shows what the honest analysis looks like, and what it
refuses to report.

Measured responses in declared designs belong to *Design Foundations* and
*Effects, Conover Comparisons and Factorials*. Measured responses over a
gradient belong to the regression vignettes.

---

# 1. Germination is counted, not measured

## 1.1 What the usual analysis assumes

The common route is to turn the counts into cumulative germination percentages
and fit a curve through them. That makes three assumptions at once, and all
three are false for this kind of data:

| Assumption | Why it fails |
|---|---|
| the germination time was observed | it is only known to lie inside an interval |
| successive percentages are independent | each cumulative value contains all the earlier ones |
| the lot eventually reaches 100% | many lots do not, and forcing the curve up invents germination for seeds that never germinated |

The nonparametric maximum likelihood estimator of the time-to-event
distribution avoids all three. It uses the intervals as intervals, assumes no
functional form for the curve, and leaves probability mass on "never".

```{r data, eval = requireNamespace("drcte", quietly = TRUE)}
data(verbascum, package = "drcte")
str(verbascum)
```

Three *Verbascum* species, 100 seeds each, inspected daily. `timeBef` and
`timeAf` bound the interval in which each seed germinated; seeds that never
germinated appear with `timeAf = Inf`. `Dish` is the Petri dish, which is the
experimental unit.

## 1.2 Capacity and speed are two properties, not one

```{r fit, eval = requireNamespace("drcte", quietly = TRUE)}
tte <- agri_np_timetoevent(nSeeds ~ timeBef + timeAf, verbascum,
                           by = Species, units = Dish, B = 199, seed = 1)
tte
```

Read the table in two halves.

**Capacity** is `responded`: the share of the lot that germinates at all.
*V. arcturus* reaches 0.32, *V. blattaria* 0.84, *V. creticum* 0.97. These are
three very different seed lots, and no measure of speed will tell you that.

**Speed** is the quantiles, and they are reported twice on purpose.
`t50_responders` is measured among the seeds that did germinate and always
exists. `t50_lot` is measured on the whole lot and is `NA` for *arcturus*,
because a lot in which only 32% of seeds germinate never reaches half. That
`NA` is the result. Reporting a median germination time there would require
inventing a germination date for seeds that never germinated.

A single "t50" therefore cannot summarise this experiment, and any analysis that
produces one for *arcturus* has assumed something the data deny.

```{r cdf, eval = requireNamespace("drcte", quietly = TRUE), fig.cap = "Estimated time-to-event distributions. The step function is the estimate itself, not a smooth curve fitted through cumulative percentages. A curve that stops below one is a lot that does not fully germinate."}
plot(tte, type = "cdf")
```

```{r capacity, eval = requireNamespace("drcte", quietly = TRUE), fig.cap = "Capacity alone. A lot that germinates slowly and a lot that never germinates are different agronomic problems and call for different decisions."}
plot(tte, type = "capacity")
```

## 1.3 Comparing curves, at the level the randomization happened

```{r test, eval = requireNamespace("drcte", quietly = TRUE)}
tte$test
```

The comparison is a permutation test on rank scores, so it assumes no
distribution for the germination times.

The `clustered_by` column matters more than the p-value. Seeds sharing a Petri
dish share its water, its temperature and its handling, so they are not
independent observations. Permuting individual seeds would treat 100 seeds as
100 independent replicates when the experiment has far fewer. Passing `units`
permutes whole dishes instead.

The function warns when `units` is omitted rather than quietly producing the
anticonservative answer:

```{r no-units, eval = requireNamespace("drcte", quietly = TRUE)}
res <- tryCatch(
  agri_np_timetoevent(nSeeds ~ timeBef + timeAf, verbascum, by = Species,
                      B = 49, seed = 1),
  warning = function(w) conditionMessage(w))
cat(res, "\n")
```

## 1.4 The censored rows are data, not gaps

The commonest data-entry error in this area is to omit the ungerminated seeds
entirely. The consequence is silent and severe: the lot is then treated as fully
germinable and its capacity is overstated.

```{r censoring, eval = requireNamespace("drcte", quietly = TRUE)}
dropped <- verbascum[is.finite(verbascum$timeAf), ]
res <- withCallingHandlers(
  agri_np_timetoevent(nSeeds ~ timeBef + timeAf, dropped, by = Species,
                      units = Dish, B = 49, seed = 1),
  warning = function(w) invokeRestart("muffleWarning"))
res$summary[, c("level", "subjects", "responded")]
```

Every species now appears to germinate completely, and *arcturus* has lost 68 of
its 100 seeds from the analysis. The function warns about this, because the
difference between "did not germinate" and "was not recorded" is the whole
result.

---

# 2. When the datum is an order

## 2.1 Ranks are not a summary, they are the analysis

Every rank-based test in this package already works on within-block ranks.
Friedman and the Conover comparisons convert the measured response into ranks
inside each block and never look at the measurements again.
`agri_rankings()` shows those ranks.

```{r blocked}
set.seed(5)
d <- expand.grid(variety = factor(paste0("V", 1:5)),
                 farm = factor(paste0("F", 1:8)))
d$yield <- 3 + c(V1 = 0, V2 = 0.6, V3 = 1.1, V4 = 0.3, V5 = 1.4)[d$variety] +
  as.numeric(d$farm) * 0.2 + rnorm(nrow(d), 0, 0.5)

r <- agri_rankings(yield ~ variety, d, block = farm)
r$summary
```

The design is complete, every variety appears on every farm, so the rank sums
are comparable and the usual machinery applies:

```{r blocked-test}
fit <- np_rcbd(yield ~ variety, d, block = farm)
fit$omnibus
```

```{r blocked-plot, fig.cap = "Mean rank within farm. Lower is better."}
plot(r, type = "items")
```

## 2.2 The pairwise record

```{r pairwise}
head(r$pairwise, 5)
```

Each row uses only the blocks that contained both items. In a complete design
that is every block; in the incomplete design of the next section it is not, and
that is precisely why this table survives incompleteness when rank sums do not.

```{r pairwise-plot, fig.cap = "Share of blocks in which the row item was placed above the column item."}
plot(r, type = "pairwise")
```

## 2.3 On-farm trials are incomplete, and that changes what is admissible

In a tricot trial each farmer receives three varieties out of many. The design is
incomplete by construction.

```{r tricot}
set.seed(9)
vars <- paste0("V", 1:9)
tri <- do.call(rbind, lapply(1:40, function(i) {
  pick <- sample(vars, 3)
  q <- c(V1 = 1, V2 = 2, V3 = 3, V4 = 1.5, V5 = 3.5,
         V6 = 2.2, V7 = 0.8, V8 = 2.8, V9 = 1.2)[pick] + rnorm(3, 0, 0.8)
  data.frame(farm = paste0("F", i), variety = pick, position = rank(-q),
             stringsAsFactors = FALSE)
}))
rt <- agri_rankings(position ~ variety, tri, block = farm, ranked = TRUE)
rt$completeness
```

120 orderings out of the 360 a complete design would have produced.

```{r tricot-summary}
rt$summary
```

`rank_sum` is withheld. This is the substantive point of the section, so it is
worth stating plainly: with an incomplete design, a variety that happened to be
allocated to favourable farms collects flattering ranks for a reason that has
nothing to do with the variety. Summing those ranks and comparing the totals
compares allocation as much as genetics. `mean_rank` carries the same caution,
which is why the printed output says so.

What does survive is the pairwise record, because every comparison in it was
made inside one farm:

```{r tricot-pairwise}
head(rt$pairwise, 6)
```

Note the `blocks` column. Some pairs were seen together on many farms and some
on one or two. A pair seen once carries almost no information, and the table
shows that instead of hiding it behind a p-value.

## 2.4 The model-based companion

`PlackettLuce` fits a likelihood model to rankings and returns one worth per
item, which is what allows incomplete rankings to be combined onto a single
scale. `agri_rankings()` reports it when the package is installed, in its own
block and labelled:

```{r worth}
if (is.null(rt$worth)) {
  cat("PlackettLuce is not installed in this environment.\n")
} else {
  print(rt$worth[order(-rt$worth$worth), ])
}
```

Two things are worth keeping straight.

It is **not** a distribution-free summary. It assumes the rankings arise from one
common worth per item, and that assumption is doing real work: it is what makes
the incomplete design analysable on a single scale at all. Where the worth
ordering and the pairwise record disagree, the assumption is the difference.

Everything in sections 2.1 to 2.3 is distribution free and needs no model. The
package therefore reports it first, and degrades to it when `PlackettLuce` is
absent.

---

# 3. Reporting checklist

For a germination or emergence trial:

1. the number of subjects and how many never responded, stated as data rather
   than dropped;
2. **capacity**, the share responding, before any measure of speed;
3. quantiles of the responders and of the whole lot, side by side, with the
   `NA` preserved where the lot never reaches that share;
4. the estimated distribution as a step function, not a smooth curve fitted
   through cumulative percentages;
5. the permutation comparison with the unit that was randomized, dish or tray,
   named.

For a ranking or on-farm trial:

6. whether the design is complete, before any rank sum is reported;
7. the pairwise record with its `blocks` column, so a comparison resting on two
   farms is visible as such;
8. if a Plackett-Luce worth is reported, its label as a model-based estimate and
   its agreement or disagreement with the pairwise record.

Every figure is a `ggplot` and every table a data frame, so both can be restyled
with `agri_theme()` or exported with `agri_save_figure()`.

---

# Part IV. Common mistakes, and the function that prevents each

# 4A. Fitting a curve to cumulative germination percentages

**The mistake.** Converting counts to cumulative percentages and fitting a
log-logistic or Weibull curve to them.

**Why it is wrong.** It asserts three things at once: that the germination time
was observed, that successive cumulative values are independent when each
contains all the earlier ones, and that the lot reaches 100%.

**What prevents it.** `agri_np_timetoevent()` uses the intervals as intervals,
assumes no functional form, and leaves mass on "never". Parametric germination
models are deliberately not offered.

---

# 4B. Deleting the seeds that never germinated

**The mistake.** Removing rows with no event before analysis.

**Why it is wrong.** The demonstration in section 1.4 is unambiguous: every
species then appears to germinate completely, and *arcturus* loses 68 of its 100
seeds from the analysis.

**What prevents it.** `end = Inf` is required for censored subjects, and the
function warns when no row carries it.

---

# 4C. Reporting a single t50

**The mistake.** "Median germination time was 11.5 days."

**Why it is wrong.** For a lot in which 32% of seeds germinate, there is no time
by which half the lot has germinated. The number quoted is the median among the
**responders**, which is a different quantity.

**What prevents it.** Quantiles are reported twice, `*_responders` and `*_lot`,
and the second is `NA` where the lot never reaches that share.

---

# 4D. Permuting seeds instead of dishes

**The mistake.** Comparing germination curves by permuting individual seeds.

**Why it is wrong.** Seeds sharing a Petri dish share its water, temperature and
handling. Permuting them treats 100 seeds as 100 independent replicates when the
experiment has far fewer, and the p-value is anticonservative.

**What prevents it.** `units =` names the dish, and the function warns when it
is omitted rather than returning the anticonservative answer silently.

---

# 4E. Confusing capacity with speed

**The mistake.** Ranking seed lots by t50 alone.

**Why it is wrong.** A lot that germinates quickly but only partially and a lot
that germinates slowly but completely are different agronomic problems, and a
single number cannot distinguish them.

**What prevents it.** `responded` is reported first, before any quantile, and
`plot(type = "capacity")` shows it alone.

---

# 4F. Assigning scores to ranks and treating them as measurements

**The mistake.** Scoring first place as 3, second as 2, third as 1, and running
an ANOVA.

**Why it is wrong.** It imports an interval scale the data never had: it asserts
that the gap between first and second equals the gap between second and third.

**What prevents it.** `agri_rankings()` works on the ranks themselves, and the
Conover machinery of the comparisons vignette does the same.

---

# 4G. Summing ranks across an incomplete design

**The mistake.** Comparing rank sums in a tricot trial.

**Why it is wrong.** A variety allocated to favourable farms collects flattering
ranks for a reason unrelated to the variety. The comparison measures allocation
as much as genetics.

**What prevents it.** `agri_rankings()` detects incompleteness, reports it, and
withholds `rank_sum`.

---

# 4H. Ignoring the `blocks` column of the pairwise record

**The mistake.** Reading a pairwise share of 1.00 as decisive.

**Why it is wrong.** A pair seen on one farm has a share of 0 or 1 by
construction, and carries almost no information.

**What prevents it.** The `blocks` column is reported beside every share, and
the sign-test p-value reflects the number of separating blocks.

---

# 4I. Reporting Plackett-Luce worth as a distribution-free summary

**The mistake.** Presenting worth estimates as the nonparametric result.

**Why it is wrong.** Plackett-Luce is a likelihood model for rankings. It assumes
one common worth per item, and that assumption is what allows incomplete
rankings to be combined onto a single scale.

**What prevents it.** The worth block is printed separately and labelled, and its
absence changes nothing else in the output.

---

# Part V. Compact selection guide

# 4J. Choose by what was recorded

| The datum is | Use |
|---|---|
| a measurement, in a declared design | the design vignettes |
| a count inside an interval, some never responding | `agri_np_timetoevent()` |
| a count with exact times | the same, with narrow intervals |
| an order, complete design | `agri_rankings()`, then `agri_conover()` |
| an order, incomplete design | `agri_rankings()`, pairwise record only |
| an order, and a single scale is needed | Plackett-Luce, labelled as model-based |

# 4K. Choose the germination summary by the question

| The question | Report |
|---|---|
| how good is this seed lot | `responded`, first |
| how fast does it germinate | `t50_responders` |
| by when will half the lot be up | `t50_lot`, and accept `NA` |
| do these lots differ | the permutation test, clustered by dish |
| how uniform is germination | the spread between `t10` and `t90` of responders |

---

# Part VI. Glossary

# 4L. Terms used in this vignette

| Term | Meaning here |
|---|---|
| **interval censoring** | the event is known to have occurred inside an interval |
| **right censoring** | the event had not occurred by the end of the trial |
| **cured fraction** | the share that never responds |
| **capacity** | the share of a lot that responds at all |
| **speed** | how quickly the responding part responds |
| **NPMLE** | nonparametric maximum likelihood estimator of the distribution |
| **Turnbull intervals** | the intervals on which the NPMLE places mass |
| **experimental unit** | the dish, tray or plot, not the individual seed |
| **complete design** | every item appears in every block |
| **incomplete design** | each block contains only a subset of the items |
| **pairwise record** | how often one item was placed above another, inside blocks |
| **worth** | a Plackett-Luce parameter placing items on one scale |

---

# Part VII. Where to go next

| If you now want | Read |
|---|---|
| measured responses in declared designs | *Design Foundations, CRD, and RCBD* |
| comparisons and compact letters | *Effects, Conover, Contrasts, and Factorial Inference* |
| plant death as a treatment effect in a longitudinal trial | *Repeated Measures and Missing Longitudinal Data* |
| figures and tables for these outputs | *Graphics, Tables, Reports, and Reproducibility* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# References

Onofri, A., Mesgaran, M. B. and Ritz, C. (2022). A unified framework for the
analysis of germination, emergence, and other time-to-event data in weed
science. *Weed Science*, 70(3), 259-271. DOI: 10.1017/wsc.2022.8.

Turnbull, B. W. (1976). The empirical distribution function with arbitrarily
grouped, censored and truncated data. *Journal of the Royal Statistical Society
Series B*, 38(3), 290-295.

Turner, H. L., van Etten, J., Firth, D. and Kosmidis, I. (2020). Modelling
rankings in R: the PlackettLuce package. *Computational Statistics*, 35,
1027-1057. DOI: 10.1007/s00180-020-00959-3.

van Etten, J., Beza, E., Calderer, L. et al. (2019). First experiences with a
novel farmer citizen science approach. *Experimental Agriculture*, 55(S1),
275-296. DOI: 10.1017/S0014479716000739.
