---
title: "From a Curve to a Decision"
author: "agriRank Development Team"
date: "2026-08-24"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a Curve to a Decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r agrirank-setup, include = FALSE}
library(agriRank)
set.seed(2026)
```

# 1. Why this vignette exists

A fitted curve is not a recommendation. Between the two lie four questions that
the earlier regression vignettes did not answer, and that this one does.

**What does the interval cover?** An interval around a curve covers the mean
response. A grower does not sow the mean, they sow one plot, and the interval
that covers one plot is wider. Reporting the first where the second was meant
understates the uncertainty, sometimes by a factor of three.

**How well does the model predict where nothing was measured?** Cross-validated
error can answer two different questions, and the flattering one is the default
almost everywhere. Whether a plot in an observed block is predicted well, and
whether an unobserved block is predicted well, are not the same, and only the
second is what a recommendation claims.

**What rate should be applied?** The top of the curve is the agronomic optimum.
It is almost never the rate to apply, because the last increments of input buy
less produce than they cost.

**What if the field itself has a gradient?** Blocking was invented for a field
whose fertility varies in patches the size of a block. It does nothing about a
slope, an old road, a drainage line, and those are the common case.

## 1.1 What else is here

| If you want | Read |
|---|---|
| the engines and how a curve is fitted | *Nonparametric and Shape-Aware Regression* |
| where the response is still changing | *Distribution-Free Uncertainty and Model Checking* |
| whether the model's assumptions survive | *Testing What a Model Assumes* |
| the design side, before any of this | *Design Foundations, CRD, and RCBD* |

# 2. Learning objectives

By the end you should be able to:

1. tell a confidence interval from a prediction interval, and say which one a
   recommendation needs;
2. choose between the two cross-validation scopes, and say which question each
   answers;
3. report an economic optimum with its interval, and read it as a function of
   the price ratio rather than as a single number;
4. find the joint optimum of two rates, and explain why the rectangle of two
   marginal intervals is not its confidence region;
5. add a field-position term and say what it absorbed;
6. recognise when a prediction leaves the support of the data.

# 3. The module map

| Question | Function | Added in |
|---|---|---|
| what covers the next plot | `agri_np_predict(interval = "prediction")` | 0.14.0 |
| is this request inside the data | `agri_np_predict(extrapolation = )` | 0.14.0 |
| which prediction error | `cv_scope` in `agri_np_compare()` and `agri_np_diagnostics()` | 0.14.0 |
| what rate pays | `agri_np_optimum_economic()` | 0.14.0 |
| what pair of rates pays | `agri_np_optimum_surface()` | 0.14.0 |
| what the field position hides | `spatial =` in `agri_np_regression()` | 0.14.0 |
| refit changing one thing | `update()` | 0.14.0 |
| into a report | `agri_tidy()`, `agri_glance()`, `agri_augment()` | 0.14.0 |

---

# Part I. What the interval covers

# 4. The trial

```{r trial}
d <- expand.grid(N = seq(0, 200, by = 50), block = factor(1:5), rep = 1:2)
d$yield <- 3 + 0.030 * d$N - 0.00009 * d$N^2 +
  as.numeric(d$block) * 0.30 + rnorm(nrow(d), 0, 0.30)
str(d)
```

Five blocks, five nitrogen rates, two plots per rate per block. The response
rises and turns over inside the tested range, which is the case where a
recommendation is possible at all.

```{r fit}
fit <- agri_np_regression(yield ~ N, d, method = "gam", block = block, k = 5)
fit
```

## 4.1 Two intervals, two questions

```{r intervals}
nd <- data.frame(N = c(50, 100, 150),
                 block = factor("1", levels = levels(d$block)))

ci <- agri_np_predict(fit, nd, interval = "confidence")
pi <- agri_np_predict(fit, nd, interval = "prediction",
                      scope = "within_block", seed = 1)

data.frame(N = nd$N,
           confidence_width = round(ci$upper - ci$lower, 3),
           prediction_width = round(pi$upper - pi$lower, 3))
```

## 4.2 Interpretation

The prediction interval is wider, and it must be. The confidence interval covers
the average yield of plots at that rate; the prediction interval covers the
yield of **one** plot, which carries the plot-to-plot variation on top of the
uncertainty about the average.

A recommendation is about a plot. If a methods section reports the narrow
interval and the discussion talks about what a grower will harvest, the two do
not refer to the same quantity.

The prediction interval comes from split conformal prediction, so its coverage
is a finite-sample guarantee rather than a consequence of an assumed error
distribution. Nothing about normality is used.

## 4.3 Which plot, in which block

```{r scope}
pw <- agri_np_predict(fit, nd, interval = "prediction",
                      scope = "within_block", seed = 1)
pn <- agri_np_predict(fit, nd, interval = "prediction",
                      scope = "new_block", seed = 1)

data.frame(N = nd$N,
           within_block = round(pw$upper - pw$lower, 3),
           new_block = round(pn$upper - pn$lower, 3))
```

`within_block` covers a new plot in a block already measured. `new_block` covers
a plot in a field, or a year, not yet seen. The second is the one a
recommendation to other growers implies, and it is the one nobody reports.

The argument is required, without a default, when the fit declares a block.
Guessing on the user's behalf would understate the interval half the time.

## 4.4 The error that this invites

```{r wrong-interval}
# Reported as if it covered a plot, this is too narrow by a factor of about
round((pi$upper - pi$lower) / (ci$upper - ci$lower), 2)
```

# 5. Asking beyond the data

```{r extrapolation, error = TRUE}
# The trial went to 200 kg/ha. What happens at 400?
agri_np_predict(fit, data.frame(N = 400,
                                block = factor("1", levels = levels(d$block))),
                extrapolation = "error")
```

## 5.1 Interpretation

Outside the range of the data a smoother has no information. Whatever number
comes back describes the spline basis, not the experiment. This is different in
kind from a parametric model, where extrapolation is at least an extrapolation
of a stated functional form; here there is no form to extrapolate.

The default is `"warn"`, which returns the value and flags the offending rows:

```{r extrapolation-warn}
p <- agri_np_predict(fit, data.frame(N = c(100, 400),
                                     block = factor("1", levels = levels(d$block))))
attr(p, "extrapolated") %||% p$extrapolated
```

Cross-validation and bootstrap loops are exempt from the check, because held-out
folds and resampled replicates leave the training range by construction.

---

# Part II. Which prediction error

# 6. Two scopes, two questions

```{r cv-scope}
w <- agri_np_diagnostics(fit, cv = TRUE, kfold = 5, seed = 1,
                         cv_scope = "within_block")
n <- agri_np_diagnostics(fit, cv = TRUE, kfold = 5, seed = 1,
                         cv_scope = "new_block")

data.frame(scope = c("within_block", "new_block"),
           cv_r2 = round(c(w$r2$cv_r2, n$r2$cv_r2), 4))
```

## 6.1 Interpretation

`within_block` stratifies the folds, so every block appears in training. Plot
mates therefore sit on both sides of the split, and the estimate answers "how
well is another plot in **this** block predicted". That is the optimistic
question.

`new_block` holds out whole blocks. It answers "how well is a block never
measured predicted", which is what a recommendation to a different field claims.
The number is always worse, and it is the honest one.

## 6.2 A substantive consequence

Under `new_block` the block term is **dropped** from the fold models. A block
that was held out has no estimated effect, so a model carrying it could only
return `NA` for every held-out row. Dropping it makes the validated error refer
to a block of average fertility, which is the only thing an unobserved block can
mean.

That is not a workaround. It is the statement that you cannot predict a specific
new block, only an average one, and any procedure that appears to do otherwise
is estimating something it has no data for.

## 6.3 The incoherence this fixed

Before 0.14.0 `agri_np_compare()` stratified its folds by block while the
routine behind `agri_np_diagnostics(cv = TRUE)` assigned rows at random. The
same model reported two different validated errors depending on which function
was asked, and nothing said so. Both now share one fold rule and one argument.

---

# Part III. What rate pays

# 7. The agronomic optimum is not the recommendation

```{r agronomic}
agri_np_optimum(fit)
```

The top of the curve. Applying it means paying for the last increments of
nitrogen that return less grain than they cost.

# 8. The economic optimum

```{r economic}
ec <- agri_np_optimum_economic(fit, price_ratio = 0.006, B = 99, seed = 1)
ec
```

## 8.1 What the price ratio is

The price of one unit of input divided by the price of one unit of produce, in
the units of the fitted model. If nitrogen costs 1.20 per kg and grain sells for
0.30 per kg, the ratio is 4, and the optimum is where the last kilogram of
nitrogen returns four kilograms of grain.

The function solves

$$\frac{\partial y}{\partial x} = r$$

and takes the root at the first crossing from above, because up to that point
another unit of input still pays for itself.

## 8.2 The check that the solver is right

```{r economic-zero}
agri_np_optimum_economic(fit, price_ratio = 0, B = 99, seed = 1)$optimum$optimum
agri_np_optimum(fit, n = 200)$optimum
```

With a price ratio of zero the economic optimum must be the agronomic one,
because the input is free. It is, to within the grid spacing. This is not a
recommendation, it is a check.

## 8.3 The ratio moves the answer more than the data do

```{r economic-sensitivity}
ev <- agri_np_optimum_economic(fit, price_ratio = c(0.002, 0.006, 0.012, 0.020),
                               B = 99, seed = 1)
ev$optimum[, c("price_ratio", "optimum", "lower", "upper", "identified")]
```

## 8.4 Interpretation

Read that table as a sensitivity analysis, not as four recommendations. The
price ratio is treated as known and it is not: prices move between the trial and
the season in which the recommendation is used, and they usually move the answer
further than the resampling interval does.

Compare the width of any single interval with the spread across ratios. If the
second is larger, and it usually is, then the honest limitation of the
recommendation is economic rather than statistical, and the methods section
should say which price ratio was assumed.

## 8.5 When there is no rate to report

`p_boundary` is the share of replicates whose root lands on an end of the
searched range, and `identified` turns `FALSE` at one half. At that point the
trial does not contain the answer: either the response never stops paying inside
the tested range, or it never starts. Reporting a rate anyway would be reporting
the edge of the design.

# 9. Two rates at once

```{r surface-data}
s <- expand.grid(N = seq(0, 200, 50), P = seq(0, 60, 15), block = factor(1:3))
s$yield <- 3 + 0.030 * s$N - 0.00013 * s$N^2 + 0.070 * s$P -
  0.0009 * s$P^2 - 0.00010 * s$N * s$P +
  as.numeric(s$block) * 0.2 + rnorm(nrow(s), 0, 0.15)
```

## 9.1 The model must let them interact

```{r surface-additive, error = TRUE}
fa <- agri_np_regression(yield ~ N + P, s, method = "gam", block = block, k = 4)
agri_np_optimum_surface(fa, B = 49, n = 12)
```

An additive surface has an optimum in each input that is the same at every level
of the other. A joint optimum would then carry no information the two separate
optima do not, and reporting a joint region for it would describe the model
rather than the trial. The function refuses.

```{r surface-tensor}
ft <- agri_np_regression(yield ~ N + P, s, method = "gam", block = block,
                         k = 4, gam_structure = "tensor")
os <- agri_np_optimum_surface(ft, B = 99, seed = 1, n = 20)
os
```

## 9.2 The rectangle is not the region

```{r surface-region}
head(os$region)
```

`box_lower` and `box_upper` are the two marginal intervals. Together they form a
rectangle, and **the rectangle is not the confidence region for the pair**. When
the surface has a ridge, more of one input compensating for less of the other,
the cloud of resampled optima lies along a diagonal and the rectangle admits
corners that no replicate ever visited.

The rank correlation between the two coordinates across replicates is printed
for that reason. A value far from zero is the signature of a ridge.

## 9.3 Interpretation

Report the pair, the region, and the correlation. A reader who sees only two
marginal intervals will imagine a rectangle, and will conclude that the two
rates can be chosen independently, which on a ridge they cannot.

---

# Part IV. What the field position hides

# 10. Blocking is a coarse instrument

```{r spatial-data}
f <- expand.grid(row = 1:8, col = 1:6)
f$N <- rep(seq(0, 200, by = 50), length.out = nrow(f))
f$block <- factor(ceiling(f$row / 2))
# A gradient that runs continuously across the whole trial, which no block can
# absorb: a slope, an old road, a drainage line.
f$yield <- 3 + 0.03 * f$N - 0.00009 * f$N^2 +
  0.25 * f$row + 0.15 * f$col + rnorm(nrow(f), 0, 0.20)
```

```{r spatial-fits}
f0 <- agri_np_regression(yield ~ N, f, method = "gam", block = block, k = 4)
fx <- agri_np_regression(yield ~ N, f, method = "gam", block = block, k = 4,
                         spatial = "smooth_xy", coords = c("row", "col"))
fr <- agri_np_regression(yield ~ N, f, method = "gam", block = block, k = 4,
                         spatial = "row_col", coords = c("row", "col"))

data.frame(model = c("block only", "block + s(row, col)", "block + row + col"),
           RMSE = round(c(f0$metrics$RMSE, fx$metrics$RMSE, fr$metrics$RMSE), 4))
```

## 10.1 What each does

```{r spatial-formulas}
deparse(f0$formula_used)
deparse(fx$formula_used)
deparse(fr$formula_used)
```

`smooth_xy` adds a two-dimensional thin-plate smooth of the coordinates, which
absorbs a continuous trend of any orientation and costs the effective degrees of
freedom that the fitting method selects for it. `row_col` adds additive row and
column factors, which is cheaper and right when the layout is a lattice and the
trend follows it.

## 10.2 Interpretation

The residual fell by more than half. That variation was in the trial all along,
the block could not absorb it because it crosses blocks, and it was inflating
every interval and every p-value computed from that residual.

The coordinates are nuisance terms, not predictors. Prediction grids hold them
at their reference value, so a reported curve refers to a plot of average
position, which is the right default.

## 10.3 The engines that can carry it

```{r spatial-refuse, error = TRUE}
agri_np_regression(yield ~ N, f, method = "loess",
                   spatial = "smooth_xy", coords = c("row", "col"))
```

Only the penalised additive engines have a term to hold a field trend. Asked of
another engine the request is refused, because dropping it silently would leave
the trend in the residual while the output suggested otherwise.

---

# Part V. Working with the objects

# 11. Changing one thing

```{r update}
f_k7 <- update(fx, k = 7)
c(original_k = fx$settings$k, updated_k = f_k7$settings$k)
c(spatial_preserved = f_k7$spatial)
```

Comparing two engines or trying a shape constraint no longer means retyping the
whole call, which is where a predictor or a block quietly goes missing between
the two versions being compared. The refit uses the rows stored in the fit, so
the two models are fitted to the same data.

All the guards still apply:

```{r update-guard, error = TRUE}
update(fit, method = "loess")
```

# 12. Into a report

```{r broom}
head(agri_tidy(fit, n = 8))
agri_glance(fit)
head(agri_augment(fit), 3)
```

## 12.1 Why `agri_tidy()` returns a curve

Most of the sixteen engines have no coefficients, and those that do have them
for a spline basis rather than for any quantity worth reporting. So the tidied
object is the **fitted curve**, one row per grid point.

There is no `p.value` column, because no test was performed. The rank-based side
does return one row per term with a p-value, because there one was.

# 13. Resampling faster

```{r parallel, eval = FALSE}
# Requires future.apply and a plan set by the user
future::plan(future::multisession, workers = 4)
agri_np_optimum_economic(fit, price_ratio = 0.006, B = 999, seed = 1,
                         parallel = TRUE)
future::plan(future::sequential)
```

The answer does not depend on it. Each replicate is drawn from its own
L'Ecuyer-CMRG substream, so replicate `b` is the same object whichever worker
computes it and in whatever order, and a run on four cores returns the same
interval as a run on one. Below a few hundred replicates the workers cost more
than they save, which is why the default is sequential.

---

# Part VI. Common mistakes, and the function that prevents each

# 14. Reporting a confidence interval as if it covered a plot

**The mistake.** Quoting `interval = "confidence"` in a sentence about what a
grower will harvest.

**Why it is wrong.** The two cover different quantities and differ by a factor
of two or three. The narrower one is about the average of many plots.

**What prevents it.** `interval = "prediction"` exists and is one argument away.

# 15. Reporting the flattering cross-validation

**The mistake.** Quoting a validated error obtained with stratified folds as
evidence that the model predicts new fields well.

**Why it is wrong.** Plot mates sat on both sides of the split, so the estimate
describes prediction inside an observed block.

**What prevents it.** `cv_scope = "new_block"`, and the note printed with the
result names which question was answered.

# 16. Recommending the top of the curve

**The mistake.** Reporting `agri_np_optimum()` as the rate to apply.

**Why it is wrong.** It ignores the price of the input entirely. On a
decelerating response the economic optimum is always lower, often by a third.

**What prevents it.** `agri_np_optimum_economic()`, and the printed comparison
against the agronomic optimum that comes with it.

# 17. Treating the price ratio as known

**The mistake.** Reporting one economic optimum without saying what ratio was
assumed.

**Why it is wrong.** Prices move between the trial and the season in which the
recommendation is used, usually further than the resampling interval.

**What prevents it.** `price_ratio` accepts a vector, and the printed output
says to read the table as a sensitivity analysis.

# 18. Reading the rectangle as the joint region

**The mistake.** Taking the two marginal intervals of a joint optimum as a
confidence region for the pair.

**Why it is wrong.** On a ridge the resampled optima lie along a diagonal, and
the rectangle admits corners no replicate visited.

**What prevents it.** `$region` and the printed rank correlation, and the
labelling of `box_lower` and `box_upper` as what they are.

# 19. Optimising two inputs one at a time

**The mistake.** Running `agri_np_optimum()` on N with P fixed, then on P with N
fixed, and reporting the pair.

**Why it is wrong.** That is the top of the surface only if the two act
additively, which is what a factorial rate trial exists to test.

**What prevents it.** `agri_np_optimum_surface()`, which refuses an additive
fit.

# 20. Leaving a field gradient in the residual

**The mistake.** Blocking and stopping there, in a trial that runs across a
slope.

**Why it is wrong.** A gradient that crosses blocks is not absorbed by them. It
stays in the residual, inflating every interval and every p-value.

**What prevents it.** `spatial =` with `coords =`, and the refusal to accept it
for an engine that cannot carry it.

# 21. Extrapolating a smoother

**The mistake.** Predicting at a rate beyond the largest one tested.

**Why it is wrong.** A nonparametric fit carries no information outside its
support. The value returned describes the basis.

**What prevents it.** `extrapolation = "warn"` by default and `"error"` on
request, plus the `extrapolated` flag on the returned rows.

---

# Part VII. Selection guide

# 22. Choose the interval by the sentence you are writing

| The sentence is about | Use |
|---|---|
| the average response at a rate | `interval = "confidence"` |
| what one plot in this trial will yield | `interval = "prediction"`, `scope = "within_block"` |
| what a plot in another field will yield | `interval = "prediction"`, `scope = "new_block"` |
| the shape of the whole curve | `agri_np_bootstrap(band = "simultaneous")` |

# 23. Choose the optimum by the decision

| The decision | Use |
|---|---|
| where the response peaks | `agri_np_optimum()` |
| how well that peak is located | `agri_np_optimum_test()` |
| what rate pays, given prices | `agri_np_optimum_economic()` |
| what pair of rates pays | `agri_np_optimum_surface()` |
| how many plants, sprays, irrigations | `agri_integer_optimum()` |
| whether an optimum exists at all | `agri_np_significant_slope()` |

---

# Part VIII. Minimum reporting checklist

# 24. What the methods section must contain

1. the engine, and whether a shape was imposed;
2. how the block entered, fixed or shrunk;
3. whether a field-position term was fitted, and which kind;
4. for an interval: confidence or prediction, and for a prediction interval, the
   scope;
5. for a validated error: the cross-validation scope;
6. for an economic optimum: the price ratio assumed, and the range explored;
7. for a joint optimum: that the region is a hull of resampled optima and not
   the rectangle of the marginal intervals;
8. the number of resampling replicates and the seed.

# 25. A worked methods paragraph

> Yield response to nitrogen was fitted by a penalised additive model with the
> block as a fixed effect and a two-dimensional smooth of row and column
> position to absorb the field gradient (agriRank 0.14.0). The agronomic optimum
> was 158 kg/ha. At a nitrogen to grain price ratio of 0.006 the economic
> optimum was 119 kg/ha, with a 95% cluster-bootstrap interval of 117 to 124
> kg/ha from 999 replicates resampling whole blocks (seed 1). Across the price
> ratios 0.002 to 0.020 the economic optimum ranged from 200 to 62 kg/ha, so the
> recommendation is more sensitive to price than to sampling error. Prediction
> intervals for individual plots were obtained by split conformal prediction
> with `scope = "new_block"`, which refers to a field not represented in the
> trial.

# 26. Where to go next

| If you now want | Read |
|---|---|
| whether the predictors and the shape survive a test | *Testing What a Model Assumes* |
| repeated measurements, several responses, missing plots | *Testing What a Model Assumes*, Part III |
| where the response is still changing | *Distribution-Free Uncertainty and Model Checking* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# Glossary

| Term | Meaning here |
|---|---|
| **confidence interval** | covers the mean response at a covariate setting |
| **prediction interval** | covers the next individual plot; always wider |
| **scope** | which plot the prediction interval refers to, in an observed block or a new one |
| **cross-validation scope** | which question the validated error answers, the same distinction applied to folds |
| **agronomic optimum** | the rate at which the response peaks |
| **economic optimum** | the rate at which the marginal product equals the price ratio |
| **price ratio** | price of a unit of input divided by price of a unit of produce |
| **joint optimum** | the pair of rates that optimises the surface, not the two separate optima |
| **ridge** | a surface along which one input compensates for the other |
| **field trend** | continuous fertility variation that crosses blocks |
| **extrapolation** | prediction outside the range of the observed predictor |
| **substream** | an independent random number stream, one per replicate |

# Selected references

- Cerrato, M. E. and Blackmer, A. M. (1990). Comparison of models for describing corn yield response to nitrogen fertilizer. *Agronomy Journal*, 82(1), 138-143.
- Lawless, C., Semenov, M. A. and Jamieson, P. D. (2008). Quantifying the effect of uncertainty in soil moisture characteristics on plant growth. *Field Crops Research*, 106(2), 138-147.
- Lei, J., G'Sell, M., Rinaldo, A., Tibshirani, R. J. and Wasserman, L. (2018). Distribution-free predictive inference for regression. *Journal of the American Statistical Association*, 113(523), 1094-1111.
- L'Ecuyer, P., Simard, R., Chen, E. J. and Kelton, W. D. (2002). An object-oriented random-number package with many long streams and substreams. *Operations Research*, 50(6), 1073-1075.
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R*, 2nd edition. Chapman and Hall.
