---
title: "Statistical foundations and state of the art"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical foundations and state of the art}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

This vignette summarizes the methodological foundations behind `agriRank` and states clearly which components are established methods, which are adapters to established software, and which are experimental package-level integrations.



## 1. From “nonparametric test” to estimand

Classical rank tests are often introduced as assumption-free substitutes for ANOVA. That description is incomplete. Modern nonparametric factorial inference defines treatment effects through distributional functionals such as relative treatment effects and uses rank or pseudo-rank estimators that remain meaningful for continuous, discrete, or ordinal outcomes.

## 2. Ranks versus pseudo-ranks

In unbalanced designs, ordinary pooled ranks induce sample-size-weighted effects. Brunner et al. showed that this can create counterintuitive changes in the target effect when allocation ratios change. Pseudo-ranks estimate unweighted relative effects and are therefore the preferred default when cell-size weighting is not scientifically intended.

## 3. ATS, WTS, and permutation

Wald-type statistics are asymptotically pivotal but can be liberal in smaller samples. ANOVA-type statistics provide useful approximations but are not generally exact. Studentized permutation procedures can improve calibration while retaining general factorial applicability. `rankFD` provides a mature implementation for independent factorial designs; `agriRank` uses it as a backend rather than reimplementing the same mathematics.

## 4. Aligned Rank Transform

ART aligns observations separately for each factorial effect before ranking and applying ANOVA machinery. It is useful for factorial main effects and interactions, and ART-C extends the framework to multifactor contrasts. ART is not identical to relative-effect inference and should therefore be presented as an alternative inferential paradigm rather than as a numerically interchangeable backend.

## 5. Randomization-restricted permutation

Permutation validity depends on exchangeability. In an RCBD, treatments are exchangeable within blocks; in split-plot designs, exchangeability is hierarchical; in repeated measures, within-subject observations are dependent. A design-aware package must compile these restrictions rather than permute the response globally.

## 6. Repeated measurements

`nparLD` established practical R workflows for rank-based longitudinal factorial inference. Wild-bootstrap procedures subsequently improved finite-sample behavior and enabled simultaneous contrasts and confidence intervals. `MANOVA.RM` provides a complementary resampling framework for multivariate and repeated outcomes.

## 7. Incomplete repeated measurements

Amro et al. extended nonparametric repeated-measures inference to incompletely observed data using quadratic-form statistics and wild bootstrap. The current native `agriRank` engine follows this framework for unblocked repeated designs, preserves subject-level dependence by using one multiplier per subject, and records the missingness assumption. It is explicitly marked experimental pending independent numerical benchmarking.

## 8. ANCOVA

The current `agri_ancova()` function is a Freedman-Lane permutation adapter through `permuco`, optionally applied to response mid-ranks. The 2026 resampling NANCOVA method is discussed as state of the art but is **not** claimed as implemented until a validated engine is added.

## 9. Reporting implications

The method section of an agronomic manuscript should state: design and experimental unit, factors and repeated structure, estimand, inferential statistic, resampling scheme and number of resamples, multiplicity control, missingness assumption, software and version, and random seed when applicable.

## 10. Package-level novelty

The package contribution is the integration of these established methods into a design-aware agricultural workflow: design declaration, admissible engine routing, preservation of randomization semantics, unified effects/contrasts, scientific graphics, missingness diagnostics, sensitivity analysis, and reproducible reporting. It does not claim invention of rankFD, ART, nparLD, MANOVA.RM, or the underlying wild-bootstrap theory.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.

# Nonparametric regression and quantitative agronomic gradients

A treatment factor is not the only useful representation of a quantitative agronomic intervention. When dose, salinity, irrigation depth, temperature or time is intrinsically ordered and sufficiently sampled, a smooth conditional-response function may address the scientific question more directly than a large collection of pairwise group comparisons. Version 0.10 therefore distinguishes strict nonparametric regression (LOESS, smoothing splines, kernel regression, isotonic regression and constrained quantile splines) from rank-robust and semiparametric companions (Theil-Sen/Siegel, quantile regression, GAM and SCAM).

Kernel methods in `np` are particularly relevant when continuous gradients coexist with ordered or unordered categorical predictors. GAMs provide penalized smooth estimation with modern smoothness selection, while SCAMs allow scientifically prespecified monotonicity and curvature constraints. Shape restrictions should be justified independently of observed p-values. See `vignette("v16-nonparametric-regression", package = "agriRank")`.
