---
title: "From experimental design to a reproducible agronomic report"
author: "agriRank Development Team"
date: "2026-08-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From experimental design to a reproducible agronomic report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
set.seed(151)
base <- expand.grid(block = factor(1:5), genotype = factor(c("G1", "G2", "G3")),
                    salinity = factor(c("0.5", "1.5", "2.5", "3.5")))
base$salinity_numeric <- as.numeric(as.character(base$salinity))
base$biomass <- rgamma(nrow(base), shape = 7, scale = 2) +
  2 * (base$genotype == "G2") - 1.5 * base$salinity_numeric +
  1.2 * (base$genotype == "G3" & base$salinity == "3.5")
```

This integrated example follows a salinity experiment from design declaration to factorial inference, contrasts, visualization, repeated chlorophyll measurements, missing-data diagnostics, and reproducible reporting.



## Agronomic question

Suppose three passion-fruit genotypes are evaluated under four salinity levels in five randomized complete blocks. Final biomass is measured once per plot, while SPAD chlorophyll readings are collected repeatedly at four dates.

## 1. Create a reproducible teaching data set

```r
set.seed(151)
base <- expand.grid(block = factor(1:5), genotype = factor(c("G1", "G2", "G3")),
                    salinity = factor(c("0.5", "1.5", "2.5", "3.5")))
base$salinity_numeric <- as.numeric(as.character(base$salinity))
base$biomass <- rgamma(nrow(base), shape = 7, scale = 2) +
  2 * (base$genotype == "G2") - 1.5 * base$salinity_numeric +
  1.2 * (base$genotype == "G3" & base$salinity == "3.5")
```

## 2. Declare and validate the factorial RCBD

```r
d_final <- agri_design(biomass ~ genotype * salinity, base, design = "rcbd", block = block)
validate_agri_design(d_final, error = FALSE)
design_summary(d_final)
```

## 3. Fit the final-biomass analysis

```r
fit_final <- if (requireNamespace("ARTool", quietly = TRUE))
  agri_rank(d_final, method = "ART") else if (requireNamespace("permuco", quietly = TRUE))
  agri_rank(d_final, method = "permuco") else NULL
fit_final
```

## 4. Effects, simple comparisons, and plots

```r
if (!is.null(fit_final)) agri_effects(fit_final)
if (!is.null(fit_final)) agri_pairs(fit_final, by = "salinity", adjust = "holm")
agri_plot(d_final, type = "data")
agri_plot(d_final, type = "interaction")
```

## 5. Create repeated SPAD measurements

```r
spad <- merge(base[c("block", "genotype", "salinity")], data.frame(time = factor(1:4)))
spad$plot <- interaction(spad$block, spad$genotype, spad$salinity, drop = TRUE)
set.seed(152)
spad$SPAD <- 38 + 2 * (spad$genotype == "G2") - 2.2 * as.numeric(spad$salinity) +
  1.4 * as.numeric(spad$time) + rnorm(nrow(spad), 0, 2.5)
```

This is a **blocked repeated experiment**. With complete observations, `permuco` is the current design-preserving route:

```r
d_spad <- agri_design(SPAD ~ genotype * salinity * time, spad, design = "repeated",
                      block = block, subject = plot, within = time)
if (requireNamespace("permuco", quietly = TRUE))
  fit_spad <- agri_repeated(d_spad, backend = "permuco", iter = 499)
```

## 6. Introduce missing repeated observations

```r
spad_miss <- spad
set.seed(153)
spad_miss$SPAD[sample(seq_len(nrow(spad_miss)), 12)] <- NA
 d_spad_miss <- agri_design(SPAD ~ genotype * salinity * time, spad_miss,
                            design = "repeated", block = block, subject = plot, within = time)
agri_missing_report(d_spad_miss)
```

The native incomplete wild-rank method is **not** allowed to ignore the agronomic block:

```r
try(agri_repeated(d_spad_miss, backend = "auto", B = 299))
```

This failure is a feature: the package refuses to produce a confirmatory result under a randomization structure that the selected method cannot represent.

## 7. Demonstrate the incomplete-data engine in an admissible unblocked repeated design

```r
u <- simulate_agri("repeated_missing", seed = 154, n = 12, missing_rate = .15)
d_u <- agri_design(height ~ treatment * time, u, design = "repeated",
                   subject = subject, within = time)
fit_u <- incomplete_wild_rank_test(d_u, B = 499, statistic = "ATS",
                                   missing_assumption = "MCAR", seed = 155)
fit_u$omnibus
agri_missing_sensitivity(d_u, B = 299, seed = 156)
```

## 8. Reproducible output

```r
if (!is.null(fit_final)) {
  agri_table(fit_final, "omnibus")
  agri_report(fit_final, tempfile(fileext = ".md"), format = "md", language = "en")
  export_results(fit_final, tempfile(fileext = ".rds"))
}
```

## Manuscript-ready interpretation template

A defensible report distinguishes the randomized block structure, the factorial interaction, the chosen nonparametric estimand, and the multiplicity procedure. Repeated SPAD measurements are analyzed with subject dependence preserved. When repeated observations are missing in a blocked experiment, the current package reports the methodological boundary rather than silently simplifying the design.
## Core methodological references

The documentation uses the bibliographic year of the version of record. Core references were cross-checked against the publisher/journal record and at least one independent bibliographic source; see `inst/references/REFERENCE_VERIFICATION.md` and `inst/references/agriRank-methods-verified.ris`.

- Pauly, M., Brunner, E., & Konietschke, F. (2015). Asymptotic permutation tests in general factorial designs. *Journal of the Royal Statistical Society: Series B*, 77(2), 461–473. DOI: 10.1111/rssb.12073.
- Brunner, E., Konietschke, F., Pauly, M., & Puri, M. L. (2017). Rank-based procedures in factorial designs: hypotheses about non-parametric treatment effects. *Journal of the Royal Statistical Society: Series B*, 79(5), 1463–1485. DOI: 10.1111/rssb.12222.
- Umlauft, M., Konietschke, F., & Pauly, M. (2017). Rank-based permutation approaches for non-parametric factorial designs. *British Journal of Mathematical and Statistical Psychology*, 70(3), 368–390. DOI: 10.1111/bmsp.12089.
- Friedrich, S., Konietschke, F., & Pauly, M. (2017). A wild bootstrap approach for nonparametric repeated measurements. *Computational Statistics & Data Analysis*, 113, 38–52. DOI: 10.1016/j.csda.2016.06.016.
- Umlauft, M., Placzek, M., Konietschke, F., & Pauly, M. (2019). Wild bootstrapping rank-based procedures: multiple testing in nonparametric factorial repeated measures designs. *Journal of Multivariate Analysis*, 171, 176–192. DOI: 10.1016/j.jmva.2018.12.005.
- Noguchi, K., Gel, Y. R., Brunner, E., & Konietschke, F. (2012). nparLD: An R software package for the nonparametric analysis of longitudinal data in factorial experiments. *Journal of Statistical Software*, 50(12), 1–23. DOI: 10.18637/jss.v050.i12.
- Friedrich, S., Konietschke, F., & Pauly, M. (2019). MANOVA.RM: Resampling-based analysis of multivariate data and repeated measures designs. *The R Journal*, 11(2), 380–400. DOI: 10.32614/RJ-2019-051.
- Frossard, J., & Renaud, O. (2021). Permutation tests for regression, ANOVA, and comparison of signals: the permuco package. *Journal of Statistical Software*, 99(15). DOI: 10.18637/jss.v099.i15.
- Hothorn, T., Hornik, K., van de Wiel, M. A., & Zeileis, A. (2008). Implementing a class of permutation tests: the coin package. *Journal of Statistical Software*, 28(8), 1–23. DOI: 10.18637/jss.v028.i08.
- Brunner, E., Konietschke, F., Bathke, A. C., & Pauly, M. (2021). Ranks and pseudo-ranks: surprising results of certain rank tests in unbalanced designs. *International Statistical Review*, 89(2), 349–366. DOI: 10.1111/insr.12418.
- Konietschke, F., & Brunner, E. (2023). rankFD: An R software package for nonparametric analysis of general factorial designs. *The R Journal*, 15(1), 142–158. DOI: 10.32614/RJ-2023-029.
- Amro, L., Konietschke, F., & Pauly, M. (2024). Incompletely observed nonparametric factorial designs with repeated measurements: a wild bootstrap approach. *Biometrical Journal*, 66(8), e70008. DOI: 10.1002/bimj.70008.
- Jan, S.-L., & Shieh, G. (2025). Power and sample size procedures for the Friedman test. *Sankhya B*, 87, 686–711. DOI: 10.1007/s13571-025-00362-2.
- Thiel, N. et al. (2026). Resampling NANCOVA for small samples. *Computational Statistics & Data Analysis*, 215, 108290. DOI: 10.1016/j.csda.2025.108290.
- Wobbrock, J. O., Findlater, L., Gergle, D., & Higgins, J. J. (2011). The aligned rank transform for nonparametric factorial analyses using only ANOVA procedures. *CHI 2011*, 143–146. DOI: 10.1145/1978942.1978963.
- Elkin, L. A., Kay, M., Higgins, J. J., & Wobbrock, J. O. (2021). An aligned rank transform procedure for multifactor contrast tests. *UIST 2021*, 754–768. DOI: 10.1145/3472749.3474784.

# Optional quantitative-gradient analysis

When salinity is recorded on a quantitative scale with enough distinct levels, the treatment can also be examined as a continuous gradient without replacing the design-aware factorial analysis. The regression analysis answers a different question: it estimates the conditional response curve rather than comparing a finite set of treatment levels.

```{r}
if (requireNamespace("mgcv", quietly = TRUE)) {
  reg <- agri_np_regression(biomass ~ salinity_numeric, base, method = "gam", block = block)
  agri_np_plot(reg)
  agri_np_derivative(reg)
}
```

This complementary analysis should be prespecified when it is intended to support confirmatory inference. It should not be added only because factor-based comparisons were inconvenient or nonsignificant.
