---
title: "Testing What a Model Assumes, and Data That Resist"
author: "agriRank Development Team"
date: "2026-08-24"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing What a Model Assumes, and Data That Resist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r agrirank-setup, include = FALSE}
library(agriRank)
set.seed(2026)
```

# 1. Why this vignette exists

Every previous regression vignette took the model as given and asked what it
said. This one asks the model to justify itself, and then takes on three kinds
of data that the module could not previously handle at all.

Two questions about the model:

**Does this predictor earn its place?** The regression module fits a curve and
reports an interval around it, but until 0.14.0 it had no test that the
predictor mattered at all, except for two of the sixteen engines through a
route that ignored the block.

**Is the shape I imposed compatible with the data?** `shape = "increasing"`
buys precision when the response really is increasing and biases the curve when
it is not. Nothing checked which case applied.

Three kinds of data:

**Repeated measurements.** Four harvests on the same plots are four rows per
plot, not four independent observations.

**Several responses.** Nitrogen raises yield and moves protein. The two optima
come from the same plots, so their errors are correlated, and separate analyses
cannot say how far apart they are.

**Missing plots.** The plot that was lost is often the flooded one.

## 1.1 The single device behind Part I

The two tests share one machine, the **cluster wild bootstrap**: refit under the
null, then build replicate responses from the null fitted values plus the
residuals multiplied by random signs drawn **once per block** rather than once
per plot.

Signing plots independently would treat the plots of a block as independent,
which is the error this whole package exists to prevent. Signing whole blocks
leaves the within-block dependence intact whatever its form. Only the null model
differs between the two tests.

# 2. Learning objectives

By the end you should be able to:

1. test whether a predictor contributes, for any engine, respecting the block;
2. read the limit that the number of blocks places on any such p-value;
3. test a shape constraint and say what a large p-value does and does not mean;
4. fit a curve to repeated measurements without inflating the replication;
5. report two optima with the distance between them, not two separate intervals;
6. use multiple imputation as a sensitivity analysis rather than as an answer.

# 3. The module map

| Question | Function | Null or structure |
|---|---|---|
| does this term matter | `agri_np_effect_test()` | the model without it |
| is this shape compatible | `agri_np_shape_test()` | the constrained fit |
| repeated measurements | `agri_np_longitudinal()` | subject shrunk, subject resampled |
| several responses | `agri_np_multiresponse()` | one shared bootstrap |
| missing plots | `agri_np_impute()` | MAR, reported beside complete case |

---

# Part I. Making the model justify itself

# 4. The trial, with a predictor that does nothing

```{r trial}
d <- expand.grid(N = seq(0, 200, by = 50), block = factor(1:5), rep = 1:2)
d$noise <- rnorm(nrow(d))            # a covariate with no effect at all
d$yield <- 3 + 0.030 * d$N - 0.00009 * d$N^2 +
  as.numeric(d$block) * 0.30 + rnorm(nrow(d), 0, 0.30)

fit <- agri_np_regression(yield ~ N + noise, d, method = "gam",
                          block = block, k = 5)
```

`noise` was generated independently of yield. A test worth having should be able
to tell it from nitrogen.

## 4.1 The test

```{r effect-test}
et <- agri_np_effect_test(fit, B = 99, seed = 1)
et
```

## 4.2 Interpretation

Look at the **statistic** before the p-value. Nitrogen and the noise variable
differ by three orders of magnitude, which is the discrimination the test
actually achieved.

The p-values are much closer together, and the printed note explains why: with
five blocks there are only $2^5 = 32$ distinct sign patterns, so no p-value
below about 0.03 can be produced however large `B` is. That is a limit of the
**design**, not of the resampling. More blocks is the remedy; more replicates
is not.

This is worth stating plainly because it is easy to misread a p-value of 0.08
from five blocks as weak evidence, when in fact the test had almost no room to
produce anything smaller.

## 4.3 What the null actually is

The null is that the term does not enter the response **at all**, not that its
effect is linear or small. A term that is not rejected has not been shown to be
absent, and with five blocks that statement is very weak indeed.

## 4.4 It works where the old test could not

```{r effect-any-engine}
fs <- agri_np_regression(yield ~ N, d, method = "smoothing_spline")
agri_np_effect_test(fs, B = 99, seed = 1, cluster = NA)$table
```

A smoothing spline has no coefficients to test and is outside the reach of
`agri_np_significance()`, which calls `np::npsigtest()` and works only for the
kernel engines. This test looks only at fitted values and residuals, so it works
for all sixteen.

Note `cluster = NA` here, because this fit declares no block. Signing rows
independently is legitimate only for a completely randomized layout, and the
printed header says which assumption was made.

## 4.5 The asymmetry this removed

Before 0.14.0 every **interval** in the regression module resampled whole
blocks, while the only **p-value** available resampled rows. A p-value and an
interval from the same fit therefore rested on different assumptions, and
nothing in the output said so. That is now visible in both directions:
`agri_np_significance()` prints its caveat, and `agri_np_effect_test()` exists.

# 5. Is the shape I imposed compatible with the data

```{r shape-fit}
f1 <- agri_np_regression(yield ~ N, d, method = "gam", block = block, k = 5)
```

## 5.1 A shape agronomy expects

```{r shape-concave}
agri_np_shape_test(f1, shape = "increasing_concave", B = 99, seed = 1)
```

## 5.2 A shape the data should contradict

```{r shape-increasing}
agri_np_shape_test(f1, shape = "increasing", B = 99, seed = 1)
```

## 5.3 Interpretation

The response rises and then turns over inside the tested range, so a purely
increasing constraint has to distort the fit near the top, and the second test
should see more distortion than the first.

**A large p-value does not prove the shape.** It says these data do not
contradict it, which is the most a test of this kind can say, and with five
blocks it is a weak statement for the reason given in section 4.2.

## 5.4 Read it beside the SiZer

```{r sizer, eval = requireNamespace("SiZer", quietly = TRUE)}
agri_np_significant_slope(agri_np_sizer(f1), stability = 0.8)
```

A shape test compresses the whole curve into one number. The SiZer says **where**
the free fit changes direction, across a column of bandwidths, and that is
usually the more useful statement. Use the test to decide whether to impose the
constraint and the SiZer to describe the response.

## 5.5 A diagnostic the print method performs for you

If the constrained fit has the smaller RMSE, which cannot happen by
optimisation, the two engines differ in more than the constraint and the printed
output says so. Compare like with like before reading the p-value.

---

# Part II. Repeated measurements

# 6. Four harvests on the same plots

```{r long-data}
lg <- expand.grid(N = seq(0, 200, 50), plot = factor(1:12), time = factor(1:3))
lg$yield <- 3 + 0.03 * lg$N - 0.00009 * lg$N^2 +
  as.numeric(lg$plot) * 0.15 + as.numeric(lg$time) * 0.40 +
  rnorm(nrow(lg), 0, 0.25)
nrow(lg)
```

## 6.1 The error this invites

Fitting `yield ~ N` to these `r nrow(lg)` rows treats them as `r nrow(lg)`
independent observations. There are twelve plots. The apparent replication is
inflated threefold, and every interval is too narrow by roughly the square root
of that.

## 6.2 The fit

```{r long-fit}
lf <- agri_np_longitudinal(yield ~ N, lg, subject = plot, time = time, k = 4)
lf
```

## 6.3 What happened structurally

The subject became the **block**. That is not a trick: the subject *is* the unit
of resampling, so every tool in the module then does the right thing without
knowing anything about repeated measurement.

```{r long-downstream}
b <- agri_np_bootstrap(lf, B = 99, n = 25, seed = 1)
attr(b, "cluster")
```

The bootstrap resamples whole plots. `agri_np_conformal()` splits by plot.
`agri_np_compare(cv_scope = "new_block")` holds out whole plots, which is the
honest question: how well is a plot predicted that was never measured.

## 6.4 Letting the shape change over time

```{r long-varying}
lv <- agri_np_longitudinal(yield ~ N, lg, subject = plot, time = time,
                           time_effect = "varying", k = 4)
deparse(lv$formula_used)
```

`"smooth"` lets the level drift between occasions while the shape stays the
same. `"varying"` lets the shape itself differ, which is what a multi-harvest
trial usually asks: does the response to nitrogen change as the season advances.

## 6.5 What this is not

**It is not a GAMM with a modelled within-subject correlation.** Nothing here
estimates an autocorrelation over time. The dependence is handled by resampling
whole subjects, which assumes nothing about its form and, in exchange, cannot
recover the efficiency that a correct correlation model would.

If the occasions are many and closely spaced, a weekly series over a season say,
a model that represents the correlation is the better tool and this one will be
conservative. Say which you used.

## 6.6 Data that are not repeated measurements

```{r long-refuse, error = TRUE}
lg1 <- lg[!duplicated(lg$plot), ]
agri_np_longitudinal(yield ~ N, lg1, subject = plot, time = time)
```

---

# Part III. Several responses to one gradient

# 7. Yield and protein

```{r mv-data}
mv <- expand.grid(N = seq(0, 200, 50), block = factor(1:5))
mv$yield   <- 3 + 0.040 * mv$N - 0.00020 * mv$N^2 +
  as.numeric(mv$block) * 0.2 + rnorm(nrow(mv), 0, 0.15)
mv$protein <- 9 + 0.026 * mv$N - 0.00009 * mv$N^2 +
  as.numeric(mv$block) * 0.1 + rnorm(nrow(mv), 0, 0.15)
```

## 7.1 The analysis

```{r mv-fit}
mr <- agri_np_multiresponse(cbind(yield, protein) ~ N, mv, block = block,
                            B = 99, seed = 1, n = 60, k = 4)
mr
```

## 7.2 What the shared bootstrap buys

Every response is refitted on the **same** resampled blocks within a replicate.
Calling `agri_np_optimum_test()` twice would give each response its own
resampled experiment and destroy exactly the dependence the joint question turns
on.

The consequence is the row that separate analyses cannot produce: the
**distance** between the two optima, with its own interval and p-value. That
distance, not either optimum on its own, is what a joint recommendation has to
face.

## 7.3 The joint region

```{r mv-region}
head(mr$region)
```

For two responses the region is the convex hull of the replicate pairs. It is
not the rectangle formed by the two marginal intervals printed above, for the
same reason as in the two-predictor case: the pairs may lie along a diagonal,
and the rectangle admits corners that no replicate visited.

## 7.4 One maximised, one minimised

```{r mv-mixed}
mv$lodging <- 20 + 0.10 * mv$N + rnorm(nrow(mv), 0, 2)
agri_np_multiresponse(cbind(yield, lodging) ~ N, mv, block = block,
                      objective = c("max", "min"), B = 99, seed = 1,
                      n = 40, k = 4)$optima
```

A single `objective` would have been wrong here. Yield is maximised and lodging
minimised, and the function accepts one entry per response for that reason.

## 7.5 Interpretation, and what the function refuses to do

**A joint region is not a compromise rate.** Choosing one rate for two responses
is a decision about their relative value, an economic and agronomic judgement,
not a statistical question. This function supplies how far apart the optima are
and how well that distance is determined, and stops there.

When prices are what decide, run `agri_np_optimum_economic()` on each response
with its own price ratio, and read the two together.

---

# Part IV. Missing plots

# 8. The plot that was lost is often the flooded one

```{r imp-data}
mi <- d[, c("N", "block", "yield")]
mi$yield[sample(nrow(mi), 8)] <- NA
sum(is.na(mi$yield))
```

## 8.1 What the module did before

Complete-case analysis, with `na_action = "fail"` as the default so that rows
were never dropped silently. That default is right and stays. But complete-case
analysis is unbiased only when the missingness is unrelated to the response, and
in a field trial it usually is not.

## 8.2 Multiple imputation, always beside the complete case

```{r imp-fit}
im <- agri_np_impute(yield ~ N, mi, block = block, m = 5, B = 99,
                     seed = 1, n = 40, k = 4)
im
```

## 8.3 Interpretation

**This is the one place in the package that assumes a missingness mechanism.**
Multiple imputation is valid when the data are missing at random given the
observed variables, and that is an assumption about *why* the plots were lost,
not a property these data can confirm.

Everything else in agriRank avoids such assumptions. This function makes one
deliberately, and never returns only the imputed answer: the complete-case fit
is always computed and printed beside it, and the gap between the two optima is
reported as a percentage of the tested range.

Read the two together:

- if they agree, the missingness is not driving the conclusion and the
  complete-case analysis can be reported as the primary one;
- if they disagree, **that disagreement is the finding**. The conclusion depends
  on what was assumed about the plots that were lost, and the methods section
  has to say so.

## 8.4 How much the imputation is doing

```{r imp-fmi}
head(im$curve[, c("x", "complete_case", "pooled", "fmi")])
```

`fmi` is the fraction of missing information at each point of the curve. Where
it is near zero the imputation changed nothing there; where it is large the
pooled answer rests mostly on the assumption.

## 8.5 The pooling

The pooled variance at each grid point is the average within-imputation
bootstrap variance plus $(1 + 1/m)$ times the variance of the $m$ point
estimates, which is Rubin's rule. The **block survives into the within part**,
because that part uses the package's own cluster bootstrap rather than a
model-based standard error.

## 8.6 With nothing missing

```{r imp-refuse, error = TRUE}
agri_np_impute(yield ~ N, mi[!is.na(mi$yield), ], block = block, m = 3)
```

---

# Part V. Common mistakes, and the function that prevents each

# 9. Quoting a p-value that the design could not have produced

**The mistake.** Reporting p = 0.08 from five blocks as weak evidence of no
effect.

**Why it is wrong.** With five blocks the cluster wild bootstrap has 32 possible
outcomes and cannot produce anything below about 0.03. The test had almost no
room.

**What prevents it.** `agri_np_effect_test()` prints the $2^G$ limit when the
number of blocks is small, and says that more blocks, not more replicates, is
the remedy.

# 10. Reading a large shape-test p-value as proof of the shape

**The mistake.** "The monotonicity test was not significant, so the response is
monotone."

**Why it is wrong.** Failing to reject is not evidence for the null, especially
with few blocks.

**What prevents it.** The printed note says exactly this, and points to
`agri_np_sizer()` for the statement that can actually be made.

# 11. Treating repeated measurements as replicates

**The mistake.** Fitting `yield ~ N` to a long-format table with four harvests
per plot.

**Why it is wrong.** The replication is inflated by the number of occasions and
every interval is too narrow.

**What prevents it.** `agri_np_longitudinal()`, which also refuses data in which
units are measured only once.

# 12. Analysing two responses separately and comparing their intervals

**The mistake.** Running `agri_np_optimum_test()` on yield and on protein, then
saying the optima differ because the intervals do not overlap.

**Why it is wrong.** The two optima come from the same plots. Their errors are
correlated, so the overlap of marginal intervals is not the test of their
difference.

**What prevents it.** `agri_np_multiresponse()`, whose shared bootstrap gives
the difference its own interval and p-value.

# 13. Reporting only the imputed analysis

**The mistake.** Running multiple imputation and reporting the pooled curve as
the result.

**Why it is wrong.** The pooled curve rests on an untestable assumption about
why the plots were lost.

**What prevents it.** `agri_np_impute()` always computes and prints the
complete-case fit beside it, and quantifies the gap.

# 14. Using imputation to rescue an underpowered trial

**The mistake.** Imputing a large fraction of the response and reporting the
narrower pooled interval.

**Why it is wrong.** Imputation propagates uncertainty, it does not create
information. A high `fmi` says most of the answer at that point is assumption.

**What prevents it.** The `fmi` column, and the largest value along the curve
printed with the result.

---

# Part VI. Selection guide

# 15. Choose the test by the question

| The question | Use |
|---|---|
| does this term contribute at all | `agri_np_effect_test()` |
| is the constraint I imposed compatible | `agri_np_shape_test()` |
| where is the response still changing | `agri_np_sizer()` |
| how well is the optimum located | `agri_np_optimum_test()` |
| kernel engine, and I want the standard route | `agri_np_significance()` |

# 16. Choose the structure by the data

| The data | Use |
|---|---|
| one measurement per plot | `agri_np_regression()` |
| several occasions per plot | `agri_np_longitudinal()` |
| several responses per plot | `agri_np_multiresponse()` |
| plots missing | `agri_np_impute()`, beside complete case |
| plots missing, ranks rather than a curve | `agri_missing_sensitivity()` |

---

# Part VII. Minimum reporting checklist

# 17. What the methods section must contain

1. for any test here: the resampling unit, and that signs were drawn per block;
2. the number of blocks, since it bounds the attainable p-value;
3. the number of replicates and the seed;
4. for a shape test: which constraint, and which free engine it was compared to;
5. for longitudinal data: the subject, the number of occasions, whether the
   design is balanced, and that dependence was handled by resampling subjects
   rather than by a modelled correlation;
6. for several responses: that one shared bootstrap was used, and the distance
   between optima with its interval;
7. for imputation: the mechanism assumed, the number of imputations, the
   imputation method, the fraction of missing information, and **both** the
   pooled and complete-case results.

# 18. A worked methods paragraph

> The contribution of each predictor was assessed by a cluster wild bootstrap
> with Rademacher signs drawn once per block, 999 replicates, seed 1 (agriRank
> 0.14.0). With five blocks this procedure admits 32 distinct sign patterns, so
> no p-value below 0.031 is attainable and the statistic is reported alongside.
> The assumption that the response is increasing and concave was tested against
> an unconstrained additive fit by the same device and was not contradicted
> (p = 0.52), which we read as compatibility rather than as evidence for the
> constraint. Yield and grain protein were analysed jointly with a single shared
> bootstrap, so that the correlation between their optima is preserved; the two
> optima differed by 37 kg/ha (95% interval 33 to 47). Eight plots were lost.
> Results are reported for the complete cases, with a multiple-imputation
> sensitivity analysis (m = 5, predictive mean matching) that moved the optimum
> by less than 2% of the tested range.

# 19. Where to go next

| If you now want | Read |
|---|---|
| the economic and joint optima | *From a Curve to a Decision* |
| the engines themselves | *Nonparametric and Shape-Aware Regression* |
| conformal intervals and simulated residuals | *Distribution-Free Uncertainty and Model Checking* |
| the rank-based repeated-measures side | *Repeated Measures and Missing Longitudinal Data* |
| the whole workflow on one experiment | *Integrated Agronomic Case Study* |

---

# Glossary

| Term | Meaning here |
|---|---|
| **cluster wild bootstrap** | replicate responses built from null fitted values plus residuals times random signs, drawn once per cluster |
| **Rademacher signs** | random values of $-1$ or $+1$ |
| **null model** | the fit under the hypothesis being tested: without the term, or with the constraint |
| **$2^G$ limit** | with $G$ blocks only $2^G$ sign patterns exist, bounding the smallest attainable p-value |
| **shape constraint** | monotonicity or curvature imposed on the fitted curve |
| **subject** | the unit measured more than once |
| **shared bootstrap** | one resample of blocks per replicate, reused for every response |
| **MAR** | missing at random given the observed variables |
| **fmi** | fraction of missing information, how much of the pooled variance comes from the imputation |
| **Rubin's rules** | pooling of estimates and variances across imputations |

# Selected references

- Cameron, A. C., Gelbach, J. B. and Miller, D. L. (2008). Bootstrap-based improvements for inference with clustered errors. *The Review of Economics and Statistics*, 90(3), 414-427.
- Haerdle, W. and Mammen, E. (1993). Comparing nonparametric versus parametric regression fits. *The Annals of Statistics*, 21(4), 1926-1947.
- Webb, M. D. (2023). Reworking wild bootstrap-based inference for clustered errors. *Canadian Journal of Economics*, 56(3), 1113-1141.
- Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- van Buuren, S. and Groothuis-Oudshoorn, K. (2011). mice: Multivariate imputation by chained equations in R. *Journal of Statistical Software*, 45(3), 1-67.
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R*, 2nd edition. Chapman and Hall.
