---
title: "Nonparametric and Shape-Aware Regression for Agronomic Gradients"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonparametric and Shape-Aware Regression for Agronomic Gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

# Why a regression module belongs in agriRank

Agronomic experiments frequently contain a quantitative treatment or environmental gradient: fertilizer dose, salinity, irrigation depth, temperature, soil water potential, plant age, spectral index, or a continuous soil property. Treating every observed value as an unrelated factor can discard ordering and may generate an unnecessarily large set of pairwise comparisons. Conversely, imposing a polynomial or other parametric curve without scientific justification can force the response into an inappropriate shape.

`agriRank` therefore separates **strictly nonparametric smoothers**, **rank-robust regression**, **conditional-quantile regression**, and **semiparametric smooth regression** under one interface. The model is not selected by the smallest p-value.

# State of the art

## Rank-robust slopes

The Theil-Sen estimator defines a slope from the median of pairwise slopes and provides a classical rank-based alternative when a monotone approximately linear relationship is scientifically plausible but outliers make least squares unattractive. Sen's 1968 treatment linked this estimator to Kendall's tau. Siegel repeated medians increase resistance to contamination further. In `agriRank` these are exposed through `method = "theil_sen"` and `method = "siegel"` using the `mblm` backend.

## Local smoothers and smoothing splines

`method = "loess"` uses local polynomial regression. `method = "smoothing_spline"` uses penalized smoothing splines from base R. Both are useful exploratory and descriptive tools when the relationship is continuous but its form is unknown. They are currently restricted to settings in which a declared agronomic block need not be adjusted by the model.

## Mixed-data kernel regression

`method = "kernel"` uses `np`. Kernel regression is especially attractive when continuous gradients occur together with ordered or unordered categorical predictors. Bandwidth selection is treated as a central estimation problem rather than a cosmetic plotting choice. The `np` framework also supports bootstrap-based uncertainty and significance testing in its native API.

## Quantile regression and constrained quantile splines

The conditional mean can conceal heterogeneous agronomic responses. For example, a fertilizer dose may have a modest effect on median yield but a much larger effect in the lower tail under stressful environments. `method = "quantile"` uses `quantreg::rq()` for conditional quantiles. `method = "cobs"` uses constrained quantile B-splines and supports monotonic or curvature constraints for a single numeric gradient.

## GAM and SCAM

Generalized additive models estimate smooth functions while allowing non-Gaussian response families and multiple covariates. `method = "gam"` uses `mgcv`. When an agronomic relationship has a defensible shape restriction, such as non-decreasing biomass with accumulated thermal time or a monotone decrease with salinity, `method = "scam"` uses shape-constrained additive models. Supported constraints include increasing, decreasing, convex, concave, and monotonicity combined with curvature constraints.

Shape constraints must be scientifically declared. They are not learned by trying several shapes and retaining whichever produces the smallest p-value.

# The unified API

The examples below use `agri_dose`, an exported data set: eight nitrogen rates,
from 0 to 280 kg ha^-1^, evaluated in five complete blocks, with yield in
Mg ha^-1^. Because it ships with the package, every object created here stays
available for you to inspect afterwards.

```{r dose-data}
data(agri_dose)
str(agri_dose)
```

```{r}
dose_data <- agri_dose

fit_ss <- agri_np_regression(
  yield ~ dose,
  data = dose_data,
  method = "smoothing_spline"
)
fit_ss
```

The same downstream verbs apply to every engine:

```{r}
agri_np_predict(fit_ss, data.frame(dose = c(50, 100, 150, 200)))
agri_np_diagnostics(fit_ss)
agri_np_derivative(fit_ss)
agri_np_optimum(fit_ss)
agri_np_plot(fit_ss)
```

## Automatic engine selection

`method = "auto"` is the default. It does not compare p-values or goodness of
fit. It applies a fixed rule to the declaration you made: a shape constraint
routes to SCAM or COBS, a non-Gaussian family routes to GAM, a single numeric
predictor with no block routes to a smoothing spline, and anything else routes
to GAM or kernel regression depending on what is installed.

```{r auto-selection}
auto_simple <- agri_np_regression(yield ~ dose, dose_data)
auto_shaped <- agri_np_regression(yield ~ dose, dose_data, shape = "concave")

data.frame(
  declaration = c("one numeric predictor, no constraint",
                  "concavity declared"),
  engine = c(auto_simple$method, auto_shaped$method)
)
```

The rule is deterministic and stated in advance, so the engine is a consequence
of the scientific declaration rather than of the data. When the declaration is
ambiguous, name the engine explicitly.

For an integer-valued treatment such as plants per hill or number of
irrigations, the decision support must also be declared. That case is covered
in `vignette("v17-integer-support-regression")`.

# Example 1: fertilizer dose-response without imposing a quadratic curve

```{r}
p1 <- agri_np_plot(fit_ss, type = "fit")
p1

agri_np_plot(fit_ss, type = "derivative")
agri_np_optimum(fit_ss)
```

The optimum returned here is **descriptive**. It is the maximum of the selected smooth over the observed search range. It is not an economic optimum and does not replace a prespecified response model when a mechanistic or economic model is required.

# Example 2: salinity response with a scientific monotonicity constraint

If biomass is expected to decrease as salinity increases over the experimental domain, a shape-constrained smoother can prevent biologically implausible reversals caused only by local noise.

```{r}
set.seed(2002)
sal <- data.frame(ec = seq(0.5, 6.0, length.out = 60))
sal$biomass <- 31 * exp(-0.12 * sal$ec^1.45) + rnorm(nrow(sal), 0, 1.1)

fit_iso <- agri_np_regression(
  biomass ~ ec,
  data = sal,
  method = "isotonic",
  shape = "decreasing"
)
agri_np_plot(fit_iso)
```

When `scam` is installed, the same scientific restriction can be embedded in a smooth additive model:

```{r}
if (requireNamespace("scam", quietly = TRUE)) {
  fit_scam <- agri_np_regression(
    biomass ~ ec,
    data = sal,
    method = "scam",
    shape = "decreasing"
  )
  agri_np_plot(fit_scam, interval = TRUE)
}
```

# Example 3: RCBD-adjusted smooth regression

A quantitative treatment in an RCBD should not lose its block structure merely because the treatment is modeled continuously. Engines that cannot adjust for the declared block are rejected by `agriRank`.

```{r}
set.seed(2003)
rcbd <- expand.grid(
  block = factor(1:5),
  N = seq(0, 200, length.out = 12)
)
rcbd$yield <- 6 + 0.075 * rcbd$N - 0.00021 * rcbd$N^2 +
  as.numeric(rcbd$block) * 0.35 + rnorm(nrow(rcbd), 0, 0.55)

if (requireNamespace("mgcv", quietly = TRUE)) {
  fit_gam <- agri_np_regression(
    yield ~ N,
    data = rcbd,
    method = "gam",
    block = block
  )
  summary(fit_gam)
  agri_np_plot(fit_gam, predictor = "N", interval = TRUE)
}
```

The block is included as an adjustment factor. A method such as one-dimensional smoothing spline is not silently substituted because doing so would discard design information.

# Conditional quantiles

```{r}
if (requireNamespace("quantreg", quietly = TRUE)) {
  fit_q50 <- agri_np_regression(yield ~ N, rcbd, method = "quantile", tau = 0.50, block = block)
  fit_q10 <- agri_np_regression(yield ~ N, rcbd, method = "quantile", tau = 0.10, block = block)
  agri_np_predict(fit_q50, rcbd[1:5, ])
  agri_np_predict(fit_q10, rcbd[1:5, ])
}
```

Different quantiles answer different scientific questions. Quantile regression should not be presented as a generic replacement for mean regression.

# Kernel regression with mixed predictors

The current `np` package supports continuous, ordered and unordered predictors in the same kernel framework. This can be useful for continuous fertilizer dose together with cultivar or soil class.

```{r}
if (requireNamespace("np", quietly = TRUE)) {
  set.seed(2004)
  kd <- expand.grid(
    cultivar = factor(c("A", "B")),
    dose = seq(0, 180, length.out = 20)
  )
  kd$yield <- 5 + 0.06 * kd$dose - 0.00018 * kd$dose^2 +
    ifelse(kd$cultivar == "B", 1.2, 0) + rnorm(nrow(kd), 0, .45)
  fit_kernel <- agri_np_regression(yield ~ dose + cultivar, kd, method = "kernel")
  agri_np_diagnostics(fit_kernel)
}
```

# Predictive model comparison

`agri_np_compare()` evaluates predictive error by cross-validation. It does **not** select an inferential model by searching for the smallest p-value.

```{r}
agri_np_compare(
  yield ~ dose,
  dose_data,
  methods = c("smoothing_spline", "loess"),
  kfold = 5,
  metric = "RMSE"
)
```

For blocked data, fold assignment can be stratified within blocks:

```{r}
if (requireNamespace("mgcv", quietly = TRUE)) {
  agri_np_compare(
    yield ~ N,
    rcbd,
    methods = "gam",
    block = block,
    kfold = 4
  )
}
```

# Bootstrap uncertainty and agronomic experimental units

When a block is declared, `agri_np_bootstrap()` resamples complete blocks by default. This avoids pretending that all observations are independent rows.

```{r}
if (requireNamespace("mgcv", quietly = TRUE)) {
  bb <- agri_np_bootstrap(fit_gam, B = 99, n = 80, seed = 2005)
  head(bb)
}
```

For final inference use a substantially larger number of bootstrap replicates than the small values used in teaching examples.

# Recommended workflow

1. Identify the experimental unit and whether blocks or repeated units must be retained.
2. State the scientific estimand: conditional mean, conditional quantile, monotone trend, or descriptive smooth response.
3. Decide whether any shape restriction is scientifically justified before fitting the data.
4. Fit one or more scientifically admissible models.
5. Examine predictive error and residual behavior without selecting models by the smallest p-value.
6. Display observed data together with fitted curves and uncertainty.
7. Treat numerical derivatives and nonparametric optima as model-dependent estimates.
8. Use cluster-aware resampling when the experimental design requires it.

# Two-dimensional agronomic response surfaces

When two treatment or environmental gradients are quantitative, a flexible surface can be more informative than separate one-dimensional curves. This is useful for nutrient-by-water gradients, temperature-by-soil-moisture responses, or interacting continuous soil properties. The surface remains conditional on any other covariates held fixed.

```{r}
if (requireNamespace("mgcv", quietly = TRUE)) {
  set.seed(2006)
  surf <- expand.grid(
    N = seq(0, 200, length.out = 15),
    water = seq(0.5, 1.0, length.out = 10)
  )
  surf$yield <- 5 + 0.065 * surf$N - 0.00018 * surf$N^2 +
    5 * surf$water + 0.015 * surf$N * surf$water + rnorm(nrow(surf), 0, .55)

  fit_surface <- agri_np_regression(yield ~ N + water, surf, method = "gam", gam_structure = "tensor")
  agri_np_plot(
    fit_surface,
    type = "surface",
    surface_predictors = c("N", "water"),
    n = 50
  )
}
```

The surface should not be extrapolated beyond the observed experimental domain. Sparse corners of a factorial gradient deserve particular caution even when the fitted surface appears visually smooth.

# Group-specific curves

A factor can be retained in the regression and displayed as separate conditional curves. For genuinely different curve shapes, specify the interaction or factor-specific smooth in the backend-compatible formula rather than assuming that a main-effect adjustment is sufficient.

```{r}
if (requireNamespace("mgcv", quietly = TRUE)) {
  set.seed(2007)
  gd <- expand.grid(cultivar = factor(c("A", "B", "C")), dose = seq(0, 180, length.out = 18))
  gd$yield <- 5 + .06 * gd$dose - .00017 * gd$dose^2 +
    c(A = 0, B = 1.2, C = -0.6)[gd$cultivar] + rnorm(nrow(gd), 0, .5)
  gfit <- agri_np_regression(yield ~ dose + cultivar, gd, method = "gam")
  agri_np_plot(gfit, predictor = "dose", group = "cultivar")
}
```

# Qualitative factors in regression models

A quantitative gradient is frequently observed together with a qualitative factor: cultivar, soil class, irrigation system, or management practice. `agri_np_regression()` treats such predictors explicitly. A character column is read as the qualitative factor it is, and a factor needs at least two levels to enter a model, because a single level carries no contrast against the response. Engines that cannot adjust for a qualitative predictor refuse it by name; `quantile`, `kernel`, GAM and SCAM keep the factor as an adjustment term.

For engines with interpretable coefficients, each non-reference level of a factor enters as one coefficient, and its bootstrap interval can be read against zero in a forest plot. The coefficient bootstrap aligns replicates by term name, and block adjustment terms are excluded as nuisance parameters, so a block-adjusted fit reports intervals for the scientific coefficients of the declared formula.

```{r}
if (requireNamespace("quantreg", quietly = TRUE)) {
  dzc <- dose_data
  dzc$cultivar <- factor(rep(c("Ana", "Bela"), length.out = nrow(dzc)))
  dzc$yield <- dzc$yield + ifelse(dzc$cultivar == "Bela", 0.8, 0)

  fit_qf <- agri_np_regression(
    yield ~ dose + cultivar,
    data = dzc,
    method = "quantile",
    block = block
  )
  coef(fit_qf)

  # B = 199 keeps the vignette fast; a final analysis needs B >= 999.
  # suppressWarnings() hides only the quantreg backend notice that the
  # linear-program solution is nonunique at tied responses; it is a property
  # of the estimator at this tau, not a defect of the analysis.
  bt_coef <- suppressWarnings(
    agri_np_bootstrap(fit_qf, target = "coefficients", B = 199, seed = 2008)
  )
  agri_np_forest(fit_qf, bootstrap = bt_coef)
}
```

The figure stacks the nitrogen slope and the cultivar contrast in their own panels and draws the reference level of the factor at zero, so every level appears. Cultivar Bela shifts the whole dose-response curve upward at every nitrogen rate. The same figure is available as `agri_np_plot(fit_qf, type = "forest", bootstrap = bt_coef)`.

# From fit to journal: tables, curves with bands, and export

Two readings of a qualitative factor are legitimate and the package keeps them as separate outputs. The coefficient forest plot states each level as a **contrast against the reference level**. The level summary states what the model predicts **at each level itself**, which is what manuscripts usually report.

```{r}
if (exists("fit_qf")) {
  agri_table(fit_qf, "coefficients", method = "bootstrap", B = 199, seed = 2010)
  lv <- agri_np_levels(fit_qf, B = 199, seed = 2010)
  lv
}
```

`agri_np_levels()` returns the observed sample size, median/MAD and mean/sd of the response at each level, beside the fitted marginal response with a pointwise bootstrap interval at the reference values of the other covariates. The corresponding figure shows the observed values beneath the fitted response:

```{r}
if (exists("lv")) {
  agri_np_plot(fit_qf, type = "levels", bootstrap = attr(lv, "bootstrap"))
}
```

Fitted curves are presented with the observed values and a resampling band, with and without a qualitative factor. Without a factor, one band wraps the curve:

```{r}
b_ss <- agri_np_bootstrap(fit_ss, B = 199, n = 80, seed = 2011)
agri_np_plot(fit_ss, type = "fit", bootstrap = b_ss)
```

With a factor, one curve and one band are drawn per level, from a single bootstrap loop over the combined grid:

```{r}
if (exists("fit_qf")) {
  agri_np_plot(
    fit_qf,
    type = "fit",
    predictor = "dose",
    group = "cultivar",
    bootstrap = 199,
    seed = 2011
  )
}
```

`B = 199` keeps the vignette fast; a final analysis needs `B >= 999`.

Every figure is a plain `ggplot` object, so units and finishing stay editable. `agri_theme()` is the common visual finish, already applied to the regression graphics, and `agri_save_figure()` writes the archival formats at journal widths:

```{r}
p_fin <- agri_np_plot(fit_ss, type = "fit", bootstrap = b_ss) +
  ggplot2::labs(x = expression("Nitrogen rate (kg ha"^-1*")"),
                y = expression("Yield (Mg ha"^-1*")"))
p_fin

fig <- agri_save_figure(p_fin, tempfile(fileext = ".png"), layout = "column")
basename(fig)
# For submission, write the same object to TIFF or a vector format, e.g.
# agri_save_figure(p_fin, "figure1.tif", layout = "column", dpi = 600)
```

Prefer a vector format (PDF, SVG, EPS) when the journal allows it, because text and lines stay editable after export; use LZW-compressed TIFF where a raster format is required.

Every figure produced by the package is a plain `ggplot` object. This means that the theme, colours, scales, labels and every visual element can be replaced or extended with ordinary ggplot2 layers. The package theme `agri_theme()` is already applied, but it can always be overridden: `agri_np_plot(fit) + ggplot2::theme_bw()` or `+ ggplot2::theme_minimal(base_size = 12)` work without restriction. Tables returned by `agri_table(..., format = "data.frame")` are plain data frames that can be passed to `knitr::kable()`, `gt::gt()` or `knitr::flextable()`.

# Interactive exploration

`agri_np_interactive()` converts the corresponding `ggplot2` object to Plotly without changing the statistical fit. It is intended for teaching, dashboards, and exploratory HTML reports; static figures remain preferable for archival manuscripts.

The chunk below runs the conversion and reports the resulting object class. The
widget itself is not embedded here: a single Plotly widget carries about 4 MB of
JavaScript, which would dominate the installed size of the package for no
statistical gain. Run the same call in an interactive session to see it.

```{r interactive-conversion}
if (exists("fit_surface") && requireNamespace("plotly", quietly = TRUE)) {
  widget <- agri_np_interactive(
    fit_surface,
    type = "surface",
    surface_predictors = c("N", "water"),
    n = 40
  )
  class(widget)
}
```

# Tables, reports, dashboards, and export

Regression fits use the same reporting infrastructure as experimental-design fits.

```{r}
fit_report <- agri_np_regression(yield ~ dose, dose_data, method = "smoothing_spline")
agri_table(fit_report, "metrics")
agri_table(fit_report, "derivative", n = 40)
agri_table(fit_report, "optimum")
report_file <- agri_report(fit_report, tempfile(fileext = ".md"), format = "md")
export_results(fit_report, tempfile(fileext = ".rds"))
```

The generated report records the regression engine, formula, response family, shape restriction, block adjustment, diagnostic metrics, interpretation boundaries, backend class, and R version.

# Nonparametric significance testing for kernel regression

A flexible fitted curve does not by itself answer whether a predictor contributes information conditionally on the other predictors. For kernel regression, `agri_np_significance()` wraps the mixed-data bootstrap significance test implemented by `np::npsigtest()`. The procedure can test continuous, ordered, or unordered predictors individually or jointly while other variables remain in the conditioning set.

```{r}
if (exists("fit_kernel") && requireNamespace("np", quietly = TRUE)) {
  agri_np_significance(
    fit_kernel,
    variables = "cultivar",
    B = 99,
    boot_type = "I"
  )
}
```

Type-I calibration is used above only to keep the teaching example shorter. The `np` documentation notes that Type-I calibration can be somewhat oversized in modest samples in some settings and recommends the more computationally intensive Type-II calibration as a sensitivity check when a result is close to a conventional decision threshold. Final analyses should also use substantially more bootstrap replicates than the vignette examples.

When a block is included in a kernel model, it remains a categorical adjustment in the conditioning set unless the user explicitly requests a test of the block itself. The resulting significance test is a model-based bootstrap procedure, not a permutation test induced by the field randomization scheme.

# Testing whether a polynomial model is too restrictive

Agronomic dose-response analyses often begin with a scientifically prespecified linear, quadratic, or other low-dimensional response equation. `agri_np_specification()` provides a complementary question: is that functional form compatible with the observed conditional mean, or is there evidence that it is too restrictive relative to a flexible mixed-data alternative?

```{r}
if (requireNamespace("np", quietly = TRUE)) {
  candidate_linear <- lm(yield ~ dose, data = dose_data, x = TRUE, y = TRUE)
  candidate_quadratic <- lm(yield ~ dose + I(dose^2), data = dose_data, x = TRUE, y = TRUE)

  agri_np_specification(candidate_linear, B = 99)
  agri_np_specification(candidate_quadratic, B = 99)
}
```

The test is based on the kernel specification framework of Hsiao, Li, and Racine. A small p-value is evidence against the **prespecified parametric functional form**. It does not establish that LOESS, a smoothing spline, GAM, or any other particular flexible model is uniquely correct. The function is intentionally restricted to continuous Gaussian outcomes because the backend uses residual bootstrapping.

These tools should therefore be used to answer different questions:

- `agri_np_compare()` compares out-of-sample predictive error among scientifically admissible fitted models;
- `agri_np_significance()` tests the contribution of predictors within mixed-data kernel regression;
- `agri_np_specification()` tests whether a prespecified Gaussian parametric functional form appears inadequate;
- `agri_np_diagnostics()` summarizes residual and predictive behavior;
- none of these functions should be used as an automated search for the smallest p-value.

# Missing values in regression are never silently deleted

The regression module deliberately differs from R functions that may apply `na.omit` inside a backend without making the data reduction prominent. If the modeled response, predictor, block, or supplied weight contains missing/non-finite values, `agri_np_regression()` stops by default.

```{r}
miss <- dose_data
miss$yield[c(4, 19)] <- NA

try(
  agri_np_regression(yield ~ dose, miss, method = "smoothing_spline")
)
```

Complete-row regression must be requested explicitly:

```{r}
fit_complete <- agri_np_regression(
  yield ~ dose,
  miss,
  method = "smoothing_spline",
  na_action = "complete"
)
fit_complete$n_original
fit_complete$n_omitted
```

This is only explicit complete-row analysis. It is **not** imputation and it is **not** a missing-data model. Repeated-measures experiments with incomplete observations should use the dedicated incomplete repeated-measures workflow described in the missing-data vignette whenever its assumptions and design scope are appropriate.

# References

Hayfield T, Racine JS (2008). Nonparametric Econometrics: The np Package. *Journal of Statistical Software* 27(5). doi:10.18637/jss.v027.i05.

Pya N, Wood SN (2015). Shape constrained additive models. *Statistics and Computing* 25(3):543-559. doi:10.1007/s11222-013-9448-7.

Sen PK (1968). Estimates of the Regression Coefficient Based on Kendall's Tau. *Journal of the American Statistical Association* 63(324):1379-1389. doi:10.1080/01621459.1968.10480934.

Wood SN (2025). Generalized Additive Models. *Annual Review of Statistics and Its Application* 12:497-526. doi:10.1146/annurev-statistics-112723-034249.

Ng P, Maechler M (2007). A fast and efficient implementation of qualitatively constrained quantile smoothing splines. *Statistical Modelling* 7(4):315-328. doi:10.1177/1471082X0700700403.

Koenker R, Bassett G Jr (1978). Regression Quantiles. *Econometrica* 46(1):33-50. doi:10.2307/1913643.

Racine JS, Hart J, Li Q (2006). Testing the Significance of Categorical Predictor Variables in Nonparametric Regression Models. *Econometric Reviews* 25(4):523-544. doi:10.1080/07474930600972590.

Hsiao C, Li Q, Racine JS (2007). A consistent model specification test with mixed discrete and continuous data. *Journal of Econometrics* 140(2):802-826. doi:10.1016/j.jeconom.2006.07.015.

