---
title: "Integer-Support Nonparametric Regression for Agronomy"
author: "agriRank Development Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integer-Support Nonparametric Regression for Agronomy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r agrirank-setup, include=FALSE}
library(agriRank)
```

# Introduction

Many agronomic predictors are quantitative and ordered but **not continuous decision variables**. Examples include the number of plants per plot, insects per sampling unit, fruits retained per plant, irrigation events, sprays, shoots, branches, traps, or animals per pen. A conventional smoother can mathematically evaluate a response at \(x=5.37\), but a decision such as 5.37 plants or 8.42 insects is impossible.

This distinction affects more than presentation. If the scientific decision variable lives on an integer support, then the estimand itself should live on that support. Let

\[
\mathcal{X}_I=\{x_1,x_2,\ldots,x_K\}\subset\mathbb{Z}
\]

denote the admissible integer decisions. The fitted maximum is defined as

\[
x_I^*=\arg\max_{x\in\mathcal{X}_I}\widehat m(x),
\]

and the corresponding minimum replaces \(\max\) with \(\min\). `agriRank` therefore does **not** estimate a continuous optimum and round it. It evaluates the fitted response directly at admissible integer values.

This vignette introduces four complementary strategies:

1. ordered-discrete kernel regression;
2. unimodal isotonic regression;
3. umbrella-order constrained regression;
4. a continuous/flexible latent model projected onto an integer decision grid.

It also introduces finite differences, practical thresholds, bootstrap distributions of the optimum, and discrete confidence sets.

# Why rounding a continuous optimum is not enough

Suppose a continuous smoother has a maximum at \(5.62\) plants per plot. Rounding produces 6, but rounding is not equivalent to solving

\[
\arg\max_{x\in\{1,2,\ldots,10\}}\widehat m(x).
\]

The fitted responses at 5 and 6 may not be symmetric around the continuous maximum, shape constraints can create plateaus, and some integer values may never have been evaluated experimentally.

`agriRank` distinguishes three supports.

## Observed integer support

```r
predictor_support = "observed_integer"
```

If the experiment evaluated

\[
\{1,3,5,7,9\},
\]

only those five decisions are admissible. This is the most conservative option because it avoids interpolation to untested treatments.

## Full integer range

```r
predictor_support = "integer_range"
```

If the range is 1 to 9, the decision support becomes

\[
\{1,2,3,\ldots,9\}.
\]

This permits interpolation to unobserved integers and should be used only when such interpolation is scientifically defensible.

## Custom integer support

```r
predictor_support = "custom_integer"
integer_values = c(2, 4, 6, 8, 10)
```

This is useful when operational constraints define the feasible treatment set.

## The three supports side by side

The declaration changes which decisions the package is willing to recommend, and
therefore it changes the answer. The comparison below fits the same data three
times and reports the admissible support and the resulting discrete optimum.

```{r three-supports}
set.seed(717)
spacing <- data.frame(plants = rep(1:9, each = 8))
spacing$yield <- 18 + 6.2 * spacing$plants - 0.52 * spacing$plants^2 +
  rnorm(nrow(spacing), 0, 0.9)

fit_support <- function(support, ...) {
  f <- agri_np_regression(
    yield ~ plants, spacing, method = "integer_grid",
    integer_base_method = "gam", predictor_support = support, ...
  )
  data.frame(
    support = support,
    admissible = paste(range(f$integer_support), collapse = " to "),
    n_decisions = length(f$integer_support),
    optimum = agri_integer_optimum(f)$optima$plants[1]
  )
}

rbind(
  fit_support("observed_integer"),
  fit_support("integer_range", integer_range = c(1, 14)),
  fit_support("custom_integer", integer_values = c(1:9, 12))
)
```

Note that `integer_range = c(1, 14)` admits decisions up to 14 plants although
the experiment never tested more than 9. Extrapolation of that kind must be
justified agronomically, never by convenience.

A support that does not contain every observed predictor value is rejected:

```{r support-rejection, error=TRUE}
agri_np_regression(
  yield ~ plants, spacing, method = "integer_grid",
  integer_base_method = "gam", predictor_support = "custom_integer",
  integer_values = c(2, 4, 6)
)
```

# Method 1: ordered-discrete kernel regression

The `np` framework supports continuous, unordered categorical, and ordered categorical predictors in the same nonparametric regression model. An integer predictor in `agriRank` is internally encoded as an ordered factor for `method = "discrete_kernel"`.

For an ordered predictor, one available kernel is the Wang--van Ryzin kernel. In the notation used by current `np` documentation,

\[
K_\lambda(x_i,x)=
\begin{cases}
1-\lambda, & x_i=x,\\
\dfrac{1-\lambda}{2}\lambda^{|x_i-x|}, & x_i\neq x.
\end{cases}
\]

Nearby integer levels receive more weight than distant levels. The smoothing parameter is selected as part of the kernel regression workflow rather than by treating the predictor as continuous.

The method is especially useful when:

- the response shape is unknown;
- there may be several predictors;
- integer \(X\) occurs together with factors such as cultivar or block;
- no unimodality or monotonicity constraint is scientifically justified.

## Example 1: insect density and crop damage

```{r discrete-kernel}
if (requireNamespace("np", quietly = TRUE)) {
  set.seed(4101)

  insects <- data.frame(
    insects = rep(0:10, each = 7)
  )
  insects$damage <- 3.5 + 1.25 * insects$insects +
    0.08 * insects$insects^2 +
    rnorm(nrow(insects), 0, 1.4)

  fit_dk <- agri_np_regression(
    damage ~ insects,
    data = insects,
    method = "discrete_kernel",
    predictor_support = "observed_integer",
    integer_kernel = "wangvanryzin"
  )

  fit_dk
  agri_integer_predict(fit_dk)
  agri_np_plot(fit_dk)
}
```

A request for 4.5 insects is rejected:

```{r discrete-reject, error=TRUE}
if (exists("fit_dk")) {
  agri_np_predict(fit_dk, data.frame(insects = 4.5))
}
```

The rejection is intentional. It protects the estimand, not merely the printing format.

# Method 2: unimodal isotonic regression

Many agronomic responses have a biologically plausible single optimum. For example, productivity may initially increase with plant density because more plants occupy the area, then decline when competition becomes excessive.

Unimodal isotonic regression estimates a sequence

\[
\theta_1,\ldots,\theta_K
\]

that minimizes weighted squared error subject to the existence of a mode \(m\):

\[
\theta_1\le\cdots\le\theta_m
\quad\text{and}\quad
\theta_m\ge\cdots\ge\theta_K.
\]

When the mode is not prespecified, `Iso::ufit()` searches the admissible \(x\)-locations and selects the one giving the smallest residual sum of squares. This is attractive for integer agronomic treatments because the fitted mode itself is an admissible location.

## Example 2: number of plants per plot

```{r unimodal}
if (requireNamespace("Iso", quietly = TRUE)) {
  set.seed(4102)

  density <- data.frame(
    plants = rep(1:10, each = 6)
  )
  density$yield <- 28 +
    8.5 * pmin(density$plants, 6) -
    5.2 * pmax(density$plants - 6, 0) +
    rnorm(nrow(density), 0, 2)

  fit_ui <- agri_np_regression(
    yield ~ plants,
    data = density,
    method = "unimodal_isotonic",
    predictor_support = "observed_integer"
  )

  fit_ui
  agri_integer_predict(fit_ui)
  agri_integer_optimum(fit_ui)
}
```

This model is appropriate only when a single increase-then-decrease structure is scientifically defensible. If the response may contain several peaks, use a less restrictive strategy.

# Method 3: umbrella-order constrained regression

Umbrella ordering expresses the same broad increase-then-decrease scientific pattern within a constrained generalized additive/order-restricted model.

For a response with systematic component \(\eta_i\),

\[
\eta_i=f_{\text{umb}}(x_i)+\mathbf z_i^\top\boldsymbol\beta,
\]

where

\[
f_{\text{umb}}(x_1)\le\cdots\le f_{\text{umb}}(x_m)
\ge\cdots\ge f_{\text{umb}}(x_K).
\]

The `cgam` package implements umbrella ordering and also permits parametric covariates. `agriRank` uses this capability to retain an agronomic block as an adjustment factor rather than discarding the experimental structure.

## Example 3: plant density in an RCBD

```{r umbrella}
if (requireNamespace("cgam", quietly = TRUE)) {
  set.seed(4103)

  rcbd <- expand.grid(
    block = factor(1:5),
    plants = 1:9
  )

  rcbd$yield <- 32 +
    9 * pmin(rcbd$plants, 6) -
    5.5 * pmax(rcbd$plants - 6, 0) +
    as.numeric(rcbd$block) * 0.7 +
    rnorm(nrow(rcbd), 0, 1.8)

  fit_umb <- agri_np_regression(
    yield ~ plants,
    data = rcbd,
    method = "umbrella",
    block = block,
    predictor_support = "observed_integer"
  )

  fit_umb
  agri_integer_predict(fit_umb)
  agri_integer_optimum(fit_umb)
  agri_np_plot(fit_umb)
}
```

The block is a model adjustment. It should not be removed simply because its effect is small.

# Method 4: integer-grid decisions from a flexible latent regression

Sometimes a continuous latent response curve is scientifically reasonable, while the operational decision remains integer-valued. For example, competition among plants can vary smoothly as density changes, even though the field manager can plant only whole plants per experimental unit.

`method = "integer_grid"` fits a selected latent model

\[
\widehat m_c(x)
\]

but defines all public decisions on

\[
\mathcal{X}_I.
\]

Thus

\[
x_I^*=
\arg\max_{x\in\mathcal{X}_I}\widehat m_c(x).
\]

No rounding step is involved.

Supported latent backends include GAM, SCAM, kernel regression, quantile regression, LOESS, smoothing splines, and COBS, subject to their usual design restrictions.

## Example 4: smoothing spline with integer-only decisions

```{r integer-grid}
set.seed(4104)

ig <- data.frame(
  plants = rep(1:12, each = 6)
)

ig$yield <- 24 + 7.5 * ig$plants -
  0.48 * ig$plants^2 +
  rnorm(nrow(ig), 0, 1.4)

fit_ig <- agri_np_regression(
  yield ~ plants,
  data = ig,
  method = "integer_grid",
  integer_base_method = "smoothing_spline",
  predictor_support = "integer_range",
  integer_range = c(1, 12)
)

agri_integer_predict(fit_ig)
agri_integer_optimum(fit_ig)
agri_np_plot(fit_ig)
```

# Finite differences instead of instantaneous derivatives

For a continuous predictor, a derivative such as

\[
m'(x)
\]

describes infinitesimal change. For a count decision, the direct agronomic question is often:

> What is the fitted gain when I move from \(k\) to \(k+1\) plants?

The first finite difference is

\[
\Delta \widehat m(k)
=
\widehat m(k+1)-\widehat m(k).
\]

The second finite difference on unit-spaced support is

\[
\Delta^2\widehat m(k)
=
\widehat m(k+1)-2\widehat m(k)+\widehat m(k-1).
\]

```{r differences}
agri_integer_difference(fit_ig, order = 1)
agri_integer_difference(fit_ig, order = 2)
```

Calling `agri_np_derivative()` on an integer-support fit redirects the interpretation to finite differences.

# Practical efficiency and threshold rules

A biological maximum is not always the most useful field decision. `agriRank` therefore exposes practical integer criteria that must be declared explicitly.

## Response relative to the fitted maximum

```{r efficiency}
eff <- agri_integer_efficiency(fit_ig)
eff
```

This table reports the fitted response and the gain associated with moving between adjacent admissible counts.

## Smallest integer attaining a fraction of the maximum

Define

\[
x_\rho=
\min\left\{
x\in\mathcal X_I:
\widehat m(x)\ge
\rho\max_{s\in\mathcal X_I}\widehat m(s)
\right\}.
\]

For \(\rho=0.95\):

```{r threshold95}
agri_integer_threshold(
  fit_ig,
  criterion = "fraction_of_maximum",
  value = 0.95
)
```

This can be more useful than the absolute maximum when several integer treatments produce nearly equivalent fitted performance.

## Minimum count giving a target gain

```{r threshold-gain}
agri_integer_threshold(
  fit_ig,
  criterion = "gain_from_baseline",
  value = 15,
  baseline = 1
)
```

## Diminishing marginal gain

```{r threshold-marginal}
agri_integer_threshold(
  fit_ig,
  criterion = "marginal_gain",
  value = 1
)
```

This identifies the first transition for which the fitted gain per additional integer unit is no larger than the declared threshold.

These are biological or operational rules. An **economic optimum** additionally requires prices, costs, and an economic objective function.

# Bootstrap uncertainty for the integer optimum

A continuous confidence interval such as 5.3--6.8 plants is awkward because much of the interval consists of impossible decisions. A more direct approach is to refit the model in bootstrap samples and record the integer optimum in each replicate.

Let

\[
x_b^*=
\arg\max_{x\in\mathcal X_I}\widehat m_b(x)
\]

for bootstrap replicate \(b\). The empirical distribution

\[
\widehat P(x^*=k)
=
\frac{1}{B}
\sum_{b=1}^B I(x_b^*=k)
\]

summarizes uncertainty over actual field decisions. When a fitted plateau gives several tied optima, `agriRank` divides that replicate's mass equally among them.

```{r bootstrap-optimum}
boot_opt <- agri_integer_bootstrap(
  fit_ig,
  B = 199,
  seed = 4105
)

boot_opt
```

For blocked models, whole blocks are resampled by default.

# A discrete confidence set

`agri_integer_confset()` builds a highest-probability set containing only admissible integers. Support values are ranked by bootstrap probability and accumulated until the requested probability mass is reached.

```{r confset}
agri_integer_confset(
  boot_opt,
  level = 0.95
)
```

A result such as

\[
\{5,6,7\}
\]

has a direct agronomic interpretation that a fractional interval does not.

# Observed support versus integer interpolation

This choice should be made explicitly.

Suppose an experiment tested only

\[
\{1,3,5,7,9\}.
\]

```{r support-comparison}
set.seed(4106)

sparse <- data.frame(
  plants = rep(c(1, 3, 5, 7, 9), each = 6)
)
sparse$yield <- 20 + 7*sparse$plants -
  0.48*sparse$plants^2 +
  rnorm(nrow(sparse), 0, 1.2)

fit_observed <- agri_np_regression(
  yield ~ plants,
  data = sparse,
  method = "integer_grid",
  integer_base_method = "smoothing_spline",
  predictor_support = "observed_integer"
)

fit_full <- agri_np_regression(
  yield ~ plants,
  data = sparse,
  method = "integer_grid",
  integer_base_method = "smoothing_spline",
  predictor_support = "integer_range",
  integer_range = c(1, 9)
)

agri_integer_predict(fit_observed)
agri_integer_predict(fit_full)
```

The second analysis includes interpolated decisions at 2, 4, 6, and 8. Those are integer-valued but were not directly tested. `agriRank` therefore records the support mode in the fitted object and in generated reports.

# Choosing among the four approaches

A useful hierarchy is:

| Scientific situation | Preferred starting point |
|---|---|
| Unknown shape; integer predictor may coexist with factors | `discrete_kernel` |
| Strong evidence for one increase-then-decrease pattern; one predictor | `unimodal_isotonic` |
| Umbrella pattern plus block or other parametric adjustment | `umbrella` |
| A smooth latent curve is plausible, but decisions must be whole numbers | `integer_grid` |

The four approaches answer related but not identical questions. They should not be run merely to select whichever yields the most attractive optimum.

# Counts as predictors versus counts as responses

An integer-valued predictor does not imply a Poisson response model.

If

\[
X=\text{number of insects}
\quad\text{and}\quad
Y=\text{yield},
\]

then \(X\) is an ordered discrete predictor while the response model is determined by the distribution and scientific target of \(Y\).

If instead

\[
Y=\text{number of insects},
\]

then the **response** is a count and a count-response model may be appropriate. These are different modeling decisions.

# Missing data and dependence safeguards

Integer support does not override the package's general safeguards.

- Missing modeled values fail by default.
- `na_action = "complete"` must be requested explicitly.
- Repeated/longitudinal observations are not treated as independent merely because the predictor is an integer.
- A block is retained only by engines that can represent it.
- `unimodal_isotonic` currently rejects a declared block because an unadjusted isotonic fit would not preserve the RCBD structure.
- `umbrella`, `discrete_kernel`, and appropriate `integer_grid` backends can retain block adjustment.

# Reporting

Integer-support information propagates to the reporting layer:

```{r reporting}
agri_table(fit_ig, what = "integer_predictions")
agri_table(fit_ig, what = "integer_optimum")
agri_table(fit_ig, what = "integer_efficiency")
```

The Markdown/Quarto report also records the support type and admissible values.

# Reproducibility checklist

For a defensible integer-support analysis, report:

1. the experimental unit and design;
2. the scientific meaning of the integer predictor;
3. the admissible support;
4. whether untested intermediate integers were allowed;
5. the regression engine and any shape constraint;
6. block or other adjustment variables;
7. the criterion defining maximum, minimum, or practical efficiency;
8. the bootstrap scheme;
9. the discrete confidence set when uncertainty in the optimum is important;
10. software and package versions.

# State-of-the-art references

The methodological metadata below were checked against at least two independent sources before inclusion in `agriRank`.

- **Wang, M.-C. & van Ryzin, J. (1981).** A class of smooth estimators for discrete distributions. *Biometrika*, 68(1), 301--309. DOI: 10.1093/biomet/68.1.301. Oxford Academic and JSTOR independently confirm authors, volume, issue, year, and pages.
- **Racine, J. & Li, Q. (2004).** Nonparametric estimation of regression functions with both categorical and continuous data. *Journal of Econometrics*, 119(1), 99--130. DOI: 10.1016/S0304-4076(03)00157-X. Elsevier and McMaster University independently confirm the metadata.
- **Turner, T. R. & Wollan, P. C. (1997).** Locating a maximum using isotonic regression. *Computational Statistics & Data Analysis*, 25(3), 305--320. DOI: 10.1016/S0167-9473(97)00009-1. Elsevier and RePEc independently confirm the metadata.
- **Stout, Q. F. (2008).** Unimodal regression via prefix isotonic regression. *Computational Statistics & Data Analysis*, 53(2), 289--297. DOI: 10.1016/j.csda.2008.08.005. Elsevier and the author's University of Michigan publication record independently confirm the metadata.
- **Geng, Z. & Shi, N.-Z. (1990).** Isotonic Regression for Umbrella Orderings. *Journal of the Royal Statistical Society Series C: Applied Statistics*, 39(3), 397--402. DOI: 10.2307/2347399. Oxford Academic and RePEc independently confirm the metadata.
- **Liao, X. & Meyer, M. C. (2019).** cgam: An R Package for the Constrained Generalized Additive Model. *Journal of Statistical Software*, 89(5). DOI: 10.18637/jss.v089.i05. JSS and the archived preprint independently confirm the metadata.

# Interpretation boundary

The new module restricts **decisions and inferential summaries** to feasible integer values. It does not imply that the underlying biological process is intrinsically discontinuous. `integer_grid` is deliberately available for situations in which a smooth latent response is scientifically plausible but operational decisions remain discrete.

The principal rule is therefore:

> Fit the scientific relationship with an appropriate method, but define the decision on the support that can actually be implemented in the experiment or field.
